---
title: "Lecture 9"
subtitle: "Hedonics: Property value models"
author: Ivan Rudik
date: AEM 6510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'


---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, tidylog, xaringanExtra, rlang, patchwork
)
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

---

# Roadmap

- What can we use to infer the demand for environmental goods?
- What do housing prices tell us?
- When do changes in house prices give us welfare measures

---

# Environmental quantity changes

Last time we learned that we cannot directly compute the welfare consequences of a change in environmental quantities

--

There are no markets for the environmental goods

--

We don't observe the ordinary inverse demand curve

--

We can't compute CS, EV, CV, etc!

---

# Revealed preference approaches

One way to circumvent this problem is to look at private goods that interact with the environmental good

--

If there are changes in the environmental good, holding everything else fixed, that should be reflected in *some way* in changes in the price of the related private good

--

This change in price can tell us something about how people value the change in the environmental good

---

# Revealed preference approaches

There is no market for orcas

--

Suppose there's a massive decline in orcas off the Washington coast, what happens?

--

We will likely see demand for sightseeing tours go down (MB of these tours went down!)

--

This drops the price of tours

--

A non-market good had an effect on a market price

--

What does this price change mean?

---

# Hedonics: Property value models

Common market goods to use for revealed preference valuation are .hi-blue[properties]

--

When people buy a home they are purchasing a bundle of goods:
  - Rooms
  - Bathrooms
  - School quality
  - .hi[Environmental quality]

--

Homes located in pristine areas are likely to be more valuable than identical homes located near toxic facilities

---

# Hedonics: Property value models

Real estate is virtually ideal for measuring environmental changes

--

Real estate markets are often competitive and thick

--

Property purchases are large and consequential: buyers and sellers are likely to be well-informed

--

It is uncontroversial that property values should reflect local attributes

---

# The hedonic model

Property value approaches are often called .hi-blue[hedonics] because they rely on the hedonic model

--

Suppose that we have some quality-differentiated good (i.e. a home)

--

This good is characterized by a set of $J$ property characteristics $x$
  - parcel size, school quality, bedrooms, etc
  
--

It is also characterized by an environmental good $q$


---

# The hedonic model

The price of a house is determined by a .hi[hedonic price function] $P(x,q)$


--

$P$ maps characteristics of the house and local environment to the market value of the home

--

For a particular house $k$ its price is $p_k = P(x_k,q_k)$

--

$P$ arises in equilibrium from the interaction of all buyers in sellers in the market

--

Here we will assume the supply of houses is fixed in the short run so the price function arises from buyer behavior


---

# The hedonic model: consumer's choice problem

Households buy a single property given their budget constraint and their preferences

--

Here we will assume that households are effectively just choosing $(x,q)$ instead of a specific house with the following objective:
$$\max_{x,q,z} U(x,q,z;s) \,\,\,\,\, s.t. \,\,\,\,\, y = z + P(x,q)$$

--

- $z$ is the numeraire good (spending on other private goods)
- $y$ is income
- $s$ is the set of the household's characteristics like family size


---

# Unrealistic pieces of the model

One unrealistic part of this model is that we are assuming household characteristics are continuous

--

Many housing characteristics are discrete (bedrooms, bathrooms, etc)

--

Another is that you just can't purchase some sets of $x$ (i.e. a huge lot in downtown manhattan with a farm)

--

We won't touch on this in class but there is a .hi-blue[discrete choice] literature that works to alleviate these issues


---

# Choosing q

Another thing to note: the consumer *chooses* $q$ where as before it was fixed

--

The idea is that mobile households can move to get their desired level of the environmental good

--

We are thus also implicitly assuming $q$ varies across space so that households can sort into areas they prefer
  - q is really picking up .hi[local] environmental goods
  

---

# What is $P(x,q)$

In the model we are thinking of $P(x,q)$ as the annual .hi-blue[rental rate], not the purchase price

--

This allows us to mesh more cleanly with annual income and view the problem as static rather than dynamic

--

This clearly works well for renting households

--

For homeowners we are basically assuming they rent from themselves every year

---

# The hedonic model: consumer's choice problem

$$\max_{x,q,z} U(x,q,z;s) \,\,\,\,\, s.t. \,\,\,\,\, y = z + P(x,q)$$

The FOCs for this problem are:
\begin{align}
  {\partial U \over \partial x_j} =& \lambda {\partial P \over \partial x_j} \,\,\, j=1,\dots,J \\
  {\partial U \over \partial q} =& \lambda {\partial P \over \partial q} \\
  {\partial U \over \partial z} =& \lambda 
\end{align}

Next, combine the last two FOCs

---

# The hedonic model: consumer's choice problem

\begin{align}
  {\partial U \over \partial q} =& \lambda {\partial P \over \partial q} \\
  {\partial U \over \partial z} =& \lambda 
\end{align}
gives us that
$${\partial P \over \partial q} = {\partial U \over \partial q} \bigg/ {\partial U \over \partial z}$$

At a utility-maximizing choice, a household equates their MRS between $q$ and $z$ and the marginal implicit cost of $q$


---

# The hedonic model: consumer's choice problem

$${\partial P \over \partial q} = {\partial U \over \partial q} \bigg/ {\partial U \over \partial z}$$

Recall that $z$ is the numeraire good so we can think of it in terms of dollars

--

This means that ${\partial U \over \partial q} \big/ {\partial U \over \partial z}$ is the WTP for $q$, the reduction in income needed to compensate for an additional unit of $q$

--

Knowledge of the hedonic price function $P$ is enough to tell us about household WTP for $q$!


---

# The hedonic model: bid functions

Now let's dive deeper by looking at some reference utility level $\bar{u}$:
$$U(x,q,z;s) = \bar{u}$$

--

Next we will define something called a .hi-blue[bid function] $b(x,q,y,s,\bar{u})$ where:
$$U(x,q,y -b(x,q,y,s,\bar{u}) ;s) = \bar{u}$$

The bid function $b$ is the maximum amount the household is willing to pay for:  
- A house with characteristics $x,q$
- Given income $y$ and household characteristics $s$
- Holding utility fixed

---

# The hedonic model: bid functions

$$U(x,q,z;s) = \bar{u}$$

We can also invert this to solve for $z$:<sup>1</sup>
$$z = U^{-1}(x,q,\bar{u},s)$$

.footnote[
<sup>1</sup> We can do this because $U$ is monotonically increasing in $z$
]

--

Income, the bid function and $z$ are related by:
$$b(x,q,y,s,\bar{u}) ;s) = y - U^{-1}(x,q,\bar{u},s)$$

--

Now we have everything we need to derive a marginal WTP function for $q$

---

# The hedonic model: deriving MWTP

$$U(x,q,y -b(x,q,y,s,\bar{u}) ;s) = \bar{u}$$

Differentiate with respect to $q$ to get:
$${\partial U \over \partial q} + {\partial U \over \partial z} {\partial b \over \partial q} = 0$$

--

We can then rearrange to get:
$${\partial b \over \partial q}  = {\partial U \over \partial q}\bigg/{\partial U \over \partial z}$$



---

# The hedonic model: deriving MWTP

Recall that the bid function is separable in income: $b(x,q,y,s,\bar{u}) ;s) = y - U^{-1}(x,q,\bar{u},s)$

--

This lets us re-write ${\partial b \over \partial q}$ as:
$$\pi^q(x,q,s,\bar{u}) = {\partial b \over \partial q}  = {\partial U \over \partial q}\bigg/{\partial U \over \partial z}$$

--

Conditional on $x$, this defines our .hi-blue[compensated inverse demand function] for $q$!

---

# The hedonic model: deriving MWTP

$$\pi^q(x,q,s,\bar{u}) = {\partial b \over \partial q}  = {\partial U \over \partial q}\bigg/{\partial U \over \partial z}$$

By assuming $q$ is a part of a larger bundle of house characteristics, we are able to characterize its demand through its relationship to the housing market

--

We can then use the bid function (which maps into prices) to understand the marginal WTP for $q$

--

Our ultimate empirical goal is to estimate $\pi^q(x,q,s,\bar{u})$


---

# Bid functions and housing prices

```{r bid_functions_1, echo = FALSE, fig.show = 'hide', warning = F}
bid1 <- function(x) -x^2/2*1.2 + 5*x + 3
bid2 <- function(x) -x^2/2*1.2 + 5*x
bid3 <- function(x) -x^2/2*1.2 + 5*x - 3
hedonic_p <- function(x) -x^2/2.2 + 4.25*x + 1
ggplot() +
  stat_function(fun = bid1, color = "#638ccc", size = 1.0) +
  stat_function(fun = bid2, color = "#638ccc", size = 1.0) +
  stat_function(fun = bid3, color = "#638ccc", size = 1.0) +
  stat_function(fun = hedonic_p, color = "#ca5670", size = 1.0) +
    annotate("segment", x = 2.75, xend = 2.75, y = 0, yend = 9.2,
           linetype = "dotted", size = 1, color = "grey55") +
    annotate("segment", x = 0, xend = 2.75, y = 9.2, yend = 9.2,
           linetype = "dotted", size = 1, color = "grey55") +
  annotate("text", size = 6, label = expression(b(x,q,y,s, u^2)), x = 3.25, y = 14, color = "#638ccc") +
  annotate("text", size = 6, label = "P(x,q)", x = 3.25, y = 11, color = "#ca5670") +
  annotate("text", size = 6, label = expression(b(x,q,y,s, u^1)), x = 3.25, y = 5, color = "638ccc") +
  annotate("text", size = 6, label = expression(b(x,q,y,s, u^0)), x = 3.25, y = 8.5, color = "#638ccc") +
  annotate("text", size = 6, label = "a", x = 2.6, y = 8, color = "black") +
  theme_minimal() +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 16, color = "#000000"), axis.text.y = element_text(size = 16, color = "#000000"),
    axis.title.x = element_text(size = 16, color = "#000000"), axis.title.y = element_text(size = 16, color = "#000000"),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee",colour = NA),
    plot.background = element_rect(fill = "#eeeeee",colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 4), breaks = c(2.75), labels = expression(q[0])) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 15), breaks = c(9.2), labels = expression(P(x,q[0]))) +
  labs(x = "q",
       y = "$",
       title = "Hedonic price function\nand bid functions")
  
```

.pull-left[

The red line is the hedonic price function

The blue lines are a single household's bid functions at different reference utility levels where $u_1 > u_0 > u_2$

Higher utility $\rightarrow$ lower bids because same level of $q$ can be achieved with higher $z$

]

.pull-right[

![bid function is tangent to hedonic price function](`r knitr::fig_chunk("bid_functions_1", "png")`)

]


---

# Bid functions and housing prices

.pull-left[

Optimal choice is where the household's bid function is tangent to the hedonic price schedule: $a$

This gives us an observed consumption level $q_0$, observed price $P(x,q_0)$, and realized utility $u^0$

]

.pull-right[

![bid function is tangent to hedonic price function](`r knitr::fig_chunk("bid_functions_1", "png")`)

]


---

# Compensated MWTP

```{r bid_functions_2, echo = FALSE, fig.show = 'hide', warning = F}
cwtp <- function(x) 12-3*x
ggplot() +
  stat_function(fun = cwtp, color = "#638ccc", size = 1.0) +
    annotate("segment", x = 1.3, xend = 1.3, y = 0, yend = 12-3*1.3,
           linetype = "dotted", size = 1, color = "grey55") +
    annotate("segment", x = .5, xend = .5, y = 0, yend = 12-3*.5,
           linetype = "dotted", size = 1, color = "grey55") +
    annotate("segment", x = 0, xend = 1.3, y = 12-3*1.3, yend = 12-3*1.3,
           linetype = "dotted", size = 1, color = "grey55") +
  annotate("text", size = 6, label = expression(paste(pi^q, "(x,q,s,", u^0, ")")), x = 3.25, y = 5, color = "638ccc") +
  annotate("text", size = 6, label = "b", x = 1.25, y = 12-3*1.3+1, color = "black") +
  theme_minimal() +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 16, color = "#000000"), axis.text.y = element_text(size = 16, color = "#000000"),
    axis.title.x = element_text(size = 16, color = "#000000"), axis.title.y = element_text(size = 16, color = "#000000"),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee",colour = NA),
    plot.background = element_rect(fill = "#eeeeee",colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 4), breaks = c(0.5, 1.3), labels = expression(q[1], q[0])) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 15), breaks = c(12-3*1.3), labels = expression(paste(partialdiff, P(x, q[0]), "/", partialdiff, q))) +
  labs(x = "q",
       y = "$",
       title = "Compensated MWTP")
  
```

.pull-left[

This plot shows the corresponding compensated MWTP curve associated with $b(x,q,y,s,u^0)$

It is the slope of the bid function as $q$ changes

We observe $b$ if we can estimate $P(x,q)$ and its derivative

]

.pull-right[

![bid function is tangent to hedonic price function](`r knitr::fig_chunk("bid_functions_2", "png")`)

]


---

# Compensated MWTP

.pull-left[

We can estimate $P(x,q)$ using home sales prices and home attributes data

The slope of $P(x,q)$ is then equal to the MWTP for $q$

This gives us the consumers inverse demand for $q$

$${\partial P(x,q_0) \over \partial q} = \pi^q(x,q_0,s,u^0)$$

]

.pull-right[

![bid function is tangent to hedonic price function](`r knitr::fig_chunk("bid_functions_2", "png")`)

]