---
title: "Lecture 4"
subtitle: "Theory of environmental policy"
author: Ivan Rudik
date: AEM 6510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'


---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, tidylog, xaringanExtra, Ryacas, rlang
)
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

<!-- --- -->

<!-- # Roadmap -->

<!-- Develop a simple model of -->
<!--   - Pollution damages -->
<!--   - Abatement costs -->
<!--   - Characteristics of efficient pollution allocations -->

<!-- This will guide us in -->
<!--   - Describing the set of policy instruments and their properties -->
<!--   - Information needs for using each kind of policy -->

<!-- --- -->

<!-- # A model of damages and costs -->

<!-- Here's our set up: -->

<!-- We have a number of households in a given area -->

<!-- There are $J$ electricity-generating firms -->

<!-- The firms use coal and emit sulfur dioxide ( $SO_2$ ) that negatively affects the households -->

<!-- The firms take output prices as given, and sell electricity in the national market, households buy electricity on the national market -->

<!-- --- -->

<!-- # A model of damages and costs -->

<!-- Each firm emits $e_j$ units of pollution and total emissions are $E = \sum_{j=1}^J e_j$ -->

<!-- The model is .hi-blue[non-spatial:] -->
<!--   - All firms emissions count the same toward aggregate emissions $E$ -->
<!--   - All households experience the same level of pollution $E$ -->

<!-- --- -->

<!-- # The damage function -->

<!-- Assume households have utility: -->
<!-- $$U_i(y_i, E) = y_i - D_i(E)$$ -->
<!-- where $y_i$ is income spent on market goods and $D_i(E)$ is the household-specific disutility caused by aggregate pollution -->

<!-- -- -->

<!-- Since utility is quasi-linear in income and the marginal utility of income is 1, we can interpret $D_i(E)$ as the *dollar value of lost utility for household $i$ from aggregate emissions* -->

<!-- --- -->

<!-- # The damage function -->

<!-- We call $D_i(E)$ the .hi-blue[damage function] -->

<!-- -- -->

<!-- Assume that the damage function is increasing and convex:  -->
<!-- $$D_i'(E) > 0, \,\, D_i''(E) > 0$$ -->

<!-- -- -->

<!-- Aggregate damages are then  -->
<!-- $$D(E) = \sum_{i=1}^N D_i(E)$$ -->
<!-- where $I$ is the number of households -->

<!-- --- -->

<!-- # Abatement costs -->

<!-- Generating electricity has costs -->

<!-- -- -->

<!-- There are also costs with reducing emissions of $SO_2$ -->

<!-- -- -->

<!-- For now we assume that these costs are separable so we can focus solely on costs of abating pollution -->

<!-- -- -->

<!-- Define the .hi-blue[abatement cost function] for firm $j$ by $C_j(e_j)$ -->

<!-- --- -->

<!-- # Abatement costs -->

<!-- Let $\hat{e}_j$ be the firm's freely chosen abatement level, then $C_j(\hat{e}_j) = 0$ -->

<!-- -- -->

<!-- $C_j(e_j) > 0$ for any $e_j < \hat{e}_j$ so that $C_j'(e_j) < 0$ for $e_j < \hat{e}_j$ -->

<!-- -- -->

<!-- Define the .hi[marginal abatement cost] as: -->
<!-- $$MAC_j(e) \equiv -C_j'(e_j) > 0, e_j < \hat{e}_j$$ -->

<!-- -- -->

<!-- We will also assume that the MAC weakly increases as we reduce emissions, i.e. it is weakly convex: -->
<!-- $$MAC_j'(e) = -C_j''(e_j) \leq 0, e_j < \hat{e}_j$$ -->

<!-- --- -->

<!-- # Abatement costs assumptions -->

<!-- These assumptions on marginal abatement cost are pretty reasonable: -->

<!-- 1. They are increasing because firms will choose among different abatement technologies in order of their marginal cost if they are profit-maximizing or cost-minimizing -->
<!-- 2. Weak convexity is a reasonable approximation of piecewise constant MAC functions, which is what many MACs look like empirically -->

<!-- --- -->

<!-- # Efficient allocation of emissions -->

<!-- In our setting emissions negatively affects households and controlling emissions imposes a cost on firms -->

<!-- An efficient outcome optimally balances these two different costs to the economy -->

<!-- -- -->

<!-- The efficient emission level for each firm $j$ can be found by minimizing the social costs of emissions: -->
<!-- $$SC(e_1,\dots,e_J) = \sum_{j=1}^J C_j(e_j) + D(E)$$ -->

<!-- --- -->

<!-- # Efficient allocation of emissions -->

<!-- The first-order conditions for minimizing SC are: -->
<!-- $$-C_j'(e_j) = D'(E), \,\, \forall j=1,\dots,J$$ -->
<!-- and also that -->
<!-- $$-C_j'(e_j) = -C_k'(e_k), \,\, \forall k,j$$ -->

<!-- These are the two fundamental characteristics of the efficient allocation of pollution -->

<!-- --- -->

<!-- # Efficient allocation of emissions -->

<!-- For each firm, its marginal abatement cost is equal to the marginal damage from pollution -->
<!-- $$-C_j'(e_j) = D'(E), \,\, \forall j=1,\dots,J$$ -->

<!-- Marginal abatement costs are equal across all polluters -->
<!-- $$-C_j'(e_j) = -C_k'(e_k), \,\, \forall k,j$$ -->

<!-- An optimal regulation will satisfy these two condition -->


<!-- --- -->

<!-- # Abatement costs and damages -->

<!-- .pull-left[ -->
<!-- ```{r abatement_cost, echo = FALSE, fig.show = 'hide', warning = F} -->
<!-- abatement_cost <- function(x) 1/x -->
<!-- intersections <- tibble(x = 1, y = 1) -->
<!-- ggplot() + -->
<!--   annotate("segment", x = 0, xend = intersections$x, y = intersections$y, yend = intersections$y,  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections$x, xend = intersections$x, y = 0, yend = intersections$y,  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   stat_function(fun = abatement_cost, color = "#638ccc", size = 1.5) + -->
<!--   theme_minimal()  + -->
<!--   theme( -->
<!--     legend.position = "none", -->
<!--     title = element_text(size = 24), -->
<!--     axis.text.x = element_text(size = 24, color = "black"),  -->
<!--     axis.text.y = element_text(size = 24, color = "black"), -->
<!--     axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24), -->
<!--     panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(), -->
<!--     panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(), -->
<!--     panel.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     plot.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     axis.line = element_line(colour = "black") -->
<!--   ) + -->
<!--   labs(x = "Emissions", -->
<!--        y = "Abatement cost", -->
<!--        title = "The abatement cost function") + -->
<!--   scale_x_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = intersections$x, labels = expression(bar(e)[1])) + -->
<!--   scale_y_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = intersections$y, labels = expression(C[1](bar(e)[1]))) -->

<!-- ``` -->

<!-- ![Abatement cost functions are decreasing in emissions](`r knitr::fig_chunk("abatement_cost", "png")`) -->
<!-- ] -->

<!-- .pull-right[ -->

<!-- ```{r damages, echo = FALSE, fig.show = 'hide', warning = F} -->
<!-- abatement_cost <- function(x) exp(x) - 1 -->
<!-- intersections <- tibble(x = log(1.5), y = 0.5) -->
<!-- ggplot() + -->
<!--   annotate("segment", x = 0, xend = intersections$x, y = intersections$y, yend = intersections$y,  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections$x, xend = intersections$x, y = 0, yend = intersections$y,  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   stat_function(fun = abatement_cost, color = "#638ccc", size = 1.5) + -->
<!--   theme_minimal()  + -->
<!--   theme( -->
<!--     legend.position = "none", -->
<!--     title = element_text(size = 24), -->
<!--     axis.text.x = element_text(size = 24, color = "black"),  -->
<!--     axis.text.y = element_text(size = 24, color = "black"), -->
<!--     axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24), -->
<!--     panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(), -->
<!--     panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(), -->
<!--     panel.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     plot.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     axis.line = element_line(colour = "black") -->
<!--   ) + -->
<!--   labs(x = "Emissions", -->
<!--        y = "Damages", -->
<!--        title = "The damage function") + -->

<!--   scale_x_continuous(expand = c(0, 0), limits = c(0, 1),  -->
<!--                      breaks = intersections$x, labels = expression(E^"*")) + -->
<!--   scale_y_continuous(expand = c(0, 0), limits = c(0, 2),  -->
<!--                      breaks = intersections$y, labels = expression(D(E^"*"))) -->

<!-- ``` -->

<!-- ![Damage functions are increasing](`r knitr::fig_chunk("damages", "png")`) -->

<!-- ] -->



<!-- --- -->

<!-- # Marginal abatement cost -->

<!-- .pull-left[ -->
<!-- ```{r mac, echo = FALSE, fig.show = 'hide', warning = F} -->
<!-- abatement_cost <- function(x) 4 - x -->
<!-- intersections <- tibble(x = c(1,3,4), y = c(3,1,0)) -->
<!-- polys_mac <- data.frame(x = c(4, 3, 3), y = c(0, 1, 0)) -->
<!-- ggplot() + -->
<!--   geom_polygon(data = polys_mac, aes(x = x, y = y), fill = "slateblue", alpha = 0.2) + -->
<!--   annotate("segment", x = 0, xend = intersections$x[1], y = intersections$y[1], yend = intersections$y[1],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections$x[1], xend = intersections$x[1], y = 0, yend = intersections$y[1],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = 0, xend = intersections$x[2], y = intersections$y[2], yend = intersections$y[2],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections$x[2], xend = intersections$x[2], y = 0, yend = intersections$y[2],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("text", size = 8, label = "A", x = 3.25, y = 0.5) + -->
<!--   annotate("text", size = 8, label = "B", x = 1.5, y = 0.5) + -->
<!--   annotate("text", size = 8, label = "C", x = 1.5, y = 1.75) + -->
<!--   stat_function(fun = abatement_cost, color = "#638ccc", size = 1.5) + -->
<!--   theme_minimal()  + -->
<!--   theme( -->
<!--     legend.position = "none", -->
<!--     title = element_text(size = 24), -->
<!--     axis.text.x = element_text(size = 18, color = "black"),  -->
<!--     axis.text.y = element_text(size = 18, color = "black"), -->
<!--     axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24), -->
<!--     panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(), -->
<!--     panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(), -->
<!--     panel.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     plot.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     axis.line = element_line(colour = "black") -->
<!--   ) + -->
<!--   labs(x = "Emissions", -->
<!--        y = "Marginal abatement cost", -->
<!--        title = "The marginal abatement\ncost function") + -->
<!--   scale_x_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = intersections$x, labels = expression(bar(e)[1], e[1]^"*", hat(e)[1])) + -->
<!--   scale_y_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = intersections$y, labels = expression(MAC[1] (e[1]^"*"), MAC[1] (bar(e)[1]), )) -->

<!-- ``` -->

<!-- ![MACs are decreasing](`r knitr::fig_chunk("mac", "png")`) -->
<!-- ] -->

<!-- .pull-right[ -->

<!-- Marginal abatement costs are decreasing in emissions, increasing in abatement -->

<!-- The area under the MAC is total abatement cost -->

<!-- .hi-blue[A]: Total abatement cost of abating $\hat{e}_1 - e_1^*$ units -->

<!-- .hi-blue[A+B+C]: Total abatement cost of abating $\hat{e}_1 -\bar{e}_1$ units -->


<!-- ] -->


<!-- --- -->

<!-- # Marginal damages -->

<!-- .pull-left[ -->
<!-- ```{r md, echo = FALSE, fig.show = 'hide', warning = F} -->
<!-- damages <- function(x) x/3 -->
<!-- intersections <- tibble(x = c(3), y = c(1)) -->
<!-- polys_md <- data.frame(x = c(0, 3, 3), y = c(0, 1, 0)) -->
<!-- ggplot() + -->
<!--   geom_polygon(data = polys_md, aes(x = x, y = y), fill = "red", alpha = 0.2) + -->
<!--   annotate("segment", x = 0, xend = intersections$x[1], y = intersections$y[1], yend = intersections$y[1],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections$x[1], xend = intersections$x[1], y = 0, yend = intersections$y[1],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("text", size = 8, label = "F", x = 2, y = 0.25) + -->
<!--   stat_function(fun = damages, color = "#ca5670", size = 1.5) + -->
<!--   theme_minimal()  + -->
<!--   theme( -->
<!--     legend.position = "none", -->
<!--     title = element_text(size = 24), -->
<!--     axis.text.x = element_text(size = 18, color = "black"),  -->
<!--     axis.text.y = element_text(size = 18, color = "black"), -->
<!--     axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24), -->
<!--     panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(), -->
<!--     panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(), -->
<!--     panel.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     plot.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     axis.line = element_line(colour = "black") -->
<!--   ) + -->
<!--   labs(x = "Emissions", -->
<!--        y = "Marginal damages", -->
<!--        title = "The marginal damage function") + -->
<!--   scale_x_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = intersections$x, labels = expression(e[1]^"*")) + -->
<!--   scale_y_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = intersections$y, labels = expression(MD(e[1]^"*") )) -->

<!-- ``` -->

<!-- ![MACs are decreasing](`r knitr::fig_chunk("md", "png")`) -->
<!-- ] -->

<!-- .pull-right[ -->

<!-- Marginal damage curve is increasing in emissions, decreasing in abatement -->

<!-- The area under the MD is total damages -->


<!-- ] -->


<!-- --- -->

<!-- # The optimal allocation -->

<!-- .pull-left[ -->
<!-- ```{r mac_and_md, echo = FALSE, fig.show = 'hide', warning = F} -->
<!-- abatement_cost <- function(x) 4 - x -->
<!-- damages <- function(x) x/3 -->
<!-- intersections <- tibble(x = c(3), y = c(1)) -->
<!-- polys_md <- data.frame(x = c(0, 3, 3), y = c(0, 1, 0)) -->
<!-- polys_mac <- data.frame(x = c(4, 3, 3), y = c(0, 1, 0)) -->
<!-- ggplot() + -->
<!--   geom_polygon(data = polys_md, aes(x = x, y = y), fill = "red", alpha = 0.2) + -->
<!--   geom_polygon(data = polys_mac, aes(x = x, y = y), fill = "slateblue", alpha = 0.2) + -->
<!--   annotate("segment", x = 0, xend = intersections$x[1], y = intersections$y[1], yend = intersections$y[1],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections$x[1], xend = intersections$x[1], y = 0, yend = intersections$y[1],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("text", size = 8, label = "F", x = 2, y = 0.25) + -->
<!--   annotate("text", size = 8, label = "A", x = 3.25, y = 0.25) + -->
<!--   stat_function(fun = abatement_cost, color = "#638ccc", size = 1.5) + -->
<!--   stat_function(fun = damages, color = "#ca5670", size = 1.5) + -->
<!--   theme_minimal()  + -->
<!--   theme( -->
<!--     legend.position = "none", -->
<!--     title = element_text(size = 24), -->
<!--     axis.text.x = element_text(size = 18, color = "black"),  -->
<!--     axis.text.y = element_text(size = 18, color = "black", angle = 90, hjust = 0), -->
<!--     axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24), -->
<!--     panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(), -->
<!--     panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(), -->
<!--     panel.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     plot.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     axis.line = element_line(colour = "black") -->
<!--   ) + -->
<!--   labs(x = "Emissions", -->
<!--        y = "", -->
<!--        title = "The optimal allocation") + -->
<!--   scale_x_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = intersections$x, labels = expression(e[1]^"*")) + -->
<!--   scale_y_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = intersections$y, labels = expression(MAC[1] (e[1]^"*") == MD[1](e[1]^"*"))) -->

<!-- ``` -->

<!-- ![Where MAC and MD intersect minimizes total cost](`r knitr::fig_chunk("mac_and_md", "png")`) -->
<!-- ] -->

<!-- .pull-right[ -->

<!-- The optimal allocation is where MAC and MD intersect -->

<!-- This minimizes the total cost to $A+F$ -->


<!-- ] -->

<!-- --- -->

<!-- # Property rights -->

<!-- Do we need government intervention to solve environmental problems? -->

<!-- Is it possible to reach an efficient outcome with negotiation? -->

<!-- -- -->

<!-- Let's think about a special case of our model: 1 firm and 1 household -->

<!-- All damages are borne by the household, all abatement costs are borne by the firm -->

<!-- -- -->

<!-- In this setting, simply assigning property rights to the firm or household and allowing for negotiation may lead to the efficient outcome -->

<!-- --- -->

<!-- # Household ownership of pollution rights -->

<!-- Suppose the household owns the right of zero pollution, but the efficient level is greater than zero -->

<!-- How can we get to the efficient level without government intervention? -->

<!-- -- -->

<!-- The firm could propose a contract where the household accepts some pollution, in exchange for a transfer payment -->

<!-- --- -->

<!-- # Household ownership of pollution rights -->

<!-- The firm could propose a contract $(E,\theta)$ consisting of emissions $E$ and transfer payment $\theta$ -->

<!-- -- -->

<!-- The household can then accept or refuse the contract -->

<!-- -- -->

<!-- Negotiating is costly and has its own transactions cost $tr$ -->

<!-- -- -->

<!-- We assume both players are fully informed about each others preferences and technologies -->

<!-- --- -->

<!-- # Household ownership of pollution rights -->

<!-- When does the household accept the contract? -->

<!-- -- -->

<!-- The household is weakly better off accepting the contract if $\theta \geq D(E)$ (assuming the firm bears the transactions costs $tr$) -->

<!-- -- -->

<!-- What contract does the firm offer in equilibrium? -->

<!-- --- -->

<!-- # Household ownership of pollution rights -->

<!-- The firm will choose to offer $\theta = D(E)$, the least amount required for the household to accept -->

<!-- -- -->

<!-- This means we can write the firm's total cost as: -->
<!-- $$TC(E) = C(E) + D(E) + tr$$ -->
<!-- It's optimal choice of $E$ (and therefore $\theta$) is given by the first-order condition: -->
<!-- $$-C'(E) = D'(E)$$ -->
<!-- which matches our condition for efficiency in pollution -->

<!-- Note that this still requires $tr \leq C(0) - [C(E^*) + D(E^*)]$ -->

<!-- --- -->

<!-- # Firm ownership of pollution rights -->

<!-- If the firm has the rights to pollution, we just flip the script -->

<!-- The household proposes a contract $(E, \theta)$ where the firm reduces pollution in exchange for a transfer payment  -->

<!-- The firm accepts or rejects the contract -->

<!-- When does the firm accept the contract? -->

<!-- -- -->

<!-- The firm accepts if $\theta \geq C(E)$ -->

<!-- -- -->

<!-- The household will then offer the minimum required: $\theta = C(E)$ -->

<!-- --- -->

<!-- # Firm ownership of pollution rights -->

<!-- The household's utility maximization problem is then: -->
<!-- $$\max_{E} y - (D(E) + C(E) + tr)$$ -->
<!-- giving the first-order condition: -->
<!-- $$-C'(E) = D'(E)$$ -->
<!-- where we again reach the social optimum, as long as: -->
<!-- $$tr \leq D(\hat{E}) - (D(E^*) + C(E^*))$$ -->
<!-- where $\hat{E}$ is the firms initial emission level -->

<!-- --- -->

<!-- # The Coase theorem -->

<!-- These two examples showed that if properties rights are clearly defined and the affected parties can negotiate, private contracts between rational agents can achieve the efficient pollution level -->

<!-- -- -->

<!-- The assignment of property rights .hi-red[doesn't] matter for efficiency -->

<!-- But it does matter for the distribution of wealth -->

<!-- -- -->

<!-- These observations are known as the .hi-blue[Coase Theorem] -->

<!-- --- -->

<!-- # The Coase theorem -->

<!-- > Suppose party A imposes an externality on party B. Provided transactions costs are sufficiently small, irrespective of the initial allocation of property rights: the parties can achieve the socially optimal level of pollution $E^*$ using a transfer payment $\theta$ where both parties are at least as well off as they were before -->

<!-- With small enough transactions costs, the party that does not own the property rights can propose a contract that is mutually beneficial -->

<!-- --- -->

<!-- # The Coase theorem: real world -->

<!-- The Coase theorem is not just a useful theoretical exercise: -->

<!-- Vittel sells luxury mineral water -->

<!-- In 1988 nitrates from farm runoff was polluting their water supply -->

<!-- Vittel contacted all upstream farmers and negotiated contracts for reducing nitrogen runoff -->

<!-- .footnote[Depres, Christophe, Gilles Grolleau, and Naoufel Mzoughi. "Contracting for Environmental Property Rights: The Case of Vittel." Economica, New Series, 75, no. 299 (2008)] -->


<!-- --- -->

<!-- # The Coase theorem: real world -->

<!-- Mines de Potasse d'Alsace heavily polluted the Rhine river with chlorides in potash production -->

<!-- MdPA was thought to have low abatement costs because filtering out chlorides is cheap -->

<!-- In 1972 Switzerland, France, Germany, and the Netherlands contracted to pay MdPA 532 million francs to reduce emissions -->

<!-- .footnote[Dieperink, C. International water negotiations under asymmetry, Lessons from the Rhine chlorides dispute settlement (1931–2004). Int Environ Agreements 11, 139–157 (2011).] -->

<!-- --- -->

<!-- # Policy instruments -->

<!-- The Coase theorem is applicable for local environmental with few and easily identifiable parties -->

<!-- -- -->

<!-- Many of the largest environmental pollution problems have many parties, households and polluters, making transactions costs likely to be very high -->

<!-- -- -->

<!-- This is where there is a role for public intervention -->

<!-- --- -->

<!-- # Command and control -->

<!-- .hi-blue[Command and control] policies require polluting firms to carry out prescribed pollution-reducing actions -->

<!-- -- -->

<!-- The most common CC policies are emissions limits and technology standards -->
<!--   - Standard on absolute amount of emissions -->
<!--   - Standard on emissions per unit of output, input, etc -->
<!--   - Must install $SO_2$ scrubber -->
<!--   - Must use best available technology -->

<!-- -- -->

<!-- We're going to focus on absolute emissions standards -->

<!-- --- -->

<!-- # Emission standard -->

<!-- The simplest policy is one where the regulator requires all firms to emit no more than their socially optimal level $e_j^*$ -->

<!-- -- -->

<!-- By definition, this policy achieves the efficient level of pollution at minimum total abatement cost -->

<!-- -- -->

<!-- Is this a realistic option? -->

<!-- Why or why not? -->

<!-- --- -->

<!-- # Emission standard -->

<!-- Firm-specific emission standards aren't very realistic -->

<!-- -- -->

<!-- The regulator needs to know: -->
<!--   - The social damage function -->
<!--   - Each firm's abatement cost function -->

<!-- And the regulator needs to be able to impose a policy that is differential across firms and is unlikely to be politically feasible -->

<!-- --- -->

<!-- # Uniform emission standard -->

<!-- An alternative is to impose a .hi-blue[uniform emission standard] such that $e_j \leq \bar{e}$ for all firms $j$ -->

<!-- We could imagine setting $\bar{e} = E^*/J$ where $E^*$ is the socially efficient level of emissions -->

<!-- -- -->

<!-- If firms are identical this achieves the efficient outcome -->

<!-- -- -->

<!-- If they're not identical it won't -->


<!-- --- -->

<!-- # Uniform standard costs -->

<!-- .pull-left[ -->
<!-- ```{r uniform_standard, echo = FALSE, fig.show = 'hide', warning = F} -->
<!-- abatement_cost_1 <- function(x) 4 - x -->
<!-- abatement_cost_2 <- function(x) 2 - x -->
<!-- intersections <- tibble(x = c(1, 1, 1), y = c(0, 1, 3)) -->
<!-- polys_mac1 <- tibble(x = c(2, 1, 1), y = c(0, 0, 1)) -->
<!-- polys_mac2 <- tibble(x = c(4, 1, 1), y = c(0, 0, 3)) -->
<!-- ggplot() + -->
<!--   geom_polygon(data = polys_mac2, aes(x = x, y = y), fill = "slateblue", alpha = 0.2) + -->
<!--   geom_polygon(data = polys_mac1, aes(x = x, y = y), fill = "red", alpha = 0.2) + -->
<!--   annotate("segment", x = intersections$x[1], xend = intersections$x[3], y = intersections$y[1], yend = intersections$y[3],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = 0, xend = intersections$x[1], y = intersections$y[2], yend = intersections$y[2],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = 0, xend = intersections$x[1], y = intersections$y[3], yend = intersections$y[3],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   stat_function(fun = abatement_cost_1, color = "#638ccc", size = 1.5) + -->
<!--   stat_function(fun = abatement_cost_2, color = "#ca5670", size = 1.5) + -->
<!--   theme_minimal()  + -->
<!--   theme( -->
<!--     legend.position = "none", -->
<!--     title = element_text(size = 24), -->
<!--     axis.text.x = element_text(size = 18, color = "black"),  -->
<!--     axis.text.y = element_text(size = 18, color = "black"), -->
<!--     axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24), -->
<!--     panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(), -->
<!--     panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(), -->
<!--     panel.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     plot.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     axis.line = element_line(colour = "black") -->
<!--   ) + -->
<!--   labs(x = "Emissions", -->
<!--        y = "", -->
<!--        title = "Not efficient if\nMACs aren't equal") + -->
<!--   scale_x_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = intersections$x, labels = expression(bar(e), "", "")) + -->
<!--   scale_y_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = intersections$y, labels = expression("", MAC[1]( bar(e) ), MAC[2]( bar(e) ))) -->

<!-- ``` -->

<!-- ![MACs are not equal under a uniform standard](`r knitr::fig_chunk("uniform_standard", "png")`) -->
<!-- ] -->

<!-- .pull-right[ -->

<!-- Even though $\bar{e}\times J =  E^*$, the MACs may not be equal -->

<!-- If MACs aren't equal we can maintain $E^*$ and reduce costs -->

<!-- How? -->

<!-- Firm 1 abates 1 unit more, firm 2 abates 1 unit less -->

<!-- ] -->


<!-- --- -->

<!-- # Uniform standard costs -->

<!-- .pull-left[ -->
<!-- ```{r uniform_standard_2, echo = FALSE, fig.show = 'hide', warning = F} -->
<!-- abatement_cost_1 <- function(x) 4 - x -->
<!-- abatement_cost_2 <- function(x) 2 - x -->
<!-- intersections_1 <- tibble(x = c(0, 0.5, 0.5),  -->
<!--                         y = c(1.5, 1.5, 0)) -->
<!-- intersections_2 <- tibble(x = c(0, 1.5, 1.5),  -->
<!--                         y = c(2.5, 2.5, 0)) -->
<!-- polys_mac1 <- tibble(x = c(2, 0.5, 0.5),  -->
<!--                      y = c(0, 1.5, 0)) -->
<!-- polys_mac2 <- tibble(x = c(4, 1.5, 1.5),  -->
<!--                      y = c(0, 2.5, 0)) -->
<!-- ggplot() + -->
<!--   geom_polygon(data = polys_mac2, aes(x = x, y = y), fill = "slateblue", alpha = 0.2) + -->
<!--   geom_polygon(data = polys_mac1, aes(x = x, y = y), fill = "red", alpha = 0.2) + -->
<!--   annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[2], y = intersections_1$y[1], yend = intersections_1$y[2],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3], y = intersections_1$y[2], yend = intersections_1$y[3],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections_2$x[1], xend = intersections_2$x[2], y = intersections_2$y[1], yend = intersections_2$y[2],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections_2$x[2], xend = intersections_2$x[3], y = intersections_2$y[2], yend = intersections_2$y[3],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   stat_function(fun = abatement_cost_1, color = "#638ccc", size = 1.5) + -->
<!--   stat_function(fun = abatement_cost_2, color = "#ca5670", size = 1.5) + -->
<!--   theme_minimal()  + -->
<!--   theme( -->
<!--     legend.position = "none", -->
<!--     title = element_text(size = 24), -->
<!--     axis.text.x = element_text(size = 18, color = "black"),  -->
<!--     axis.text.y = element_text(size = 18, color = "black"), -->
<!--     axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24), -->
<!--     panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(), -->
<!--     panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(), -->
<!--     panel.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     plot.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     axis.line = element_line(colour = "black") -->
<!--   ) + -->
<!--   labs(x = "Emissions", -->
<!--        y = "", -->
<!--        title = "Not efficient if\nMACs aren't equal") + -->
<!--   scale_x_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = c(intersections_1$x[2], intersections_2$x[2]), labels = expression(e[1], e[2])) + -->
<!--   scale_y_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = c(intersections_1$y[2], intersections_2$y[2]), labels = expression(MAC[1]( e[1] ), MAC[2]( e[2] ))) -->

<!-- ``` -->

<!-- ![MACs are not equal under a uniform standard](`r knitr::fig_chunk("uniform_standard_2", "png")`) -->
<!-- ] -->

<!-- .pull-right[ -->

<!-- Now firm 1, the lower MAC firm, is abating more than firm 2 -->

<!-- This changed the total abatement cost -->

<!-- ] -->


<!-- --- -->

<!-- # Uniform standard costs -->

<!-- .pull-left[ -->
<!-- ```{r uniform_standard_3, echo = FALSE, fig.show = 'hide', warning = F} -->
<!-- abatement_cost_1 <- function(x) 4 - x -->
<!-- abatement_cost_2 <- function(x) 2 - x -->
<!-- intersections_1 <- tibble(x = c(0, 0.5, 0.5),  -->
<!--                         y = c(1.5, 1.5, 0)) -->
<!-- intersections_2 <- tibble(x = c(0, 1.5, 1.5),  -->
<!--                         y = c(2.5, 2.5, 0)) -->
<!-- polys_mac1 <- tibble(x = c(1, 1, 0.5, 0.5),  -->
<!--                      y = c(0, 1, 1.5, 0)) -->
<!-- polys_mac2 <- tibble(x = c(1, 1, 1.5, 1.5),  -->
<!--                      y = c(0, 3, 2.5, 0)) -->
<!-- ggplot() + -->
<!--   geom_polygon(data = polys_mac2, aes(x = x, y = y), fill = "slateblue", alpha = 0.2) + -->
<!--   geom_polygon(data = polys_mac1, aes(x = x, y = y), fill = "red", alpha = 0.2) + -->
<!--   annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[2], y = intersections_1$y[1], yend = intersections_1$y[2],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3], y = intersections_1$y[2], yend = intersections_1$y[3],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections_2$x[1], xend = intersections_2$x[2], y = intersections_2$y[1], yend = intersections_2$y[2],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   annotate("segment", x = intersections_2$x[2], xend = intersections_2$x[3], y = intersections_2$y[2], yend = intersections_2$y[3],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   stat_function(fun = abatement_cost_1, color = "#638ccc", size = 1.5) + -->
<!--   stat_function(fun = abatement_cost_2, color = "#ca5670", size = 1.5) + -->
<!--   theme_minimal()  + -->
<!--   theme( -->
<!--     legend.position = "none", -->
<!--     title = element_text(size = 24), -->
<!--     axis.text.x = element_text(size = 18, color = "black"),  -->
<!--     axis.text.y = element_text(size = 18, color = "black"), -->
<!--     axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24), -->
<!--     panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(), -->
<!--     panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(), -->
<!--     panel.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     plot.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     axis.line = element_line(colour = "black") -->
<!--   ) + -->
<!--   labs(x = "Emissions", -->
<!--        y = "", -->
<!--        title = "Gains and losses from\nshifting abatement")  + -->
<!--   scale_x_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = c(intersections_1$x[2], intersections_2$x[2]), labels = expression(e[1], e[2])) + -->
<!--   scale_y_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = c(intersections_1$y[2], intersections_2$y[2]), labels = expression(MAC[1]( e[1] ), MAC[2]( e[2] ))) -->

<!-- ``` -->

<!-- ![Gains from equating macs](`r knitr::fig_chunk("uniform_standard_3", "png")`) -->
<!-- ] -->

<!-- .pull-right[ -->

<!-- Now firm 1, the lower MAC firm, is abating more than firm 2 -->

<!-- This changed the total abatement cost -->

<!-- Firm 1 has costs .hi-red[increase] -->

<!-- Firm 2 has costs .hi-blue[decrease] -->

<!-- Net effect is a .hi-blue[decrease] in costs -->

<!-- ] -->


<!-- --- -->

<!-- # Uniform standard costs -->

<!-- .pull-left[ -->
<!-- ```{r uniform_standard_4, echo = FALSE, fig.show = 'hide', warning = F} -->
<!-- abatement_cost_1 <- function(x) 4 - x -->
<!-- abatement_cost_2 <- function(x) 2 - x -->
<!-- intersections <- tibble(x = c(0, 2),  -->
<!--                         y = c(2, 2)) -->
<!-- polys_mac1 <- tibble(x = c(1, 1, 0, 0),  -->
<!--                      y = c(0, 1, 2, 0)) -->
<!-- polys_mac2 <- tibble(x = c(1, 1, 2, 2),  -->
<!--                      y = c(0, 3, 2, 0)) -->
<!-- polys_mac3 <- tibble(x = c(1, 2, 2, 1),  -->
<!--                      y = c(3, 2, 1, 2)) -->
<!-- ggplot() + -->
<!--   geom_polygon(data = polys_mac2, aes(x = x, y = y), fill = "slateblue", alpha = 0.2) + -->
<!--   geom_polygon(data = polys_mac1, aes(x = x, y = y), fill = "red", alpha = 0.2) + -->
<!--   geom_polygon(data = polys_mac3, aes(x = x, y = y), linetype = "dotted", size = 2, alpha = 0.7) + -->
<!--   annotate("segment", x = intersections$x[1], xend = intersections$x[2], y = intersections$y[1], yend = intersections$y[2],  -->
<!--            linetype = "dashed", size = 1, color = "black") + -->
<!--   stat_function(fun = abatement_cost_1, color = "#638ccc", size = 1.5) + -->
<!--   stat_function(fun = abatement_cost_2, color = "#ca5670", size = 1.5) + -->
<!--   theme_minimal()  + -->
<!--   theme( -->
<!--     legend.position = "none", -->
<!--     title = element_text(size = 24), -->
<!--     axis.text.x = element_text(size = 18, color = "black"),  -->
<!--     axis.text.y = element_text(size = 18, color = "black", angle = 90, hjust = 1), -->
<!--     axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24), -->
<!--     panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(), -->
<!--     panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(), -->
<!--     panel.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     plot.background = element_rect(fill = "#eeeeee", colour = NA), -->
<!--     axis.line = element_line(colour = "black") -->
<!--   ) + -->
<!--   labs(x = "Emissions", -->
<!--        y = "", -->
<!--        title = "Gains and losses from\nshifting abatement")  + -->
<!--   scale_x_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = c(intersections_1$x[2], intersections_2$x[2]), labels = expression(e[1], e[2])) + -->
<!--   scale_y_continuous(expand = c(0, 0), limits = c(0, 5),  -->
<!--                      breaks = c(intersections_2$y[2]), labels = expression(MAC[1]( e[1] ) == MAC[2]( e[2] ))) -->

<!-- ``` -->

<!-- ![Gains from equating macs](`r knitr::fig_chunk("uniform_standard_4", "png")`) -->
<!-- ] -->

<!-- .pull-right[ -->

<!-- We can keep obtaining cost reductions until MACs are equal across firms -->

<!-- With net reductions in deadweight loss equal to the dark gray area (light blue minus light red) -->

<!-- We want low MAC firms to abate more than high MAC firms -->


<!-- ] -->

---

# Emission taxes

An emission tax is a fee that the pollution firm must pay per unit of effluent

These aren't very common in the US but are more widespread in Europe

--

Suppose the government imposes a tax of size $\tau$ per unit of pollution

---

# Emission taxes

The firm's problem is then to minimize total pollution-related costs:
$$TC_j(e_j) = C_j(e_j) + \tau e_j$$

--

The first-order conditions are then:
$$-C'_j(e_j) = \tau$$

--

The firm's optimal choice is to set marginal abatement cost equal to the tax rate

--

The firm reduces emissions as long as the cost of emissions reductions is less than the alternative: paying the tax

---

# Emission taxes

.pull-left[
```{r taxes, echo = FALSE, fig.show = 'hide', warning = F}
abatement_cost <- function(x) 4 - x
intersections <- tibble(x = c(2,2), y = c(2,0))
polys_mac <- data.frame(x = c(4, 2, 2), y = c(0, 2, 0))
polys_tax <- data.frame(x = c(0, 0, 2, 2), y = c(0, 2, 2, 0))
ggplot() +
  geom_polygon(data = polys_mac, aes(x = x, y = y), fill = "slateblue", alpha = 0.2) +
  geom_polygon(data = polys_tax, aes(x = x, y = y), fill = "red", alpha = 0.2) +
  annotate("segment", x = 0, xend = intersections$x[1], y = intersections$y[1], yend = intersections$y[1],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections$x[1], xend = intersections$x[1], y = 0, yend = intersections$y[1],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = 0, xend = intersections$x[2], y = intersections$y[2], yend = intersections$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections$x[2], xend = intersections$x[2], y = 0, yend = intersections$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("text", size = 8, label = "B", x = 2.5, y = 1) +
  annotate("text", size = 8, label = "A", x = 1, y = 1) +
  stat_function(fun = abatement_cost, color = "#638ccc", size = 1.5) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 20, color = "black"),
    axis.text.y = element_text(size = 20, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "",
       title = "Costs under taxation") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = intersections$x[1], labels = expression(e[j](tau))) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = intersections$y[1], labels = expression(tau))

```

![MACs are decreasing](`r knitr::fig_chunk("taxes", "png")`)
]

.pull-right[

Under a tax $\tau$, the emission choice is a function of the tax: $e_j(\tau)$

The firm pays total tax $A$ and incurs abatement cost $B$

Now the government has revenue $\tau \times e_j(\tau)$ that it can use for different purposes


]

---

# Emission taxes

If all firms face the same marginal tax rate, what does the firm first-order condition imply?

--

$$-C'_j(e_j) \equiv MAC_j(e_j) = \tau \,\, \forall j$$

--

This then implies that:
$$MAC_j(e_j) = MAC_j(e_k) \,\, \forall j,k$$

Marginal abatement costs across firms are equal and we have obtained the given emissions reduction at least-cost

---

# Emission taxes

If we change the tax rate what do we expect to happen to emissions?

--

Return to the firm FOC:
$$-C'_j(e_j) = \tau$$

and differentate it with respect to $\tau$ and recognize that $e_j$ is a function of $\tau$:
$$-C''_j(e_j) {de_j(\tau) \over d\tau} = 1$$

--

This gives us that: ${de_j(\tau) \over d\tau} = {1 \over -C''_j(e_j)} < 0$: higher taxes lower emissions if MACs are decreasing in emissions

---

# Auctioned pollution permits