---
title: "Lecture 4"
subtitle: "Imperfect information"
author: Ivan Rudik
date: AEM 6510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'


---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, tidylog, xaringanExtra, rlang, patchwork
)
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

---

# Roadmap

What happens when the regulator has imperfect information about:
  - Marginal abatement costs?
  - Marginal damages?

So we far have assumed perfect information which is unlikely to be true

--

We will continue assuming that firms know their own marginal abatement cost

And that regulators observe firm-level emissions

---

# Imperfect information

How do we think about this?

--

Firms are unlikely to tell regulators exactly what their marginal abatement costs are

--

Therefore regulators will have to try to estimate it when designing policy

--

Estimation will naturally result in some error

---

# Price vs quantities (Weitzman 1974)

Let's start by comparing emissions taxes and tradable permits

--

We'll start by looking at damage function uncertainty with known abatement costs

--

Then we look at abatement cost uncertainty with known damage functions

--

We will mainly be focused on efficiency outcomes and want to understand which policy delivers the highest welfare and why

---

# Damage function uncertainty

- $D(E)$ is the social damage function
--

- $C(E)$ is the aggregate abatement cost function
--

- $E^*$ is the optimal level of pollution
  - This is defined by $-C'(E) = D'(E)$
--
  
- Let $\tilde{D}(E)$ and $\tilde{C}(E)$ denote the estimated damage and abatement cost function

--

First suppose the regulator estimates $-C'(E)$ correctly, but underestimates marginal damages: $\tilde{D}'(E) < D'(E) \,\, \forall E$

---

# Damage function uncertainty

The regulator chooses policy based on her estimate $\tilde{D}'(E)$

--

This means she sets:
$$-C'(E) = \tilde{D}'(E)$$

--

which is solved by some $\tilde{E} > E^*$ because she underestimated marginal damages

--

What is the welfare loss from targeting $\tilde{E}$ instead of $E^*$?

Does the size of the loss depend on the policy instrument?

---

# Damage function uncertainty

Define welfare loss as the difference in total social costs at $\tilde{E}$ versus the efficient level $E^*$:
\begin{align}
WL &= [D(\tilde{E}) + C(\tilde{E})] - [D(E^*) + C(E^*)] \notag\\
  &= [D(\tilde{E}) -D(E^*)] + [ C(\tilde{E}) - C(E^*)]
\end{align}

This is equivalent to the area under the marginal damage and abatement cost curves between the two emission levels:
$$WL = \int_{E^*}^{\tilde{E}}D'(E) dE - \int_{E^*}^{\tilde{E}} - C'(E)dE$$


---

# Damage function uncertainty with permits


.pull-left[
```{r uncertain_damages_permit_nodwl, echo = FALSE, fig.show = 'hide', warning = F}
damage_1 <- function(x) 1 + x
damage_2 <- function(x) -1 + x
abatement_cost <- function(x) 5 - x
intersections_1 <- tibble(x = c(0, 2, 2, 3, 3, 3),
                        y = c(2, 3, 0, 4, 2, 0))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
ggplot() +
  stat_function(fun = damage_1, color = "#638ccc", size = 1.5) +
  stat_function(fun = damage_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = abatement_cost, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = "D'(E)", x = 3, y = 4.5) +
  annotate("text", size = 8, label = expression(tilde(D)*minute (E)), x = 4, y = 2.5) +
  annotate("text", size = 8, label = expression(-C*minute (E)), x = 4.5, y = 1) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "#eeeeee"),
    axis.text.y = element_text(size = 18, color = "#eeeeee"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "MAC/MD ($)",
       title = "Uncertain MD with permits") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5))

```

![Uncertain MD with permits](`r knitr::fig_chunk("uncertain_damages_permit_nodwl", "png")`)
]

.pull-right[

Here is the set up

Solve for $\tilde{E}, E^*$ and the welfare loss from setting the number of permits to be $L = \tilde{E}$

]

---

# Damage function uncertainty with permits


.pull-left[
```{r uncertain_damages_permit, echo = FALSE, fig.show = 'hide', warning = F}
damage_1 <- function(x) 1 + x
damage_2 <- function(x) -1 + x
abatement_cost <- function(x) 5 - x
intersections_1 <- tibble(x = c(0, 2, 2, 3, 3, 3),
                        y = c(2, 3, 0, 4, 2, 0))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
ggplot() +
  geom_polygon(data = polys_dwl, aes(x = x, y = y), fill = "red", alpha = 0.2) +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[5],
           y = intersections_1$y[1], yend = intersections_1$y[5],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[6],
           y = intersections_1$y[4], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = damage_1, color = "#638ccc", size = 1.5) +
  stat_function(fun = damage_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = abatement_cost, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = "A", x = 2.6, y = 3) +
  annotate("text", size = 8, label = "D'(E)", x = 3, y = 4.5) +
  annotate("text", size = 8, label = expression(tilde(D)*minute (E)), x = 4, y = 2.5) +
  annotate("text", size = 8, label = expression(-C*minute (E)), x = 4.5, y = 1) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "MAC/MD ($)",
       title = "Uncertain MD with permits") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4]), labels = expression(E^"*", tilde(E))) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[1]), labels = expression(sigma(tilde(E))))

```

![Uncertain MD with permits](`r knitr::fig_chunk("uncertain_damages_permit", "png")`)
]

.pull-right[

With a permit scheme the regulator fixes the total amount of emissions at $\tilde{E}$

Since she underestimates $D'$, she lets firms emit too much

She incurs welfare loss $A$ from emissions where marginal damage $>$ marginal abatement cost

]


---

# Damage function uncertainty with taxes


.pull-left[
```{r uncertain_damages_permit_nodwl_tax, echo = FALSE, fig.show = 'hide', warning = F}
damage_1 <- function(x) 1 + x
damage_2 <- function(x) -1 + x
abatement_cost <- function(x) 5 - x
intersections_1 <- tibble(x = c(0, 2, 2, 3, 3, 3),
                        y = c(2, 3, 0, 4, 2, 0))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
ggplot() +
  stat_function(fun = damage_1, color = "#638ccc", size = 1.5) +
  stat_function(fun = damage_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = abatement_cost, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = "D'(E)", x = 3, y = 4.5) +
  annotate("text", size = 8, label = expression(tilde(D)*minute (E)), x = 4, y = 2.5) +
  annotate("text", size = 8, label = expression(-C*minute (E)), x = 4.5, y = 1) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "#eeeeee"),
    axis.text.y = element_text(size = 18, color = "#eeeeee"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "MAC/MD ($)",
       title = "Uncertain MD with taxes") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5))

```

![Uncertain MD with permits](`r knitr::fig_chunk("uncertain_damages_permit_nodwl_tax", "png")`)
]

.pull-right[

Here is the set up

Solve for $\tilde{E}, E^*$ and $E(\tau)$ which is the firm's choice of emissions given $\tau$, and the welfare loss from setting the tax to be $\tau(\tilde{E})$ which achieves $E=\tilde{E}$ given $-\tilde{C}'(E)$

]

---

# Damage function uncertainty with taxes


.pull-left[
```{r uncertain_damages_tax, echo = FALSE, fig.show = 'hide', warning = F}
damage_1 <- function(x) 1 + x
damage_2 <- function(x) -1 + x
abatement_cost <- function(x) 5 - x
intersections_1 <- tibble(x = c(0, 2, 2, 3, 3, 3),
                        y = c(2, 3, 0, 4, 2, 0))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
ggplot() +
  geom_polygon(data = polys_dwl, aes(x = x, y = y), fill = "red", alpha = 0.2) +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[5],
           y = intersections_1$y[1], yend = intersections_1$y[5],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[6],
           y = intersections_1$y[4], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = damage_1, color = "#638ccc", size = 1.5) +
  stat_function(fun = damage_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = abatement_cost, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = "D'(E)", x = 3, y = 4.5) +
  annotate("text", size = 8, label = expression(tilde(D)*minute (E)), x = 4, y = 2.5) +
  annotate("text", size = 8, label = expression(-C*minute (E)), x = 4.5, y = 1) +
  annotate("text", size = 8, label = "A", x = 2.6, y = 3) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "MAC/MD ($)",
       title = "Uncertain MD with taxes") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4]), labels = expression(E^"*", tilde(E))) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[1]), labels = expression(tau(tilde(E))))

```

![Uncertain MD with taxes](`r knitr::fig_chunk("uncertain_damages_tax", "png")`)
]

.pull-right[

The regulator sets the tax as a function of her target emissions $\tau(\tilde{E})$

Since she underestimates $D'$, she sets $\tau(\tilde{E})$ too low

The firm then selects $E(\tau) = \tilde{E}$

She incurs welfare loss $A$ from emissions where marginal damage $>$ marginal abatement cost



]


---

# Damage function uncertainty

What did we learn?

--

With damage function uncertainty,  
the regulator will set the incorrect level of policy

--

If she underestimates MD:
--

  - She sets the total number of permits too high
  - She sets too low of a tax

--

Both lead to the exact same welfare loss  
so both policies have the same efficiency

---

# Abatement cost function uncertainty

Suppose the regulator estimates $D'(E)$ correctly and underestimates marginal abatement cost: $-\tilde{C}'(E) < -C'(E) \,\, \forall E$


The regulator chooses policy based on her estimate $-\tilde{C}'(E)$

--

This means she sets:
$$-\tilde{C}'(E) = D'(E)$$
which is solved by some $\tilde{E} < E^*$ because she underestimated marginal costs

--

What is the welfare loss from targeting $\tilde{E}$ instead of $E^*$?


---

# Abatement cost function uncertainty


.pull-left[
```{r uncertain_abatement_nodwl, echo = FALSE, fig.show = 'hide', warning = F}
damage <- function(x) 1 + x
abatement_cost_1 <- function(x) 5 - x
abatement_cost_2 <- function(x) 3 - x

intersections_1 <- tibble(x = c(0, 2, 2, 3, 3, 3, 1, 1, 0),
                        y = c(2, 3, 0, 4, 2, 0, 0, 4, 4))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
polys_dwl_2 <- tibble(x = c(2, 1, 1),
                     y = c(3, 4, 2))
ggplot() +
  stat_function(fun = abatement_cost_1, color = "#638ccc", size = 1.5) +
  stat_function(fun = abatement_cost_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = damage, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = expression(D*minute (E)), x = 3, y = 4.5) +
  annotate("text", size = 8, label = expression(-C*minute (E)), x = 4.5, y = 1) +
  annotate("text", size = 8, label = expression(-tilde(C)*minute (E)), x = 0.5, y = 3.25) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "#eeeeee"),
    axis.text.y = element_text(size = 18, color = "#eeeeee"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "MAC/MD ($)",
       title = "Uncertain MAC") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5))

```


![Uncertain MAC breaks the equivalence between taxes and permits](`r knitr::fig_chunk("uncertain_abatement_nodwl", "png")`)
]

.pull-right[

Here's the uncertain MAC problem

Solve for $\tilde{E}, E^*, \sigma(\tilde{E}),$ and the welfare loss from setting the number of permits to be $L = \tilde{E}$

Solve for $\tilde{E}, E^*,$ and $E(\tau)$ which is the firm's choice of emissions given $\tau$, and the welfare loss from setting the tax to be $\tau(\tilde{E})$

]



---

# Abatement cost function uncertainty


.pull-left[
```{r uncertain_abatement, echo = FALSE, fig.show = 'hide', warning = F}
damage <- function(x) 1 + x
abatement_cost_1 <- function(x) 5 - x
abatement_cost_2 <- function(x) 3 - x

intersections_1 <- tibble(x = c(0, 2, 2, 3, 3, 3, 1, 1, 0),
                        y = c(2, 3, 0, 4, 2, 0, 0, 4, 4))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
polys_dwl_2 <- tibble(x = c(2, 1, 1),
                     y = c(3, 4, 2))
ggplot() +
  geom_polygon(data = polys_dwl, aes(x = x, y = y), fill = "red", alpha = 0.2) +
  geom_polygon(data = polys_dwl_2, aes(x = x, y = y), fill = "blue", alpha = 0.2) +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[5],
           y = intersections_1$y[1], yend = intersections_1$y[5],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[6],
           y = intersections_1$y[4], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[7], xend = intersections_1$x[8],
           y = intersections_1$y[7], yend = intersections_1$y[8],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[9], xend = intersections_1$x[8],
           y = intersections_1$y[9], yend = intersections_1$y[8],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = abatement_cost_1, color = "#638ccc", size = 1.5) +
  stat_function(fun = abatement_cost_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = damage, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = expression(D*minute (E)), x = 3, y = 4.5) +
  annotate("text", size = 8, label = expression(-C*minute (E)), x = 4.5, y = 1) +
  annotate("text", size = 8, label = expression(-tilde(C)*minute (E)), x = 0.5, y = 3.25) +
  annotate("text", size = 8, label = "B", x = 2.6, y = 3) +
  annotate("text", size = 8, label = "C", x = 1.4, y = 3) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "MAC/MD ($)",
       title = "Uncertain MAC") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[7]), labels = expression(E^"*", E(tau), tilde(E))) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[1], intersections_1$y[9]), labels = expression(tau(tilde(E)), sigma(tilde(E))))

```


![Uncertain MAC breaks the equivalence between taxes and permits](`r knitr::fig_chunk("uncertain_abatement", "png")`)
]

.pull-right[

With permits, the regulator allows $\tilde{E}$ permits which results in a permit price of $\sigma(\tilde{E})$ where $\tilde{E}$ intersects the true MAC

This yields a welfare loss of $C$

Firm behavior sets the price even though quantity is fixed by the regulator

]


---

# Abatement cost function uncertainty


.pull-left[
![Uncertain MAC breaks the equivalence between taxes and permits](`r knitr::fig_chunk("uncertain_abatement", "png")`)
]

.pull-right[

With a tax, the regulator sets a price $\tau(\tilde{E})$ per unit of emissions, and the firms choose the quantity of emissions where $\tau(\tilde{E}) = -C'(E)$ which causes total emissions to be $E(\tau)$

This yields a welfare loss of $B$

Firm behavior sets the quantity even though price is fixed by the regulator

]

---

# Abatement cost function uncertainty


.pull-left[
![Uncertain MAC breaks the equivalence between taxes and permits](`r knitr::fig_chunk("uncertain_abatement", "png")`)
]

.pull-right[

Since $E(\tau) \neq \tilde{E}$, abatement cost uncertainty matters: tradable permits and taxes give us different emission outcomes

Is there any systematic difference in the efficiency properties of permits and taxes?

]

---

# Abatement cost function uncertainty


In general it seems like sometimes the welfare loss of taxes will be higher, and other times the welfare losses of permits will be higher

--

What we will do next is try to understand the characteristics of the MAC and MD curves that tend to drive one policy to be better than the other


---

# Steep versus flat MD

```{r prices_vs_quantities_nodwl, echo = FALSE, fig.show = 'hide', warning = F, fig.width=12, fig.height=5}
damage_steep <- function(x) -1.6 + 1.75*x
abatement_cost_1 <- function(x) 5 - x
abatement_cost_2 <- function(x) 3 - x

intersections_1 <- tibble(x = c(0.0, 2.4, 2.4, 5-1.328, 5-1.328, 5-1.328, 1.672, 1.672, 0.0),
                        y =   c(1.328, 2.6, 0.0, -1.6+1.75*(5-1.328), 1.328, 0.0, 0.0, 3.4, 3.4))
polys_dwl <- tibble(x = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[5]),
                     y = c(intersections_1$y[2], intersections_1$y[4], intersections_1$y[5]))
polys_dwl_2 <- tibble(x = c(intersections_1$x[2], intersections_1$x[7], intersections_1$x[8]),
                     y = c(intersections_1$y[2], intersections_1$y[1], intersections_1$y[8]))
steep <- ggplot() +
  stat_function(fun = abatement_cost_1, color = "#638ccc", size = 1.5) +
  stat_function(fun = abatement_cost_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = damage_steep, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = expression(D*minute (E)), x = 3, y = 4.5) +
  annotate("text", size = 8, label = expression(-C*minute (E)), x = 4.5, y = 1.15) +
  annotate("text", size = 8, label = expression(-tilde(C)*minute (E)), x = 0.5, y = 3.05) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "#eeeeee"),
    axis.text.y = element_text(size = 18, color = "#eeeeee"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "MAC/MD ($)",
       title = "Steep MD") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[7]), labels = expression(E^"*", E(tau), tilde(E))) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[1], intersections_1$y[9]), labels = expression(tau(tilde(E)), sigma(tilde(E))))


damage_flat <- function(x) 1 + x/2
abatement_cost_1 <- function(x) 5 - x
abatement_cost_2 <- function(x) 3 - x

intersections_1 <- tibble(x = c(0.0, 8/3, 8/3, 5-(3-4/3), 5-(3-4/3), 5-(3-4/3), 4/3, 4/3, 0.0),
                        y =   c(3-4/3, 5-8/3, 0.0, 1+(5-(3-4/3))/2, 3-4/3, 0.0, 0.0, 5-4/3, 5-4/3))
polys_dwl <- tibble(x = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[5]),
                     y = c(intersections_1$y[2], intersections_1$y[4], intersections_1$y[5]))
polys_dwl_2 <- tibble(x = c(intersections_1$x[2], intersections_1$x[7], intersections_1$x[8]),
                     y = c(intersections_1$y[2], intersections_1$y[1], intersections_1$y[8]))
flat <- ggplot() +
  stat_function(fun = abatement_cost_1, color = "#638ccc", size = 1.5) +
  stat_function(fun = abatement_cost_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = damage_flat, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = expression(D*minute (E)), x = 4, y = 3.5) +
  annotate("text", size = 8, label = expression(-C*minute (E)), x = 4.5, y = 1.15) +
  annotate("text", size = 8, label = expression(-tilde(C)*minute (E)), x = 0.5, y = 3.05) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "#eeeeee"),
    axis.text.y = element_text(size = 18, color = "#eeeeee"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "MAC/MD ($)",
       title = "Flat MD") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[7]), labels = expression(E^"*", E(tau), tilde(E))) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[1], intersections_1$y[9]), labels = expression(tau(tilde(E)), sigma(tilde(E))))

steep + flat & theme(
    plot.background = element_rect(fill = "#eeeeee", colour = NA))

```

![Initial distribution matters for how costs and benefits are spread out](`r knitr::fig_chunk("prices_vs_quantities_nodwl", "png")`)

Solve for the permit and tax DWLs in both of these scenarios where the only difference is the steepness of the marginal damage curve


---

# Steep versus flat MD

```{r prices_vs_quantities, echo = FALSE, fig.show = 'hide', warning = F, fig.width=12, fig.height=5}
damage_steep <- function(x) -1.6 + 1.75*x
abatement_cost_1 <- function(x) 5 - x
abatement_cost_2 <- function(x) 3 - x

intersections_1 <- tibble(x = c(0.0, 2.4, 2.4, 5-1.328, 5-1.328, 5-1.328, 1.672, 1.672, 0.0),
                        y =   c(1.328, 2.6, 0.0, -1.6+1.75*(5-1.328), 1.328, 0.0, 0.0, 3.4, 3.4))
polys_dwl <- tibble(x = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[5]),
                     y = c(intersections_1$y[2], intersections_1$y[4], intersections_1$y[5]))
polys_dwl_2 <- tibble(x = c(intersections_1$x[2], intersections_1$x[7], intersections_1$x[8]),
                     y = c(intersections_1$y[2], intersections_1$y[1], intersections_1$y[8]))
steep <- ggplot() +
  geom_polygon(data = polys_dwl, aes(x = x, y = y), fill = "red", alpha = 0.2) +
  geom_polygon(data = polys_dwl_2, aes(x = x, y = y), fill = "blue", alpha = 0.2) +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[5],
           y = intersections_1$y[1], yend = intersections_1$y[5],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[6],
           y = intersections_1$y[4], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[7], xend = intersections_1$x[8],
           y = intersections_1$y[7], yend = intersections_1$y[8],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[9], xend = intersections_1$x[8],
           y = intersections_1$y[9], yend = intersections_1$y[8],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = abatement_cost_1, color = "#638ccc", size = 1.5) +
  stat_function(fun = abatement_cost_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = damage_steep, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = expression(D*minute (E)), x = 3, y = 4.5) +
  annotate("text", size = 8, label = expression(-C*minute (E)), x = 4.5, y = 1.15) +
  annotate("text", size = 8, label = expression(-tilde(C)*minute (E)), x = 0.5, y = 3.05) +
  annotate("text", size = 8, label = "A", x = 3.15, y = 2.5) +
  annotate("text", size = 8, label = "B", x = 1.9, y = 2.5) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "MAC/MD ($)",
       title = "Steep MD") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[7]), labels = expression(E^"*", E(tau), tilde(E))) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[1], intersections_1$y[9]), labels = expression(tau(tilde(E)), sigma(tilde(E))))


damage_flat <- function(x) 1 + x/2
abatement_cost_1 <- function(x) 5 - x
abatement_cost_2 <- function(x) 3 - x

intersections_1 <- tibble(x = c(0.0, 8/3, 8/3, 5-(3-4/3), 5-(3-4/3), 5-(3-4/3), 4/3, 4/3, 0.0),
                        y =   c(3-4/3, 5-8/3, 0.0, 1+(5-(3-4/3))/2, 3-4/3, 0.0, 0.0, 5-4/3, 5-4/3))
polys_dwl <- tibble(x = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[5]),
                     y = c(intersections_1$y[2], intersections_1$y[4], intersections_1$y[5]))
polys_dwl_2 <- tibble(x = c(intersections_1$x[2], intersections_1$x[7], intersections_1$x[8]),
                     y = c(intersections_1$y[2], intersections_1$y[1], intersections_1$y[8]))
flat <- ggplot() +
  geom_polygon(data = polys_dwl, aes(x = x, y = y), fill = "red", alpha = 0.2) +
  geom_polygon(data = polys_dwl_2, aes(x = x, y = y), fill = "blue", alpha = 0.2) +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[5],
           y = intersections_1$y[1], yend = intersections_1$y[5],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[6],
           y = intersections_1$y[4], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[7], xend = intersections_1$x[8],
           y = intersections_1$y[7], yend = intersections_1$y[8],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[9], xend = intersections_1$x[8],
           y = intersections_1$y[9], yend = intersections_1$y[8],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = abatement_cost_1, color = "#638ccc", size = 1.5) +
  stat_function(fun = abatement_cost_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = damage_flat, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = expression(D*minute (E)), x = 4, y = 3.5) +
  annotate("text", size = 8, label = expression(-C*minute (E)), x = 4.5, y = 1.15) +
  annotate("text", size = 8, label = expression(-tilde(C)*minute (E)), x = 0.5, y = 3.05) +
  annotate("text", size = 8, label = "A", x = 3.15, y = 2.25) +
  annotate("text", size = 8, label = "B", x = 1.9, y = 2.5) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Emissions",
       y = "MAC/MD ($)",
       title = "Flat MD") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[7]), labels = expression(E^"*", E(tau), tilde(E))) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[1], intersections_1$y[9]), labels = expression(tau(tilde(E)), sigma(tilde(E))))

steep + flat & theme(
    plot.background = element_rect(fill = "#eeeeee", colour = NA))

```

![Initial distribution matters for how costs and benefits are spread out](`r knitr::fig_chunk("prices_vs_quantities", "png")`)

The only difference between the two plots is how steep the MD is, and subsequently the policy that the regulator sets

Welfare loss for taxes is given by $A$, welfare loss for permits is given by $B$


---

# Steep versus flat MD


![Initial distribution matters for how costs and benefits are spread out](`r knitr::fig_chunk("prices_vs_quantities", "png")`)

Permits do better with steep MD, taxes do better with flat MD!

Why?


---

# Steep versus flat MD

With a steep (and known) MD:
--

- The difference between $E^*$ and $\tilde{E}$ is restricted: steeper curves are more inelastic, errors in estimating MAC lead to small errors in $\tilde{E}-E^*$
- The difference between $E^*$ and $E(\tau)$ can be very large: errors in estimating the MAC lead to bigger errors in $\tau(\tilde{E}) - \tau(E^*)$ because MD is very steep


---

# Steep versus flat MD

With a flat (and known) MD:
--

- The difference between $E^*$ and $\tilde{E}$ is can be very large: flatter curves are more elastic, errors in estimating MAC lead to large errors in $\tilde{E}-E^*$
- The difference between $E^*$ and $E(\tau)$ is restricted: errors in estimating the MAC lead to smaller errors in $\tau(\tilde{E}) - \tau(E^*)$ because MD is very flat

<!-- --- -->

<!-- # The Weitzman theorem -->

<!-- So far we've focused on the intuition, now lets formalize the results -->

<!-- -- -->

<!-- Let the damage function be $D(E,\eta)$ where $\eta$ is a random variable representing the regulator's uncertainty about the damage function -->

<!-- -- -->

<!-- We assume $D_E, D_{EE} > 0$, and that $D_\eta, D_{E\eta} > 0$ -->

<!-- -- -->

<!-- Increases in $\eta$ shift up the damage and marginal damage curves -->

<!-- --- -->

<!-- # The Weitzman theorem -->


<!-- Let the abatement cost function be $C(E,\epsilon)$ where $\epsilon$ is a random variable representing the regulator's uncertainty about the abatement cost function -->

<!-- -- -->

<!-- We assume $-C_E > 0, -C_{EE} > 0$, and that $-C_\epsilon, -C_{E\epsilon} > 0$ -->

<!-- -- -->

<!-- Increases in $\epsilon$ shift up the total and marginal abatement cost curves -->

<!-- --- -->

<!-- # The Weitzman theorem: quantities -->

<!-- Let's start with a quantity instrument like tradable permits -->

<!-- -- -->

<!-- The regulator's objective is to find the emission target that minimizes social cost: -->
<!-- $$\min_{E} -->
