---
title: "Lecture 12"
subtitle: "Health // Difference-in-differences"
author: Ivan Rudik
date: AEM 6510
output: 
  html_document:
  theme: flatly
  highlight: haddock
  code_folding: show
  toc: yes
  toc_depth: 4
  toc_float: yes
  keep_md: true
---

<style type="text/css">
.main-container {
  max-width: 1000px;
  margin-left: auto;
  margin-right: auto;
}
</style>
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, dpi = 300)
```

## R packages 

Required packages: `ggthemes, fixest, broom, tidyverse`

```{r libs}
# load required packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  ggthemes, fixest, broom, tidyverse
)
```

```{r, echo = FALSE}
options(dplyr.summarise.inform = FALSE)
```

# Hands on: Intro to difference-in-differences

## Generating fake data

To see how DD works in practice we will first generate a fake dataset where **we know the true value of what we want to estimate.** 
This is a good skill to have in order to make sure you understand how methods work, and that your code does what you want it to do.
If you know what the result is and your code produces something different, your code is wrong.

To generate fake data we need to model the **true data generating process.** 
We will make this exactly the DD model, but with some added noise.

### Random numbers
Any time you introduce randomness into your code, you will want to set a random number generator seed so it is reproducible.

```{r}
# Initialize RNG
set.seed(123456)
```

Every time you run this notebook you will get the exact same results even though we are drawing "random" numbers.

### The model parameters

The model generating our data will be very simple:
1. A time-invariant factor specific to both NY and PA
2. A period-specific factor common across both NY and PA
3. A treatment effect in NY after the fake policy is implemented
4. Some added random noise in outcomes

First we need to define the time-invariant factors.

```{r}
# Fixed state characteristics: ny has more biodiversity
state_fes <- tribble(~state, ~state_fe,
                     "NY", 88,
                     "PA", 44)
```

Here we used a `tribble`, which is just a `tibble` but where we can write it out element-by-element.
This is helpful when you need to initialize small data frames.

Next we need the common, period-specific factors.

```{r}
# Period characteristics: biodiversity is declining
period_fes <- tribble(~period, ~period_fe,
                     "before", 341,
                     "after", 207)
```

Last, we need to define the size of the treatment effect.

```{r}
# True effect of the policy: how many additional species are saved?
treatment_effect <- 55
```

Now we have defined all the parameters required to generate our data. 
Next, we need to *actually* generate the data.
This takes a few steps.

### Generating the data

First, we need to generate N observations for each combination of before/after and NY/PA. We can do this with `crossing`.

```{r}
# All combos of states, periods, and observation numbers
bio_df <- crossing(
  state = c("NY", "PA"),
  period = c("before", "after"),
  obs = 1:500
)
bio_df
```

Second, we need to bring in the time-invariant, and period-specific factors with `join` functions.

```{r}
bio_df <- bio_df %>%
  inner_join(period_fes) %>% 
  inner_join(state_fes)
bio_df
```

Third, it will be helpful to define a variable for units that are treated: New York and after.

```{r}
bio_df <- bio_df %>% 
  mutate(treated = state == "NY" & period == "after",
         period = factor(period, levels = c("before", "after")),
         period = relevel(period, "before"))
bio_df
```

Now we need to generate our data by adding the time-invariant effect `state_fe`, the period-specific effect `period_fe`, and some random noise given by `rnorm`. If the observation is New York after the policy we will also add `treatment_effect`.

```{r}
bio_df <- bio_df %>% 
  mutate(
    biodiversity = ifelse(
      treated,
      period_fe + state_fe + treatment_effect + 100*rnorm(n()),
      period_fe + state_fe  + 100*rnorm(n())
    )
  )
bio_df
```

Let's plot the data. New York is in black, Pennsylvania is in orange.

```{r}

# Switch the order of the period variables so before comes first
bio_df$period = factor(bio_df$period, levels = c("before", "after"))

ggplot(bio_df, group = interaction(state, period)) +
  geom_jitter(aes(x = period, y = biodiversity, color = interaction(state), shape = interaction(state)), size = 3) +
  annotate("text", size = 8, label = "NY", x = 1, y = 700) +
  annotate("text", size = 8, label = "PA", x = 1, y = 100, color = "orange") +
  scale_color_colorblind() +
  theme_minimal() +
  scale_x_discrete(labels = c("Before Treatment", "After Treatment")) + 
  scale_y_continuous(limits = c(50, 700)) +
  labs(
    x = "Time",
    y = "Biodiversity",
    title = "The raw fake data"
  ) +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 24), axis.text.y = element_text(size = 24),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line = element_line(colour = "black")
  ) 
```

Next we can estimate the effect using DD.

```{r}
# Outcome ~ treatment + period fixed effect + state fixed effect, data
estimates <- fixest::feols(
  biodiversity ~ treated + period + state, 
  data = bio_df
  ) %>% 
  broom::tidy()
estimates
```

Compare the true values (left) versus the estimates (right). **DD correctly recovers the true treatment effect!**

```{r}
c(treatment_effect, estimates$estimate[2]) # treatment_effect
c(period_fes$period_fe[period_fes$period == "after"] - period_fes$period_fe[period_fes$period == "before"], estimates$estimate[3]) # period_after
c(state_fes$state_fe[state_fes$state == "PA"] - state_fes$state_fe[state_fes$state == "NY"], estimates$estimate[4]) # statePA

```

### DD addresses time-invariant and common period-specific factors

Now let's look at what happens if take more naive approaches that do not address time-invariant factors, or common-period specific factors.
We will look at how the failure to address these things biases our results in predictible ways.

Suppose we just regressed biodiversity on the existence of the conservation policy.

```{r}
# Outcome ~ treatment, data
fixest::feols(
  biodiversity ~ treated, 
  data = bio_df
  ) %>% 
  broom::tidy()
```

We **underestimate the true effect** because we did not difference out the common decline in biodiversity over time.
The estimated effect of the policy in NY is confounded by the common decline in biodiversity across both states.

What if we controlled for these common, period-specific factors?

```{r}
# Outcome ~ treatment + period fixed effects, data
fixest::feols(
  biodiversity ~ treated + period, 
  data = bio_df
  ) %>% 
  broom::tidy()
```

Now we **overestimate the true effect**.
We (still) did not difference out the fact that NY tends to have higher levels of biodiversity.
The estimated treatment effect is biased upward because NY is just more biodiverse even without the policy.


# Hands on: Hollingsworth and Rudik (2021)

Now lets work through the HR data to see how they got their results.
We will just start with the cleaned data instead of raw.
Cleaning the data is a huge chunk of the actual work of doing research, but takes a lot of code/time.

First, load the data and look at it.
```{r}
########################################
## Read in blood lead data
bll_df <- read_csv("data/hr_2021_bll.csv")

########################################
## Look at data
bll_df
```

What are the variables?

- `state`: state fips code
- `county`: county fips code
- `state_county`: full fips code

### The raw data

First, let's plot the raw data.
This is something you should generally always do.
It lets you see what is actually happening, whether you made data cleaning errors, etc.
If what you're trying to estimate pops up in the raw data that's a good sign.

Here we will be plotting the raw data with two changes:
1. Only plot for 2005-2009. After 2009 some counties stop reporting lead poisoning, so the composition of the sample changes, making the comparison tricky.
2. Plot the *weighted mean* lead poisoning rate where we weight by the number of children tested. The idea here is that counties with more testing report estimated rates that less noisy measures of the true rate. We want these counties to count more toward our nationwide average.

```{r}
bll_df %>% 
  group_by(race, year) %>% 
  summarise(percent_high_conditional_tested = weighted.mean(percent_high_conditional_tested, weight, na.rm = T)) %>% 
  filter(year >= 2005 & year <= 2009) %>% 
  ggplot(aes(x = year, y = percent_high_conditional_tested)) +
    geom_vline(xintercept = 2006.5, color = "grey55", linetype = "dashed") +
    geom_point(aes(shape = as.factor(race), color = as.factor(race)), size = 5) +
    annotate("text", size = 4, label = "Treated", x = 2005.25, y = 2.85) +
    annotate("text", size = 4, label = "Border", x = 2005.25, y = 2.5) +
    annotate("text", size = 4, label = "Control", x = 2005.25, y = 1.4) +
    annotate("text", size = 6, label = "Leaded Fuel", x = 2005.50, y = 0.25) +
    annotate("text", size = 6, label = "Unleaded Fuel", x = 2008, y = 0.25) +
    theme_minimal() +
    scale_color_colorblind() +
    labs(
      x = "Year",
      y = "Percent Lead Poisoned",
      title = "Average lead poisoning rates by type (balanced panel)"
    ) +
    theme(
      legend.position = "none",
      title = element_text(size = 14),
      axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 14),
      axis.title.x = element_text(size = 14), axis.title.y = element_text(size = 14),
      panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.line = element_line(colour = "black")
    ) 
```

### Difference-in-differences

```{r}
########################################
## Difference-in-differences estimate
fixest::feols(
  ihs_bll ~ as.factor(race)*I(year > 2007) + 
    as.factor(state)*year + 
    unemp_rate + median_income + percent_non_white + lead_emitted + ap | state_county + year, 
  weights = ~weight,
  data = bll_df) %>% 
  broom::tidy(cluster = "state")
```

### Event study

```{r}
########################################
## Event study

# Need to factor year so we can omit 2007
bll_df$year <- factor(bll_df$year, ordered = FALSE)
bll_df$year <- relevel(bll_df$year, 3)

# Estimate event study
event_study <- fixest::feols(
  ihs_bll ~ as.factor(race)*year + 
    as.factor(state)*year + 
    unemp_rate + median_income + percent_non_white + lead_emitted + ap | state_county + year, 
  weights = ~weight,
  data = bll_df) %>% 
  tidy(cluster = "state") %>% 
  filter(str_detect(term, "^as.factor\\(race\\)2:year")) %>% 
  add_row(estimate = 0, std.error = 0, .after = 2) %>% 
  mutate(year = row_number() + 2004)

# Plot event study
ggplot(data = event_study, aes(x = year, y = estimate)) +
  geom_hline(yintercept = 0, size = 0.5) +
  geom_vline(xintercept = 2006.5, color = "grey55", linetype = "dashed") +
  geom_ribbon(aes(ymin = estimate - 1.96*std.error, ymax = estimate + 1.96*std.error), 
              fill = "darkslateblue", alpha = 0.4) +
  geom_line(size = 1) +
  geom_point(size = 5) +
  annotate("text", size = 4, label = "Leaded Fuel", x = 2005.50, y = -0.26) +
  annotate("text", size = 4, label = "Unleaded Fuel", x = 2012, y = -0.26) +
  theme_minimal() +
  scale_color_colorblind() +
  scale_x_continuous(breaks = seq(2005, 2015, 2)) +
  labs(
    x = "Year",
    y = "Percent Lead Poisoned",
    title = "Percent lead poisoned relative to 2007 (first unleaded year)"
  ) +
  theme(
    legend.position = "none",
    title = element_text(size = 14),
    axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 14), axis.title.y = element_text(size = 14),
    panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line = element_line(colour = "black")
  )


########################################
## Read in mortality data
mortality_df <- read_csv("data/hr_2021_mortality.csv")

mortality_df



# Need to factor year so we can omit 2007
mortality_df$year <- factor(mortality_df$year, ordered = FALSE)
mortality_df$year <- relevel(mortality_df$year, 9)

# Estimate event study
event_study <- fixest::feols(
  age_adjusted_rate ~ as.factor(race)*year + 
    as.factor(state)*year + 
    unemp_rate + median_income + percent_non_white + lead_emitted + ap | state_county + year, 
  weights = ~weight,
  data = mortality_df) %>% 
  broom::tidy(cluster = "state") %>% 
  filter(str_detect(term, "^as.factor\\(race\\)2:year")) %>% 
  add_row(estimate = 0, std.error = 0, .after = 8) %>% 
  mutate(year = row_number() + 1998)

# Plot event study
ggplot(data = event_study, aes(x = year, y = estimate)) +
  geom_hline(yintercept = 0, size = 0.5) +
  geom_vline(xintercept = 2006.5, color = "grey55", linetype = "dashed") +
  geom_ribbon(aes(ymin = estimate - 1.96*std.error, ymax = estimate + 1.96*std.error), 
              fill = "darkslateblue", alpha = 0.4) +
  geom_line(size = 1) +
  geom_point(size = 5) +
  annotate("text", size = 4, label = "Leaded Fuel", x = 2002, y = -200) +
  annotate("text", size = 4, label = "Unleaded Fuel", x = 2012, y = -200) +
  theme_minimal() +
  scale_color_colorblind() +
  scale_x_continuous(breaks = seq(1999, 2015, 2)) +
  labs(
    x = "Year",
    y = "Age-Adjusted Mortality Rate (deaths per 100,000)",
    title = "Mortality rate relative to 2007 (first unleaded year)"
  ) +
  theme(
    legend.position = "none",
    title = element_text(size = 14),
    axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 14), axis.title.y = element_text(size = 14),
    panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line = element_line(colour = "black")
  )
```