---
title: "Lecture 7"
subtitle: "Environmental policy with pre-existing distortions"
author: Ivan Rudik
date: AEM 6510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'


---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, tidylog, xaringanExtra, rlang, patchwork
)
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

---

# Roadmap

So far we have looked at single sector economies with:
- Pollution distortions
- Competitive markets
- Market power distortions

--

Now we will learn about multi-sector economies

How does environmental policy spillover into these other sectors?

How does environmental policy interact with revenue-raising taxes (e.g. income taxes)?

---

# Environmental policy with leisure

First we extend the model so that labor supply is .hi-blue[elastic]
  - Households have a choice of either working or leisure time
  
--

To focus on the key intuition we assume: $^1$
  - There is a representative (single) firm
  - There is a representative household

--

This allows us to treat individual and aggregate behavior the same

.footnote[1: The underlying critical assumption is that utility and profit functions take what's called a Gorman form.]

---

# Environmental policy with leisure

Define the following:
- $X$ is consumption of the polluting good
- $Z$ is consumption of the *numeraire* good (i.e. the relative good)
- $N$ is the hours of leisure time
- $E$ is aggregate emissions

--

Utility is then:
$$U(X,Z,N,E)  = U(X,N) + Z - D(E)$$

--

where $U_{XX}, U_{NN} < 0$ and  $U_{XX}U_{NN}-U_{NX}^2 > 0$ and the person is endowed with some amount of time $T$ to allocate between work and leisure

---

# Environmental policy with leisure

Wages earned by the household are $w$, and we assume demand for labor is perfectly elastic

--

i.e. demand is horizontal at $w$

--

Household income is then: $w\cdot(T-N)$

--

We can now write the households utility maximization problem as:
$$\max_{X,N,Z} U(X,Z,N,E)  = U(X,N) + Z - D(E) \\
\text{ subject to: } \,\,\,\, w\cdot(T-N) = Z + pX$$

Substitute the budget constraint in for $Z$ to get an unconstrained problem

---

# Environmental policy with leisure

We can substitute the budget constraint in for $Z$ to get an unconstrained problem:
$$\max_{X,N} U(X,Z,N,E)  = U(X,N) + w\cdot(T-N) - pX - D(E)$$

--

with FOCs:
$$U_X = p \qquad U_N = w$$
which implicitly define the demand function for consumption $X(p,w)$ and the demand function for leisure $N(p,w)$

---

# Environmental policy with leisure

How do choices of $X,N$ respond to a change in price $p$?

--

Differentiate both FOCs with respect to p:

--

$$U_{XX} {\partial X \over \partial p} + U_{XN} {\partial N \over \partial p} = 1 \qquad U_{NN} {\partial N \over \partial p} + U_{XN} {\partial X \over \partial p} = 0$$

--

We have two equations and two unknowns so we can solve to get:
$${\partial N \over \partial p} = {-U_{XN} \over U_{XX}U_{NN} - U_{XN}^2} \qquad {\partial X \over \partial p} = {U_{NN} \over U_{XX}U_{NN} - U_{XN}^2}$$

---

# Environmental policy with leisure

$${\partial N \over \partial p} = {-U_{XN} \over U_{XX}U_{NN} - U_{XN}^2} \qquad {\partial X \over \partial p} = {U_{NN} \over U_{XX}U_{NN} - U_{XN}^2}$$

--

${\partial X \over \partial p}$ is negative since $U$ is concave $(U_{NN} < 0, U_{XX}U_{NN} - U_{XN}^2 > 0)$

--

The sign of ${\partial N \over \partial p}$ equals the sign on $-U_{XN}$

--

If $X$ and $N$ are substitutes, $-U_{XN} > 0$, and leisure increases in the price of the consumption good

--

If they are complements, $-U_{XN} < 0$, and leisure decreases in the price of the consumption good

---

# Environmental policy with leisure

If $N$ is going on a picnic and $X$ is hot dogs: $X$ and $N$ are complements

--

If the price of hot dogs goes up 1000% then you will go on fewer picnics

--

If $N$ is going on a picnic and $X$ is video games: $X$ and $N$ are substitutes

--

If the price of video games go up 1000% then you will go on more picnics


---

# Environmental policy with leisure

The firm side of the economy will be the same as before: it produces $X$ and emits $E$ and for simplicity we will focus on the specific case:
$$\Pi = pX - C(X) \text{ where } E = \delta X$$

--

We will also assume:

- $\delta=1$ so we can use $E$ and $X$ interchangeably
--

- $C'(X) > 0, C''(X) \geq 0$
--

- The polluting industry's demand for labor is small relative to the entire economy, i.e. wages are effectively fixed for the household

---

# Environmental policy with leisure

Now lets solve for the social optimum:
$$\max_{X} W = \underbrace{U(X,N) + w\cdot(T-N) - pX - D(X)}_{\text{Consumer Utility}} + \underbrace{pX - C(X)}_{\text{Firm profit}}$$
To focus on interactions with non-regulated industries, we assume the regulator cannot determine the allocation of leisure and labor

--

The consumer chooses $N$ according to the FOC $U_N(X^*,N) = w$ and then $Z$ given the budget constraint $Z = w(T-N) - pX^*$

--

One way you can think about this is as if the regulator imposes a quantity standard $X^*$ and then a market price $p^*$ arises which affects leisure demand

---

# Environmental policy with leisure

The FOC for the optimum is:
$$U_X - D'(X) - C'(X) + [U_N -w]{\partial N \over \partial X} = 0$$
where the last term captures the households .hi-blue[indirect] leisure response to the regulator's policy choice

--

Given household utility maximization $U_N -w = 0$ and the condition is then:
$$U_X - C'(X) = D'(X)$$

--

Marginal abatement cost $(U_X - C'(X))$ equals marginal damage $(D'(X))$ !

---

# Environmental policy with labor market distortions

.hi[Elastic labor supply/leisure doesn't change the efficiency condition]

--

Now suppose the government needs to raise revenue with a labor income tax $m$ in order to finance government services

--

It needs to finance a budget of size $G$

--

The consumer's utility maximization problem is:
\begin{gather}
  \max_{X,Z,N} U = u(X,N) + Z - D(E) \\
  \text{ subject to } (1-m)w(T-N) = Z + pX
\end{gather}

Where the budget is scaled down by $(1-m)$ reflecting the income tax

---

# Environmental policy with labor market distortions

The FOCs are:
$$u_X = p \qquad u_N = (1-m)w$$

--

.hi-blue[The labor tax causes an inefficiency in the labor market:] the marginal value of leisure $(u_N)$ is no longer equal to the marginal value of labor $(w)$

--

$u_N < w$ which means that the household overconsumes leisure

--

Another way to see this is to re-write the FOC as:
$$u_N + mw = w$$

--

The tax $m$ makes the consumer act as if there is a subsidy $mw$ on leisure

---

# Environmental policy with labor market distortions

.pull-left[
```{r labor_market_1, echo = FALSE, fig.show = 'hide', warning = F}

# demand <- function(x) 5 - x
demand <- function(x) 3.5 - x
intersections_1 <- tibble(x = c(5, 5, 0, 0, 5, 2.5, 2.5, 1.5, 1.5),
                        y = c(0, 2, 2, 1, 1, 0, 2, 0, 2))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
ggplot() +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[2],
           y = intersections_1$y[1], yend = intersections_1$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "solid", size = 1, color = "firebrick") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[5],
           y = intersections_1$y[4], yend = intersections_1$y[5],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[6], xend = intersections_1$x[7],
           y = intersections_1$y[6], yend = intersections_1$y[7],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[8], xend = intersections_1$x[9],
           y = intersections_1$y[8], yend = intersections_1$y[9],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = demand, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = expression(N(p^c,w)), x = 0.75, y = 3.5) +
  annotate("text", size = 8, label = "b", x = 2.25, y = 1.75) +
  annotate("text", size = 8, label = "G", x = 4, y = 1.5) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#ffffff", colour = NA),
    plot.background = element_rect(fill = "#ffffff", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Leisure",
       y = "Wage ($)",
       title = "Income taxation causes DWL") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[6]), labels = expression(T, 0, N^c)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[2], intersections_1$y[4], intersections_1$y[6]), labels = expression(w, "(1-m)w", ))

```


![excess burden is because MU of leisure is less than wage](`r knitr::fig_chunk("labor_market_1", "png")`)
]

.pull-right[

$w$ is the perfectly elastic demand for labor

$N^c$ is how much leisure the consumer chooses, since $(1-m)w < w$ this is too much and induces DWL equal to $b$

This is called .hi-red[excess burden]

The tax raises revenues equal to $G$: $mw\times(T-N^c)$

]


---

# Environmental policy with labor market distortions

.pull-left[
```{r labor_market_2, echo = FALSE, fig.show = 'hide', warning = F}

demand_2 <- function(x) 5 - x
demand <- function(x) 3.5 - x
intersections_1 <- tibble(x = c(5, 5, 0, 0, 5, 2.5, 2.5, 4, 4),
                        y = c(0, 2, 2, 1, 1, 0, 2, 2, 0))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
ggplot() +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[2],
           y = intersections_1$y[1], yend = intersections_1$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "solid", size = 1, color = "firebrick") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[5],
           y = intersections_1$y[4], yend = intersections_1$y[5],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[6], xend = intersections_1$x[7],
           y = intersections_1$y[6], yend = intersections_1$y[7],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[8], xend = intersections_1$x[9],
           y = intersections_1$y[8], yend = intersections_1$y[9],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = demand, color = "darkslateblue", size = 1.5) +
  stat_function(fun = demand_2, color = "#ca5670", size = 1.5) +
  annotate("text", size = 8, label = expression(N(p^c,w)), x = 0.75, y = 3.5) +
  annotate("text", size = 8, label = expression(N(p^"*",w)), x = 2.25, y = 3.5) +
  annotate("text", size = 8, label = "e", x = 3, y = 1.5) +
  annotate("text", size = 8, label = "c", x = 3.75, y = 1.75) +
  annotate("text", size = 8, label = "d", x = 4.5, y = 1.5) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#ffffff", colour = NA),
    plot.background = element_rect(fill = "#ffffff", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Leisure",
       y = "Wage ($)",
       title = "Reducing emissions has\ntwo other effects") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[6], intersections_1$x[9]), labels = expression(T, 0, N^c, N^"*")) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[2], intersections_1$y[4], intersections_1$y[6]), labels = expression(w, "(1-m)w", ))

```


![excess burden is because MU of leisure is less than wage](`r knitr::fig_chunk("labor_market_2", "png")`)
]

.pull-right[

Suppose $N$ and $X$ are substitutes, and the regulator sets $X=X^*$ where $X^* \rightarrow MAC = MD$

This raises the price of $X$, shifts leisure demand to the .hi-red[right]

New DWL is $c$, and government revenues are now only $d$

Change in DWL from $X^c \rightarrow X^*$ is indeterminant

]



---

# Environmental policy with labor market distortions

.pull-left[

![excess burden is because MU of leisure is less than wage](`r knitr::fig_chunk("labor_market_2", "png")`)
]

.pull-right[

Fixing the pollution externality had two effects:

1. Indeterminant effect on the distortion in the labor market
2. Reduced the amount of revenue the government raised through labor taxation

]

---

# Second-best environmental policy

What does the optimal environmental policy look like if there's a pre-existing labor market distortion?

--

The government has a budget $G$ it needs to finance using labor taxes or emission taxes

--

First let's consider the case where they can only raise revenue via a labor tax: this is non-revenue raising environmental policy

---

# Second-best non-revenue raising environmental policy

If we cannot raise revenue with the environmental policy, the regulator chooses $X$ (and $E$) and the marginal tax rate $m$ to maximize the sum of profit and utility, subject to the budget constraint

--

The household consumes leisure according to the FOC: 
$$U_N(\bar{X},N) = (1-m)w$$ 
given the regulator chose $X=\bar{X}$

--

The firm obtains profits:
$$\Pi = p\bar{X} - C(\bar{X})$$

---

# Second-best non-revenue raising environmental policy

The marginal value of the dirty good comes from the consumers inverse demand:
$$P(\bar{X}) = u_X(\bar{X},N)$$
which depends on $N$

--

First we need to learn how the endogenous variables $N$ and $p$ vary with $\bar{X}$

--

Let's do the comparative statics: differentiate the consumer's two FOCs with respect to $\bar{X}$

---

# Second-best non-revenue raising environmental policy

\begin{align}
  u_{XX}\frac{\partial \bar{X}}{\partial \bar{X}} + u_{XN} \frac{\partial N}{\partial \bar{X}} &= \frac{\partial p}{\partial \bar{X}} \tag{X FOC} \\
  u_{NX}\frac{\partial \bar{X}}{\partial \bar{X}} + u_{NN} \frac{\partial N}{\partial \bar{X}} &= 0 \tag{N FOC}
\end{align}

$\frac{\partial \bar{X}}{\partial \bar{X}} = 1$ so two equations, two unknowns; 
--
solving the system gives us:
\begin{align}
 \frac{\partial N}{\partial \bar{X}} &= - {u_{XN} \over u_{NN}} \\
 \frac{\partial p}{\partial \bar{X}} &= {u_{XX}u_{NN} - u_{NN}^2 \over u_{NN}} < 0
\end{align}
$\text{sign}(\frac{\partial N}{\partial \bar{X}})$ depends on whether $X$ and $N$ are complements or substitutes

---

# Second-best non-revenue raising environmental policy

Now that we know how the firm responds, return to the regulator's problem:
$$\max_{X,m} u(X,N) + Z - D(X) + pX - C(X) \quad \text{ s.t. } \quad wm(T-N) = G$$

--

We still need to decide what the government does with its revenue

--

For convenience, we assume its returned to the consumer as a lump sum transfer so that:
$$Z=(1-m)w(T-N)-pX+G = (1-m)w(T-N)-pX+wm(T-N)\\ \Rightarrow Z =  w(T-N) - pX$$
Income is unchanged for a given level of $N$ under the tax and transfer


---

# Second-best non-revenue raising environmental policy

The regulator's problem is then:
$$\max_{X,m} u(X,N) +  \underbrace{w(T-N)}_{Z} - D(X) - C(X) + \lambda[ wm(T-N)  -G]$$
$\lambda$ is called the .hi[marginal welfare cost of public funds]

--

It measures the welfare cost of raising revenue by taxing labor

--

What's the FOC for $m$?

---

# Second-best non-revenue raising environmental policy

The FOC for $m$ is:
$$(u_N - w){\partial N \over \partial m} + \lambda \left[w(T-N)-wm {\partial N \over \partial m} \right] = 0$$

--

The household's optimal choice of $N$ tells us that: $-mw = u_N - w$, we can substitute this in to get $\lambda$:

--

$$\lambda = \frac{w  m {\partial N \over \partial m}}{w(T-N)-wm{\partial N \over \partial m}}$$

--

Whats the interpretation of the right hand side?

---

# Second-best non-revenue raising environmental policy

$$\lambda = \frac{w  m {\partial N \over \partial m}}{w(T-N)-wm{\partial N \over \partial m}}$$

--

The numerator is:

--

The welfare cost of changing $m$

--

Why?

--

Higher $m$ increases leisure demand ${\partial N \over \partial m}$

--

This times the .hi[tax wedge] $mw$, the gap between $w$ and actual wage after taxes, gives us the change in excess burden (i.e. the DWL $d$ in the graph)

---

# Second-best non-revenue raising environmental policy

$$\lambda = \frac{w  m {\partial N \over \partial m}}{w(T-N)-wm{\partial N \over \partial m}}$$

The denominator is:

--

The change in tax revenue from higher $m$

--

First term is the increase in revenue on the inframarginal hours worked

--

Second term is the decrease in revenue from reduced hours worked
  - Similar to $P(X) + P'(X)X$ for a monopolist

---

# Second-best non-revenue raising environmental policy

$$\lambda = \frac{w  m {\partial N \over \partial m}}{w(T-N)-wm{\partial N \over \partial m}}$$

Numerator and denominator combined give us:

The change in welfare from a change in $m$
--
 over 
--
the change in tax revenue from a change in $m$

--

This is the change in welfare from a change in tax revenue!

---

# Second-best non-revenue raising environmental policy

Now consider the FOC for $X$:
$$u_X - D'(X) - C'(X) + \left[ u_N - w - \lambda w m \right] {\partial N \over \partial X} = 0$$

--

Recall that we know:
$$-wm = u_N - w \qquad \frac{\partial N}{\partial X} = \frac{\partial N}{\partial p}\frac{\partial p}{\partial X}$$

So that we can substitute in the consumer leisure response:

--

$$u_X - C'(X) + (1+\lambda)\left[-{\partial N \over \partial p}\frac{\partial p}{\partial X}\right]wm = D'(X)$$

---

# Second-best non-revenue raising environmental policy

$$u_X - C'(X) + (1+\lambda)\left[-{\partial N \over \partial p}\frac{\partial p}{\partial X}\right]wm = D'(X)$$

--

What are each of the terms:

--

$u_X - C'(X)$ is the marginal abatement cost

--

$D'(X)$ is marginal damage

--

$(1+\lambda)\left[-{\partial N \over \partial p}\frac{\partial p}{\partial X}\right]wm$ is new

--

What's the interpretation?

---

# Second-best non-revenue raising environmental policy

$(1+\lambda)\left[-{\partial N \over \partial p}\frac{\partial p}{\partial X}\right]wm$ is called the .hi-blue[marginal interaction effect (MIE)]


--

It tells us how the optimal choice of $X$ departs from $X^*$ because of the labor market distortion
  - Changing $\bar{X}$ changes the price $p$ which changes the household's optimal choice of $N$

--

We need to account for this because the household's choice of leisure will respond to changes in $X$

--

Suppose $N$ and $X$ are substitutes, what does this mean?

---

# Second-best non-revenue raising environmental policy

Substitutes means that $MIE > 0$

--

The marginal social cost of abatement $(MAC + MIE)$ has become .hi-red[larger]

--

Intuition?

--

Its more socially costly to reduce $X$ because the household increases $N$ in response

--

This .hi-red[exacerbates] the distortion caused by the income tax: the household was already undersupplying labor because of the income tax

Now the household undersupplies labor to an even greater extent


---

# Second-best non-revenue raising environmental policy

Complements means that $MIE < 0$

--

The marginal social cost of abatement $(MAC + MIE)$ has become .hi-blue[smaller]

--

Intuition?

--

Its less socially costly to reduce $X$ because the household decreases $N$ in response

--

This .hi-blue[alleviates] the distortion caused by the income tax: the household was undersupplying labor because of the income tax, but now reducing $X$ increases labor supply, shrinking the labor market DWL


---

# Second-best non-revenue raising environmental policy

.pull-left[
![excess burden is because MU of leisure is less than wage](`r knitr::fig_chunk("labor_market_2", "png")`)
]

.pull-right[

$N^c \rightarrow N^*$ when $p^c \rightarrow p^*$ because of a change in $X$

This is $-{\partial N \over \partial p}\frac{\partial p}{\partial X}$

This reduces tax revenue by $e+c$ which is just 
$$(N^* - N^c)(w - (1-m)w) \\
= \underbrace{(N^* - N^c)}_{\approx -{\partial N \over \partial p}\frac{\partial p}{\partial X}}mw$$

]

---

# Second-best non-revenue raising environmental policy

The marginal welfare cost of recovering the lost tax revenue (in order to maintain gov't revenues $G$) by raising $m$ is $\lambda$ giving us a total welfare cost of:
$$\lambda (N^* - N^c)mw$$

--

But $(N^* - N^c)mw$ also happens to be the increase in excess burden: its a .hi-red[direct welfare loss] in addition to the loss from having to increase $m$

So the total welfare loss is:
$$(1+\lambda) (N^* - N^c)mw$$

The discrete version of MIE!

---

# Findings recap

If there's a government revenue constraint, and it can only be met with labor taxes then:

1. The marginal social cost of reducing $X$ is higher if $X$ and $N$ are substitutes and lower if they are complements
--

2. The optimal level of pollution is larger if they are substitutes, lower if they are complements
--

3. The absolute value of the difference in first and second-best pollution levels is larger if:
  <!-- - Labor supply is more elastic -->
  - Demand for $X$ is more inelastic
  - Elasticity of substitution between $N$ and $X$ is greater
  
  
---

# Second-best non-revenue raising environmental policy

We didn't actually show the last part yet

--

First define:
- $\varepsilon_x$ as the own price elasticity ${\frac{\partial X}{\partial p} \frac{p}{X}}$
- $\eta_{XN}$ as the elasticity of substitution between $X$ and $N$: ${\frac{\partial X}{\partial w}  \frac{(1-m)w}{X}}$

--

and take advantage of the .hi-blue[Slutsky symmetry condition] $\partial N / \partial p = \partial X / \partial w$

We can then use these to substitute into the MIE and get:
$$MIE = (1+\lambda)\left[- {\eta_{XN} \over \varepsilon_X}\right]p {m \over 1-m}$$

---

# Second-best non-revenue raising environmental policy

$$MIE = (1+\lambda)\left[- {\eta_{XN} \over \varepsilon_X}\right]p {m \over 1-m}$$
MIE bigger if $|\eta_{XN}|$ is bigger (higher elasticity of substitution)

--

MIE bigger if $|\varepsilon_X|$ is smaller (more inelastic demand for $X$)

<!-- -- -->

<!-- Still need to show labor supply part -->

<!-- --- -->

<!-- # Second-best non-revenue raising environmental policy -->

<!-- Define the elasticity of labor supply .hi[at the after-tax wage] as: -->
<!-- $$\varepsilon_{L} = - {\partial N \over \partial w}{(1-m)w \over L}$$ -->

<!-- -- -->

<!-- And recognize that: -->
<!-- $${\partial N(p,(1-m)w) \over \partial m} = -w {\partial N(p,w) \over \partial w}$$ -->

<!-- -- -->

<!-- Combining these two gives: -->
<!-- $${\partial N \over \partial m}m = \varepsilon_{L} L m / (1-m)$$ -->

<!-- --- -->

<!-- # Second-best non-revenue raising environmental policy -->

<!-- Finally, put: -->
<!-- $${\partial N \over \partial m}m = \varepsilon_{L} L m / (1-m)$$ -->
<!-- in -->
<!-- $$\lambda = \frac{w  m {\partial N \over \partial m}}{w(T-N)-wm{\partial N \over \partial m}}$$ -->

<!-- To get the welfare cost of public funds in terms of fundamental economic parameters -->

<!-- --- -->

<!-- # Second-best non-revenue raising environmental policy -->

<!-- We get: -->
<!-- $$\lambda = \frac{\varepsilon_{L}m/(1-m)}{1- \varepsilon_{L}m/(1-m)}$$ -->

<!-- Side note:  -->
<!-- If labor is more elastic $\varepsilon_{L}$ is larger, the numerator is larger, denominator is smaller $\rightarrow$ MIE is bigger -->

<!-- If labor supply is perfectly inelastic (vertical), there is no welfare cost of public funds! -->

<!-- --- -->

<!-- # Second-best non-revenue raising environmental policy -->

<!-- Finally use: -->
<!-- $$\lambda = \frac{\varepsilon_{L}m/(1-m)}{1- \varepsilon_{L}m/(1-m)}$$ -->
<!-- to get: -->
<!-- $${m\over(1-m)} = {\lambda \over (1 + \lambda)\varepsilon_L}$$ -->

<!-- and then substitute into our MIE expression: -->
<!-- $$MIE = (1+\lambda)\left[- {\eta_{XN} \over \varepsilon_X}\right]p {m \over 1-m}$$ -->

<!-- --- -->

<!-- # Second-best non-revenue raising environmental policy -->

<!-- Finally we will get: -->
<!-- $$MIE = \lambda {-\eta_{XN} \over \varepsilon_X} {p \over \varepsilon_L}$$ -->

<!-- and substitute in our expression for $\lambda$: -->
<!-- $$\lambda = \frac{\varepsilon_{L}m/(1-m)}{1- \varepsilon_{L}m/(1-m)}$$ -->

<!-- To get: -->
<!-- $$MIE = p\frac{m/(1-m)}{1- \varepsilon_{L}m/(1-m)} {-\eta_{XN} \over \varepsilon_X}$$ -->

<!-- --- -->

<!-- # Second-best non-revenue raising environmental policy -->

<!-- $$MIE = p\frac{m/(1-m)}{1- \varepsilon_{L}m/(1-m)} {-\eta_{XN} \over \varepsilon_X}$$ -->

<!-- MIE is bigger and the absolute value of the difference in first and second-best pollution levels is larger if: -->

<!-- <!-- 1. Labor supply is more elastic: $\varepsilon_{L}$ bigger --> -->
<!-- 2. Demand for $X$ is more inelastic: $\varepsilon_X$ smaller -->
<!-- 3. Elasticity of substitution between $N$ and $X$ is greater: $\eta_{XN}$ bigger in magnitude -->

---

# Revenue raising environmental policy

Now suppose that the government raises revenues via emission taxation or auctioning permits

--

In our model the government has a revenue requirement:
$$G=wm(T-N) + \tau X$$
where $\tau$ is the revenue per unit of the dirty good

--

The regulator's problem is thus to select two tax rates: $m$ and $\tau$

--

For simplicity we still assume all tax revenues are returned lump sum to households

---

# Revenue raising environmental policy

First derive household spending on the numeraire good:
$$Z = (1-m)w(T-N)-pX+G = w(T-N)-pX+\tau X$$
where the second equality comes from substituting out the govt's budget constraint: $G = wm(T-N) + \tau X$

--

The endogenous variables to be determined are: $X$, $N$ and $p$, quantity of the dirty good, leisure, and the price of the dirty good

--

These are a function of the govt's choice of $m$ and $\tau$

---

# Revenue raising environmental policy

The household FOCs are:
$$u_X = p \qquad u_N = (1-m)w$$
and the firm FOC is:
$$C'(X) = p-\tau$$

--

MU = MC of consumption and leisure

MR = MC of production

--

Next, as usual, differentiate the FOCs wrt $\tau$


---

# Revenue raising environmental policy

This gives us 3 equations and 3 unknown partial derivatives:
$$u_{XX} {\partial X \over \partial \tau} + u_{XN}{\partial N \over \partial \tau} = {\partial p \over \partial \tau} \tag{Household X FOC}$$


$$u_{XN} {\partial X \over \partial \tau} + u_{NN}{\partial N \over \partial \tau} = 0 \tag{N FOC}$$

$$C''(X) {\partial X \over \partial \tau} = {\partial p \over \partial \tau} - 1 \tag{Firm X FOC}$$

Substitute and solve...

---

# Revenue raising environmental policy

Now solve for how the endogenous variables change in $\tau$

--

$${\partial X \over \partial \tau} = {u_{NN} \over H} < 0$$

$${\partial N \over \partial \tau} = {-u_{XN} \over H} \lessgtr 0$$

$${\partial p \over \partial \tau} = {u_{XX}u_{NN}-u_{XN}^2 \over H} > 0$$

where $H = u_{XX}u_{NN}-u_{XN}^2-C''(X)u_{NN} > 0$

---

# Revenue raising environmental policy

Now that we know how the endogenous variables change we can solve for the regulator's optimal taxes

--

The regulator wants to maximize social welfare given the budget constraint:
\begin{gather}
\max_{m,\tau} \underbrace{U(X,N) + Z - D(X)}_{\text{household utility}} + \underbrace{pX - C(X) - \tau X}_{\text{firm profit}} \\
\text{ subject to:      } wm(T-N)+\tau X=G
\end{gather}

--

Substitute in for Z from household spending:
$$Z = w(T-N)-pX+\tau X$$

And look at the $\tau$ FOC

---

# Revenue raising environmental policy

$$\left[u_X - C'(X) - D'(X)\right]{\partial X \over \partial \tau} + \left[\underbrace{u_N - w}_{-wm} - \lambda wm\right]{\partial N \over \partial \tau} + \lambda\left[X + \tau {\partial X \over \partial \tau} \right]  = 0$$

--

Just follow the same steps as we did with the non-revenue raising case and divide by ${\partial X \over \partial \tau}$ to get:
$$\underbrace{u_x - C'(X)}_{MAC} + \underbrace{(1+\lambda)wm\left[-{\partial N \over \partial \tau}\bigg/{\partial X \over \partial \tau}\right]}_{MIE} + \underbrace{\lambda\left[\tau + X/{\partial X \over \partial \tau}\right]}_{MRE} = D'(X)$$

--

Since the tax is per unit, we have that: ${\partial N \over \partial \tau}\big/{\partial X \over \partial \tau} = {\partial N \over \partial p}\big/{\partial X \over \partial p}$, MIE is similar in revenue and non-revenue raising contexts

---

# Revenue raising environmental policy

What is this new term, $MRE$: $\lambda\left[\tau + X/{\partial X \over \partial \tau}\right]$?

--

It's the .hi-blue[marginal revenue effect]: the amount by which emission tax revenue changes when $X$ changes, scaled by $\lambda$, the MC of public funds

--

MRE changes the marginal social cost of $X$ because changes in $\tau$ affect how much revenue we need to raise with distorting labor taxation

--

Let's get some intuition at the corner case of $\tau = 0$

What's the sign of $MRE$?

---

# Revenue raising environmental policy

$MRE(\tau=0)$: $\quad \lambda\left[x/{\partial X \over \partial \tau}\right]$

--

Here $MRE < 0$ because ${\partial X \over \partial \tau} < 0$, what's the intuition?

--

$\rightarrow$ the additional revenue from an increase in $\tau$ lets us reduce labor taxes

--

$\rightarrow$ this reduces the distortionary tax in the labor market

--

$\rightarrow$ this reduces welfare losses in the labor market

--

$\rightarrow$ this reduction in welfare losses reduces the marginal social cost of reducing $X$, decreasing the optimal level of X

---

# Revenue raising environmental policy

Is MRE always negative?

--

No

--

We can get some intuition by making a substitution:

--

$$MRE \equiv \lambda\left[\tau + X\Big/{\partial X \over \partial \tau}\right] = \lambda\left[\tau + X\Big/{\partial X \over \partial p}\right] = \lambda\left[\tau + p/\varepsilon_{X}\right] = \lambda\tau\left[1 + 1/\varepsilon^\tau_{X}\right]$$
where $\varepsilon_{X} < 0$ is the elasticity of demand for the dirty good and $\varepsilon^\tau_{X}$ is the elasticity with respect to the tax

---

# Revenue raising environmental policy

$$MRE \equiv \lambda\left[\tau + p/\varepsilon_{X}\right]$$

MRE is negative and increases total abatement if:

--

- demand for dirty good is sufficiently inelastic $(\varepsilon_{X}$ small)

--

- the price of the dirty good is sufficiently larger than the emission tax

--

Why?


---

# Revenue raising environmental policy


.pull-left[
```{r negative_mre, echo = FALSE, fig.show = 'hide', warning = F}

# demand <- function(x) 5 - x
mac_mre <- function(x) 9 - 3*x
mac <- function(x) 7 - 2*x
mre <- function(x) 2-x
intersections_1 <- tibble(x = c(5, 5, 0, 0, 5, 2.5, 2.5),
                        y = c(0, 2, 2, 1, 1, 0, 2))
ggplot() +
  annotate("segment", x = 0, xend = 5,
           y = 1, yend = 1,
           linetype = "solid", size = 1.5, color = "#ca5670") +
  annotate("segment", x = 0, xend = 5,
           y = 0, yend = 0,
           linetype = "dotted", size = 1, color = "black") +
  annotate("segment", x = 8/3, xend = 8/3,
           y = -2, yend = 1,
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = 7/2, xend = 7/2,
           y = -2, yend = 0,
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = 3, xend = 3,
           y = -2, yend = 1,
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = mac_mre, color = "darkslateblue", size = 1.5) +
  stat_function(fun = mac, color = "#638ccc", size = 1.5) +
  stat_function(fun = mre, color = "violetred", size = 1.5) +
  annotate("text", size = 8, label = expression(u[X] - "C'(X) + MRE"), x = 2.25, y = 4.5) +
  annotate("text", size = 8, label = expression(u[X] - "C'(X)"), x = 0.75, y = 4.5) +
  annotate("text", size = 8, label = "MRE", x = 0.5, y = 2) +
  annotate("text", size = 8, label = "D'(X)", x = 3.5, y = 1.25) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#ffffff", colour = NA),
    plot.background = element_rect(fill = "#ffffff", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "X",
       y = "$",
       title = "Inelastic demand means MRE < 0") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(7/2, intersections_1$x[4], 8/3, 3), labels = expression(X^C, 0, X^{RE}, X^"*")) +
  scale_y_continuous(expand = c(0, 0), limits = c(-2, 5),
                     breaks = c(intersections_1$y[4], 0), labels = expression("d", 0)) +
  coord_cartesian(xlim = c(0, 3.7))

```


![MRE can increase optimal tax](`r knitr::fig_chunk("negative_mre", "png")`)
]

.pull-right[


.hi-blue[Demand for dirty good is sufficiently inelastic]:

Suppose ${\partial N \over \partial p}= 0$ so $MIE = 0$, $C'(X) = c$, $D'(X) = d$

Inelastic demand lets us raise more revenue from a small change in the tax

]


---

# Revenue raising environmental policy


.pull-left[

![MRE can increase optimal tax](`r knitr::fig_chunk("negative_mre", "png")`)
]

.pull-right[

Inelastic demand lets us raise more revenue from a small change in the tax

This reduces the marginal social cost of reducing X

Optimal X with revenue-raising is lower than without: $X^{RE} < X^*$

]

---

# Revenue raising environmental policy


.pull-left[

![MRE can increase optimal tax](`r knitr::fig_chunk("negative_mre", "png")`)
]

.pull-right[

We can also see that if $D'(X)$ was very large, making $\tau$ larger, we would be where $MRE > 0$

]


---

# Double dividend

Is there a prospect for a .hi[double dividend?]

There is a .hi-blue[weak double dividend] if welfare is always greater when revenue raised via environmental taxation is used to reduced distortionary taxation rather than refunded lump sum
- This is always true

--

There is a .hi-red[strong double dividend] if the emission tax should always be set above the $MAC = MD$ level, resulting in greater pollution reductions and more revenue raised
- This may or may not be true

---

# Double dividend

When is there a strong double dividend?

--

Recall: $MSC = MAC + MIE + MRE$

--

To have a strong double dividend we need: 
--
$MSC < MAC \Rightarrow MIE + MRE < 0$

--

This can happen via two pathways:

**Pathway 1:** $MIE, MRE < 0$ or, $MIE < 0$ and $|MIE| > MRE > 0$

--

In this pathway we have that leisure and the polluting good are .hi-blue[complements]

--

Price of $X$ rises from $\tau$, demand for leisure goes down, labor goes up

---

# Double dividend

Is this likely to be true?

--

Not really: leisure and consumption are generally substitutes

--

**Pathway 2:** $MIE > 0 > MRE$, $|MRE| > MIE$

--

Here leisure and consumption are substitutes, but the revenue effect dominates the interaction effect

--

Let's look at this pathway in more detail

---

# Double dividend

Again, assume $C'(X) = c$, this gives us that:
\begin{align}
  MIE = \lambda\left(- {\eta_{XN} \over \varepsilon_X} \right){p \over \varepsilon_L} \qquad MRE = \lambda\left({p \over \varepsilon_X} + \tau \right)
\end{align}

where

$$\eta_{XN} = \overbrace{\frac{\partial X}{\partial w}\frac{(1-m)w}{X}}^{\text{cross-price elasticity}} \qquad \varepsilon_L = \overbrace{-\frac{\partial N}{\partial w} \frac{(1-m)w}{L}}^{\text{labor supply elasticity}} = \frac{\partial L}{\partial w} \frac{(1-m)w}{L}$$

---

# Double dividend

Suppose N and X are *average substitutes* which means the negative cross-price elasticity is equal to the the labor supply elasticity $\eta_{XN} = \varepsilon_L$

This is true if a 1% wage increase gives a $\eta_{XN}$% $= \varepsilon_L$% spending increase

--

Then:
$$MIE = \lambda\left(-{p\over \varepsilon_X} \right) < \lambda\left({p \over \varepsilon_X} + \tau \right) = MRE$$

$\Rightarrow$ we shouldn't expect a strong double dividend because MIE + MRE = $\lambda \tau > 0$



---

# Revenue raising environmental policy


.pull-left[
```{r mac_mie_mre, echo = FALSE, fig.show = 'hide', warning = F}

# demand <- function(x) 5 - x
mac_mre <- function(x) 8 - 2*x
mac <- function(x) 4 - x
mre <- function(x) 2 - x/2
intersections_1 <- tibble(x = c(5, 5, 0, 0, 5, 2.5, 2.5),
                        y = c(0, 2, 2, 1, 1, 0, 2))
ggplot() +
  annotate("segment", x = 0, xend = 5,
           y = 1, yend = 1,
           linetype = "solid", size = 1.5, color = "#ca5670") +
  annotate("segment", x = 2, xend = 2,
           y = 0, yend = 1,
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = 3, xend = 3,
           y = 0, yend = 1,
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = 7/2, xend = 7/2,
           y = 0, yend = 1,
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = mac_mre, color = "darkslateblue", size = 1.5) +
  stat_function(fun = mac, color = "#638ccc", size = 1.5) +
  stat_function(fun = mre, color = "violetred", size = 1.5) +
  annotate("text", size = 8, label = "MAC+MIE", x = 2.5, y = 4.5) +
  annotate("text", size = 8, label = "MAC+MIE+MRE", x = 1, y = 4.0) +
  annotate("text", size = 8, label = "MAC", x = 0.5, y = 2) +
  annotate("text", size = 8, label = "D'(X)", x = 4, y = 1.25) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#ffffff", colour = NA),
    plot.background = element_rect(fill = "#ffffff", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "X",
       y = "$",
       title = "Inelastic demand means MRE < 0") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(4, 7/2, 3, 2), labels = expression(X^C, X^{IE}, X^{GE}, X^"*")) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[4], 0), labels = expression("d", 0)) 

```


![MRE can increase optimal tax](`r knitr::fig_chunk("mac_mie_mre", "png")`)
]

.pull-right[


Even though there isn't a double dividend, MIE and MRE .hi-red[still matter] for the optimal second-best pollution level

Optimal pollution $X^{GE}$ is larger than first-best $X^*$, but less than the level without revenue recycling $X^{IE}$ 

]


---

# Policy instruments with labor market distortions

How do environmental policy instruments work when we have the distortionary labor tax?

--

Taxes and auctioned permits are easy, just set the tax equal to:
$$\tau = D'(X) + MIE + MRE$$
or the number of permits equal to $X^{GE}$ to obtain the optimal second-best outcome

--

The regulator obtains revenues $\tau X^{GE} = \sigma x^{GE}$ and recycles it to reduce labor taxation

--

What about freely allocated permits or command and control?

---

# Policy instruments with labor market distortions

This would lead to the same *environmental* outcome, but not achieve the the welfare maximizing outcome

Why?

--

Free allocation and command and control do not generate revenues that let us reduce labor taxation

--

Setting $X^{GE} < X^c$ raises the price of $X$, increases leisure, and reduces revenues via the interaction effect

--

Without revenue from permits or taxes, the optimal pollution level is higher
