---
title: "Lecture 6"
subtitle: "Non-competitive output markets"
author: Ivan Rudik
date: AEM 6510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'


---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, tidylog, xaringanExtra, rlang, patchwork
)
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

---

# Roadmap

Previously we assumed output markets were competitive and found that the conditions for efficiency (MAC = MD) still hold

--

But markets are often  .hi-red[not] competitive

--

Will this affect our results?

--

Why might it?

--

Now we have .hi-blue[two] distortions in our market, pollution and market power

---

# Environmental policy under monopoly

Suppose we have an industry with a .hi-red[single] firm that generates a local pollutant

--

We will again look at both the general and specific cases of our model

--

Lets begin with the specific case where abatement is only possible through output reductions and $E=\delta X$

---

# Environmental policy under monopoly

The monopolist's profit-maximization problem under an emission tax is:
$$\max_{X} \Pi(X) = P(X)X - C(X) - \tau E = P(X)X - C(X) - \tau \delta X$$

--

Where the monopolist now controls $X$: aggregate production

--

The FOC for the problem is:
$$P'(X)X + P(X) = C'(X) + \tau \delta$$
and the profit-maximizing output choice is given by $X^M(\tau)$

---

# Environmental policy under monopoly

The FOC for the problem is:
$$P'(X)X + P(X) = C'(X) + \tau \delta$$

This illustrates the monopolists decision rule: MR = MC

--

MR consists of two pieces:

1. $P(X)$: Additional revenue from increased $X$
2. $P'(X)X$: decreases in the price for .hi-blue[inframarginal units] from increased $X$

--

The monopolist now accounts for how increasing production lowers price

---

# Environmental policy under monopoly

How does the firm respond to the tax? Differentiate the FOC wrt $\tau$:
$$\left[P''(X^M)X^M+ 2P'(X^M) - C''(X^M)\right]{d X^M \over d \tau} = \delta$$

--

and rearrange:
$${d X^M \over d \tau} = {\delta \over \left[P''(X^M)X^M+ 2P'(X^M) - C''(X^M)\right]}$$

--

What's the sign on this expression? 

--

Let's make two assumptions

---

# Environmental policy under monopoly

$${d X^M \over d \tau} = {\delta \over \left[P''(X^M)X^M+ 2P'(X^M) - C''(X^M)\right]}$$

--

Assume the inverse demand function $P(X)$ is decreasing, 
--
and for all $X>0$:
$${P''(X) \over P'(X)}X > -1$$

--

This just comes from the second-order sufficient condition for a maximum being satisfied, it mean inverse demand isn't *too convex*: $P''$ is bounded above

--

It also ensures that ${d X^M \over d \tau} < 0$

---

# Environmental policy under monopoly

Now lets look at the regulators problem of maximizing the benefits minus the costs of production, accounting for damages:
$$\max_{X} \int_{0}^XP(t)dt - C(X) - D(\delta X)$$

--

She is maximizing the consumption value minus production and environmental costs

--

The FOC for this problem is:
$$P(X^*) = C'(X^*) + \delta D'(\delta X^*)$$
which doesn't map as nicely to the firm's FOC as in competitive markets

---

# Environmental policy under monopoly

Lets re-write the regulator's problem and explicitly include the firm's optimal response to $\tau$:
$$W(\tau) =  \int_{0}^{X^m(\tau)} P(t)dt - C(X^m(\tau)) - D(\delta X^m(\tau))$$

--

Then differentiate with respect to $\tau$ to get:
$$[P(X^m) - C'(X^m) - \delta D'(\delta X^m)] \times {d X^m \over d\tau} = 0$$

---

# Environmental policy under monopoly

$$[P(X^m) - C'(X^m) - \delta D'(\delta X^m)] \times {d X^m \over d\tau} = 0$$

Recognize $P(X^m) - C'(X^m) = \tau\delta - P'(X^m)X^M$ from the firm FOC

--

Rearranging gives us that the optimal tax rate is characterized by:
$$\tau = D'(\delta X^m) + {P'(X^m)X^m \over \delta}$$

--

Can we achieve the first-best with this tax?

--

Yes! If we plug it into the firm FOC it is the same as the regulator's welfare-maximizing FOC

---

# Environmental policy under monopoly

$$\tau = D'(\delta X^m) + {P'(X^m)X^m \over \delta}$$

What is the intuition behind this expression?

--

First, since $P'(X^m) < 0$, $\tau < D'(\delta X^m)$

--

Why?

--

The monopolist already reduces output to exercise its market power

--

So we don't need as big of a tax, or may even need a .hi-blue[subsidy] if the monopolist was reducing output too much, to achieve the first-best


---

# Graphical intuition for the optimal tax


.pull-left[
```{r optimal_monopoly_tax, echo = FALSE, fig.show = 'hide', warning = F}
smc <- function(x) 2.5
md <- function(x) 1.5
pmc <- function(x) 1
demand <- function(x) 5 - x
marginal_revenue <- function(x) 5 - 2*x
intersections_1 <- tibble(x = c(0, 2, 2, 2.5, 2, 2.5/2),
                        y = c(2.5, 3, 0, 2.5, 2, 5-2.5/2))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
ggplot() +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[5],
           y = intersections_1$y[2], yend = intersections_1$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = smc, color = "#ca5670", size = 1.5) +
  stat_function(fun = pmc, color = "#ca5670", size = 1.5) +
  stat_function(fun = md, color = "#ca5670", size = 1.5) +
  stat_function(fun = demand, color = "darkslateblue", size = 1.5) +
  stat_function(fun = marginal_revenue, color = "darkslateblue", size = 1.5) +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[4],
           y = intersections_1$y[1], yend = intersections_1$y[4],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[4],
           y = intersections_1$y[1], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[6],
           y = intersections_1$y[6], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[6], xend = intersections_1$x[6],
           y = intersections_1$y[3], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  annotate("text", size = 8, label = expression(D*minute + C[X]), x = 4.5, y = 2.7) +
  annotate("text", size = 8, label = expression(D*minute), x = 4.5, y = 1.7) +
  annotate("text", size = 8, label = expression(C[X]), x = 4.5, y = 1.2) +
  annotate("text", size = 8, label = expression(P(X)), x = 4.4, y = 0.15) +
  annotate("text", size = 8, label = expression(P*minute(X)), x = 3, y = 0.15) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Quantity",
       y = "MC/MD ($)",
       title = "Optimal tax is less than MD") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[6]), labels = expression(X^m, X^"*", X^"MD")) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[2], intersections_1$y[4], intersections_1$y[6]), labels = expression(P^m, P^"*", P^"MD"))

```

![optimal emission tax for a monopoly is less than marginal damage](`r knitr::fig_chunk("optimal_monopoly_tax", "png")`)
]

.pull-right[
Suppose MD and MC are constant for simplicity and we are in the specific case
 
$X^m, P^m$ is the unregulated monopoly allocation

$X^{MD}, P^{MD}$ is the outcome if we set $\tau = D'(E^*)$

Since $X^{MD} < X^m < X^*$ this clearly .hi-red[reduced welfare]
]


---

# Graphical intuition for the optimal tax


.pull-left[
```{r optimal_monopoly_tax2, echo = FALSE, fig.show = 'hide', warning = F}
smc <- function(x) 2.5
md <- function(x) 1.5
pmc <- function(x) 1
demand <- function(x) 5 - x
marginal_revenue <- function(x) 5 - 2*x
intersections_1 <- tibble(x = c(0, 2, 2, 2.5, 2, 2.5/2),
                        y = c(2.5, 3, 0, 2.5, 2, 5-2.5/2))
polys_dwl_m <- tibble(x = c(2.5/2, 2.5/2, 2.5),
                      y = c(5-2.5/2, 2.5, 2.5))
polys_dwl_md <- tibble(x = c(2, 2, 2.5),
                       y = c(3, 2.5, 2.5))
  
ggplot() +
  geom_polygon(data = polys_dwl_m, aes(x = x, y = y), fill = "red", alpha = 0.2) +
  geom_polygon(data = polys_dwl_md, aes(x = x, y = y), fill = "blue", alpha = 0.2) +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[5],
           y = intersections_1$y[2], yend = intersections_1$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = smc, color = "#ca5670", size = 1.5) +
  stat_function(fun = pmc, color = "#ca5670", size = 1.5) +
  stat_function(fun = md, color = "#ca5670", size = 1.5) +
  stat_function(fun = demand, color = "darkslateblue", size = 1.5) +
  stat_function(fun = marginal_revenue, color = "darkslateblue", size = 1.5) +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[4],
           y = intersections_1$y[1], yend = intersections_1$y[4],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[4],
           y = intersections_1$y[1], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[6],
           y = intersections_1$y[6], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[6], xend = intersections_1$x[6],
           y = intersections_1$y[3], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  annotate("text", size = 8, label = expression(D*minute + C[X]), x = 4.5, y = 2.7) +
  annotate("text", size = 8, label = expression(D*minute), x = 4.5, y = 1.7) +
  annotate("text", size = 8, label = expression(C[X]), x = 4.5, y = 1.2) +
  annotate("text", size = 8, label = expression(P(X)), x = 4.4, y = 0.15) +
  annotate("text", size = 8, label = expression(P*minute(X)), x = 3, y = 0.15) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Quantity",
       y = "MC/MD ($)",
       title = "Optimal tax is less than MD") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[6]), labels = expression(X^m, X^"*", X^"MD")) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[2], intersections_1$y[4], intersections_1$y[6]), labels = expression(P^m, P^"*", P^"MD"))

```

![optimal emission tax for a monopoly is less than marginal damage](`r knitr::fig_chunk("optimal_monopoly_tax2", "png")`)
]

.pull-right[
The purple/blue area is the original DWL under monopoly

The red + purple/blue area is the DWL if we tax at marginal damage, this .hi-red[worsened welfare]

What's the optimal tax that gets us $E^*$?
]

---

# Graphical intuition for the optimal tax


.pull-left[

![optimal emission tax for a monopoly is less than marginal damage](`r knitr::fig_chunk("optimal_monopoly_tax2", "png")`)
]

.pull-right[
We want to shift the marginal cost of the firm so that it intersects MR at $X^*$, this vertical distance is the optimal tax

MR at $X^*$ is 0, so we want to shift marginal cost down to zero

Our tax is then: $0-C_X$: we actually .hi-blue[subsidize output]
]

---

# Environmental policy under monopoly

In the specific case we could achieve the first-best with just an emission tax and solve two externalities at once

--

Does this hold more generally?

--

The general monopoly problem is:
$$\max_{X,E} \Pi(X,E) = P(X)X - C(X,E) - \tau E$$

---

# Environmental policy under monopoly

The FOCs are
$$P'(X)X + P(X) = C_X(X,E) \qquad -C_E(X,E) = \tau$$

The first is just MR = MC of production

The second is the MAC = tax for emissions

--

Recall the regulators solution will look like:
$$P(X^*) = C_X(X^*,E^*) \qquad -C_E(X^*,E^*) = D'(E^*)$$

--

Can the regulator use an emission tax alone to achieve the efficient outcome?

---

# Environmental policy under monopoly

The regulator's problem is:
$$W(\tau) = \int_0^{X^m(\tau)}P(t)dt - C(X^m(\tau),E^m(\tau)) - D(E^m(\tau))$$

--

The FOC of this problem is:
$$[P(X^M) - C_X(X^m,E^m)]{dX^m \over d\tau} - [C_E(X^m,E^m) + D'(E^m)]{dE^m \over d\tau} = 0$$

--

To get the .hi-blue[second-best] tax rate, substitute in the conditions from the firm FOCs:
$$P'(X)X + P(X) = C_X(X,E) \qquad -C_E(X,E) = \tau$$

---

# Environmental policy under monopoly

This gives us that:
$$\tau = D'(E^m) + P'(X^m)X^m {dX^m/d\tau \over dE^m/d\tau}$$
--

What is the second term?

--

It represents how a change in emissions caused by the tax $\tau$ changes how the monopolist exercises market power

--

$P'(X^m)X^m$ is the market power term in MR, ${dX^m/d\tau \over dE^m/d\tau}$ is the output response to a tax-induced change in emissions

---

# Environmental policy under monopoly

The sign of this term depends on ${dX^m/d\tau \over dE^m/d\tau}$, so lets sign the two components

--

We do this by differentiating the firm FOCs wrt $\tau$ and first solving for $dX^m/d\tau$:
$$P'(X)X + P(X) = C_X(X,E) \qquad -C_E(X,E) = \tau$$

--

to give us that (after some algebra and dropping function arguments):
$${dX^m \over d\tau} = { -C_{XE} \over C_{EE}\left[ P''X + 2P' - C_{EE}\left\{ C_{XX}C_{EE}-C^2_{XE} \right\} \right] } < 0$$

---

# Environmental policy under monopoly

$${dX^m \over d\tau} = { -C_{XE} \over C_{EE}\left[ P''X + 2P' - C_{EE}\left\{ C_{XX}C_{EE}-C^2_{XE} \right\} \right] } < 0$$


We assumed $-C_{XE}> 0$, strict convexity ensures $C_{EE}, C_{XX}C_{EE}-C^2_{XE} > 0$, and our most recent assumption that demand is not too convex ensures $P''X + 2P' < 0$

--

The numerator is positive, the denominator is negative: output declines as the tax increases
---

# Environmental policy under monopoly

For emissions we differentiate the firm FOCs wrt $\tau$ solve for ${dE^m \over d\tau}$:
$${dE^m \over d\tau} = {-1 \over C_{EE}} + { C^2_{XE} \over C^2_{EE}\left[ P''X + 2P' - C_{EE}\left\{ C_{XX}C_{EE}-C^2_{XE} \right\} \right] } < 0$$

--

As before, we assumed:
- $-C_{XE}> 0$
- Strict convexity ensures $C_{EE}, C_{XX}C_{EE}-C^2_{XE} > 0$
- Demand is not too convex assumption ensures $P''X + 2P' < 0$

So both output and emissions decline in the tax rate

---

# Environmental policy under monopoly

Can we use the tax alone to achieve the first-best?

--

Substitute in the optimal tax expression $\tau$ into the firm emission FOC and see if it reduces to the conditions for the first-best outcome:

--

$$-C_E(X,E) = D'(E^m) + P'(X^m)X^m {dX^m/d\tau \over dE^m/d\tau}$$

--

The last term doesn't equal zero so this FOC, given the second-best tax, cannot is not equal to the first-best condition

--

.hi-red[The regulator cannot achieve the first-best in the general case with a tax alone]

---

# Environmental policy under monopoly

What ended up being the difference between the specific and general case?

--

In the specific case the monopolist only had one degree of freedom, output, and the regulator had one instrument, the tax

--

With one instrument and one degree of freedom the regulator can incentivize the firm to select a specific value of output AND emissions

---

# Environmental policy under monopoly

In the general case the monopolist has two degrees of freedom: it can choose output and emissions separately

--

The regulator still has only one instrument

--

You cannot use one instrument to pin down two values (similar to solving one equation for two unknowns)

--

When you have fewer instruments than market distortions you are in a .hi-blue[second-best world]

---

# Environmental policy under monopoly

Here's some intuition:

--

Suppose $|P'(X)|$ is large so demand is very inelastic $\rightarrow$ small changes in quantity lead to big changes in price

--

In this case, the market power distortion is a big problem $(P(X^M) >>> C_X(X^M,E^M))$

--

If we tax equal to marginal damage, we will make the market power distortion worse

--

The second term in the tax expression reduces the tax to account for these concerns

---

# Environmental policy under monopoly

We can achieve the first-best if we have two instruments to address two distortions

- Use an output subsidy to incentivize the monopolist to produce the efficient level
- Use an emission tax to get the efficient level of emissions

The monopolist's problem is:
$$\Pi(X,E) = [P(X) + \xi]X - C(X,E) - \tau E$$
---

# Environmental policy under monopoly

The FOCs are:
$$P(X) + \xi + P'(X)X = C_X(X,E) \qquad -C_E(X,E) = \tau$$

If we set:
$$\xi = -P'(X^*)X^* \qquad \tau = D'(E^*)$$
the firm FOCs reduce to the regulator's efficiency conditions

--

This is a special case of the *Tinbergen Rule* that says you need as many instruments as distortions to achieve the first-best
