---
title: "Lecture 5"
subtitle: "Competitive output markets"
author: Ivan Rudik
date: AEM 6510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'


---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, tidylog, xaringanExtra, rlang, patchwork
)
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

---

# Roadmap

So far we have ignored output markets in our analysis

--

But firms actually have production costs in addition to abatement costs, and sometimes these costs cannot be disentangled

--

Now we explore models where output and abatement may not be separable

This captures a wider range of potential abatement methods and technologies

---

# The competitive output model

This model is simply an extension of our previous one

For now drop firm-specific subscripts but we assume firms are asymmetric

--

A firm's production technology is given by: 
$$x = f(l_1,\dots,l_K)$$
where $x$ is how many units of output are produced using production function $f$ and vector of inputs $\{l_1,\dots,l_K\}$

---

# The competitive output model


A firm's emission technology is given by: 
$$e = g(l_1,\dots,l_K)$$
where $e$ is how many units of emissions resulting from using the vector of inputs $\{l_1,\dots,l_K\}$

--

Note that the marginal product of $l_k$ for either $f$ or $g$ could be positive, negative, or zero

--

Why?

--

An input that reduces emissions could reduce output, output-enhancing inputs could increase emissions or be emission-neutral

---

# The competitive output model

Next we need the cost functions

--

The cost function $C(x,e)$ is derived from the firm's cost minimization problem

--

We want to minimize the costs of producing a given combination of output $(x,e)$
$$C(x,e) = \min_{l_1,\dots,l_k} \left\{ \sum_{k=1}^K w_k l_k + \lambda \underbrace{\left[ x - f(l_1,\dots,l_K) \right]}_{\text{x units of output}} + \mu \underbrace{\left[e-g(l_1,\dots,l_K)\right]}_{\text{e units of emissions}} \right\}$$
where $\{w_1,\dots,w_K\}$ is a vector of input prices


---

# The competitive output model

Let $\hat{e}^x$ be the freely chosen level of emissions at output level $x$

We need two sets of assumptions to continue

--

## Assumption Set 1 .hi-red[(general case)]

--

1. $C(x,e)$ is twice continuously-differentiable with $C_x > 0$ and for any $x$ there is an emission level $\hat{e}^x$ such that $C_e(x,\hat{e}^x) = 0$
--

2. $C_e(x,\hat{e}^x) < 0 \,\,\, \forall e < \hat{e}^x$ and $C_e(x, \hat{e}^x) \geq 0 \,\,\, \forall e \geq \hat{e}^x$
--

3. $C_{xe} = C_{ex} < 0 \,\,\, \forall e < \hat{e}^x$ 
--

4. $C_{xx} > 0, C_{ee} > 0, C_{xx}C_{ee} - C_{ex}^2 > 0$ 

---

# The competitive output model

What do these tell us?

1. MC of production is positive and there is a cost-minimizing emission level $\hat{e}^x$ for all $x$ in the absence of environmental regulation
--

2. MAC is increasing in abating below $\hat{e}^x$
--

3. MC of production is smaller with higher emissions $\leftrightarrow$ MAC shifts up if output rises
--

4. Production and abatement costs are convex (marginal costs are increasing)

---

# The competitive output model

We can also see that the non-regulated emission level $\hat{e}^x$ rises with $x$ from the $C_e(x,\hat{e}^x) = 0$ assumption

--

Differentiate $C_e(x, \hat{e}^x) = 0$ wrt $x$ to get: $C_{ex}(x,\hat{e}^x){dx \over dx} + C_{ee}(x,\hat{e}^x){d\hat{e}^x \over dx} = 0$

--

Rearrange to get: ${d\hat{e}^x \over dx} = - {C_{ex} \over C_{ee}} > 0$

---

# The competitive output model

## Assumption Set 2 .hi[(specific case)]

We will be directly linking emissions to output in the .hi[specific case] so we make these assumptions:

- Cost function $C(x)$ is twice continuously-differentiable
--

- $C'(x) > 0, C''(x) > 0$
--

- Emissions are given by $e = \delta(x)$ where $\delta'(x) > 0$
--

- In some cases we will assume $\delta(x) = \delta \cdot x$ to simplify

---

# The competitive output model

What do these tell us?

The MAC function is tired directly to the firm's marginal profit

--

If $p$ is the output price of $x$, profit is:
$$\Pi = px - c(x)$$
and if $e = \delta x$ we have that:
$${d \Pi \over de} = {p - C'(e/\delta) \over \delta}$$

---

# The competitive output model

At an unregulated optimum it must be that $p - C'(x) = 0$ so $p - C'(\hat{e}^x/\delta) = 0$ defines the unregulated level of emissions

--

For $e < \hat{e}^x$: $p - C'(e/\delta) > 0$ since $\hat{e}^x$ is privately optimal for the firm

--

Thus ${d \Pi \over de}$ is the marginal abatement cost where ${d \Pi \over de} > 0$ for $e < \hat{e}^x$

--

.hi-blue[The MC of abatement is the forgone marginal profits from reducing emissions]

--

We can also see that the MAC is increasing:
$${d^2 \Pi \over de^2} = {C''(e/\delta) \over \delta^2} \geq 0$$

---

# The competitive output model

Next we need to model the demand side of the market

--

Let consumer utility be:
$$U^i = U_i(x_i) + z_i - D_i(E)$$
where

- $x_i$ is the person's consumption level
- $z_i$ is spending on all other non- $x$ goods
- $D_i(E)$ are damages from aggregate emissions $E$
- There are $i=1,\dots,J$ consumers

---

# The competitive output model

The consumer has a budget constraint:
$$y = px_i + z_i$$
where the price of $z_i$ is normalized to 1, $p$ is the price of $x$ in terms of $z$

--

Utility maximization gives us that $u'_i(x_i) = p$

--

This defines the inverse demand for $x$ as: $p_i(x_i) = u'_i(x_i)$

--

We can then derive gross benefits from consumption: $\int_0^{x_i} u'_i(s)ds =  u_i(x_i)$

---

# The competitive output model

Next we want to derive .hi-blue[aggregate] market benefits

--

Let $X = \sum_{i=1}^I x_i$ be aggregate consumption,  $P(X)$ be the market inverse demand curve, and $D(E)$ be the aggregate damage curve
- You get $P(X)$ by just horizontally summing $p_i(x_i)$ like we did in previous classes

--

$P(X)$ and $D(E)$ allow us to .hi-blue[fully] characterize benefits and damages to households
  
---

# The competitive output model

Now we have both sides of our model, next we need to define .hi-blue[social welfare] so we can find efficient outcomes

--

Social welfare in the .hi-red[general case] is given by:
$$W(x_1,\dots,x_J,e_1,\dots,e_J) \equiv \int_0^{X\equiv\sum x_j} P(s)ds - \sum_{j=1}^J C^j(x_j, e_j) - D(E)$$
where $j$ are specific firms, and household costs and firm revenues cancel out because they are just a transfer from households to firms

Welfare is CS minus total cost, minus damages

---

# The competitive output model

Social welfare in the .hi[specific case] when $e_j = \delta_j(x_j)$ is given by:
$$W(x_1,\dots,x_J) \equiv \int_0^{X\equiv\sum x_j} P(s)ds - \sum_{j=1}^J C^j(x_j) - D(E)$$

where $E = \sum_{j=1}^J \delta_j(x_j)$

--

Now we can derive the efficiency conditions for our model to understand what defines the optimal allocation

---

# The competitive output model: Efficiency

Begin with the .hi-red[general case], the FOCs are defined by:
$${\partial W\over\partial x_j} = P(X) - C^j_{x_j}(x_j, e_j) = 0\rightarrow P(X) = C^j_{x_j}(x_j, e_j)$$
where ${\partial \over \partial x_j} \int_0^{X\equiv\sum x_j} P(s)ds = P(\sum x_j) = P(X)$ by the fundamental theorem of calculus, 
--
and
--
$${\partial W\over\partial e_j} = C^j_{e_j}(x_j,e_j) + D'(E) = 0 \rightarrow D'(E) = -C^j_{e_j}(x_j,e_j)$$

These $2J$ equations give us the solutions $x^*_j, e^*_j$ for $j=1,\dots,J$

---

# The competitive output model: Efficiency

The conditions are pretty straightforward, right?

--

- $P(X) = C^j_{x_j}(x_j, e_j)$ tells us that marginal benefit of consumption must equal marginal cost of consumption
--

- $-C^j_{e_j}(x_j,e_j) = D'(E)$ tells us that marginal abatement cost must equal marginal damage

--

For efficiency, we need to balance the environmental and production costs of producing the good with the benefits of consuming it

---

# The competitive output model: Efficiency

The .hi[specific case] can give us some more insight

--

Here only the $x_j$s are choice variables so we get the following FOCs:
$$P(X) = C'_j(x_j) + D'(E)\delta_j'(x_j) \,\,\,\, j=1,\dots,J$$

--

The left hand side is the marginal benefit of consumption

--

The right hand side is the total marginal cost:
  - Private production costs
  - External damage costs
  
  
---

# Efficiency in the specific case


.pull-left[
```{r uncertain_damages_permit, echo = FALSE, fig.show = 'hide', warning = F}
damage_1 <- function(x) 0 + x/2
damage_2 <- function(x) 1 + x
damage_3 <- function(x) 1 + 3*x/2
abatement_cost <- function(x) 5 - x
intersections_1 <- tibble(x = c(0, 2, 2, 8/5, 2, 8/5),
                        y = c(2, 3, 0, 17/5, 2, 0))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
ggplot() +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[5],
           y = intersections_1$y[2], yend = intersections_1$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[6],
           y = intersections_1$y[4], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[4],
           y = intersections_1$y[4], yend = intersections_1$y[4],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = damage_1, color = "#ca5670", size = 1.5) +
  stat_function(fun = damage_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = damage_3, color = "#ca5670", size = 1.5) +
  stat_function(fun = abatement_cost, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = "A", x = 2.6, y = 3) +
  annotate("text", size = 8, label = expression(delta*D*minute (delta*X)), x = 4, y = 2.5) +
  annotate("text", size = 8, label = expression(C*minute (X)), x = 3, y = 4.5) +
  annotate("text", size = 8, label = expression(P(E)), x = 4.5, y = 1) +
  annotate("text", size = 8, label = expression(C*minute (X) + delta*D*minute (delta*X)), x = 1.5, y = 4.75) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Output",
       y = "MC/MD ($)",
       title = "Efficient allocation in\nthe specific case") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4]), labels = expression(X^c, X^"*")) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[2], intersections_1$y[4]), labels = expression(P^c, P^"*"))

```

![aggregate efficiency sets SMC = SMB](`r knitr::fig_chunk("uncertain_damages_permit", "png")`)
]

.pull-right[
$X^c$ is the competitive allocation, this results in:
   - Too much production
   - Too low of an output price

$X^*$ is the optimal allocation where SMB = SMC
   - This results in less production than the competitive allocation at a higher price
 
]


---

# Efficiency in the specific case


.pull-left[
![aggregate efficiency sets SMC = SMB](`r knitr::fig_chunk("uncertain_damages_permit", "png")`)
]

.pull-right[

 Where do the aggregate curves come from?
 
 We get aggregate private MC $C'(X) = \sum_{j=1}^J C'_j(x_j)$ by *horizontally* summing firm MCs
 
 We get SMC by *vertically* summing PMC and MD
 
]

---

# Policy instruments

Now we will take another look at our environmental policy instruments in this model with output markets

--

We will see that:

--
  - There are additional results related to the output market that we didn't have before
  
--

  - The previous results all still hold: taxes, subsidies, permits can all achieve the efficient allocation
  
---

# Policy instruments: taxes

When facing an emission tax $\tau$ a competitive firm's problem in the .hi-red[general case] is:
$$\Pi_j(x_j, e_j) = p x_j - C^j(x_j,e_j) - \tau e_j$$
where $p$ is the market price of $x$

--

The firm's FOCs are:

--

\begin{align}
  p &= C'_{x_j}(x_j,e_j) \\
  \tau &= -C'_{e_j}(x_j,e_j)
\end{align}

---

# Policy instruments: taxes

The firm's FOCs are:
\begin{align}
  p &= C'_{x_j}(x_j,e_j) \\
  \tau &= -C'_{e_j}(x_j,e_j)
\end{align}

The firm equates MR to MC of production

The firm equates the marginal abatement cost to the tax level

So if the regulator sets $\tau = D'(E^*)$ .hi-blue[we can achieve the efficient allocation]

---

# Policy instruments: taxes

When facing an emission tax $\tau$ a competitive firm's problem in the .hi[specific case] where $e_j = d x_j$ is:
$$\Pi_j(x_j) = p x_j - C^j(x_j) - \tau \delta x_j$$
with FOC:
$$p = C'_j(x_j) - \tau \delta$$

--

If the regulator sets $\tau = D'(E^*)$ then firms behave as if
$$P(X) = C'_j(x_j) - \delta D'(E^*)$$
.hi-blue[which matches our social efficiency condition]

---

# Comparative statics

Now that we have an output market in our model we can study how taxes influence it

--

To start we will assume all firms are identical and we are in the .hi[specific case] of the model so our profit-maximizing firm FOC is:
$$P(X) = C'(x) + \delta\tau$$
where $X = x\cdot J$

--

Differentiate the FOC with respect to $\tau$ to get how output and emissions respond to a change in the tax rate

---

# Comparative statics

Differentiating gives us:
$$\left[P'(X)J - C''(x)\right]{dx \over d\tau} = \delta$$
which implies that:
$${dx\over d\tau} = {\delta \over P'(X)J - C''(x)} < 0$$
and ${dX\over d\tau} = J{dx\over d\tau} < 0$  

Emission taxes reduce output

---

# Comparative statics

With $E = \delta \cdot X$ we have how emissions respond to the tax:
$${dE\over d\tau} = J \delta {\delta \over P'(X)J - C''(x)} < 0$$
and since $p = P(X)$ is the market price of output, we can determine the relationship between output prices and the tax:
$${dp\over d\tau} = P'(X){d X \over d\tau} > 0$$


---

# Comparative statics

Recap: What do the comparative statics tell us?

Output and emissions decline in the tax:
  - A tax on emissions raises the marginal cost of production for firms
    - Supply shifts left
  - Output price $p$ goes up, quantity $x$ goes down
  
---

# Comparative statics

The .hi-blue[incidence]<sup>1</sup> of the tax is also made clear by:
$${dp\over d\tau} = P'(X){d X \over d\tau} > 0$$

.footnote[
Incidence is how the tax burden is distributed between consumers and producers.
]

--

If $P'(X)$ is small, demand is .hi-blue[elastic (flat), and consumers have low incidence because the price they pay does not change much in the tax, firms bear most of the cost of the tax

--

If demand is perfectly elastic $P'(X) = 0$ and there is no associated price increase from a tax increase

---

# Comparative statics

$${dp\over d\tau} = P'(X){d X \over d\tau} > 0$$

If $P'(X)$ is large, demand is .hi-blue[inelastic], and consumers have high incidence because the price they pay for $x$ increases substantially in the tax, firms pass-through most of the tax to consumers

--

If demand is perfectly inelastic, then consumers bear the entire cost of the tax
  
---

# Comparative statics: taxes recap

What did we learn?

Increasing a tax:
1. Decreases firm and aggregate emission levels
2. Decreases firm and aggregate output (even in the general case, see pg 103-104 in the book)
3. Increases output prices

---

# Policy instruments: permits

Now suppose the regulator issues $L=E^*$ permits instead of setting a tax

--

The regulator knows that the permit price $\sigma$ that clears the permit market will be $\sigma(L) = \sigma(E^*) = D'(E^*)$

--

Similarly, the output price will then be $p = P(X^*)$

--

The regulator can achieve the first-best efficient outcome

---

# Policy instruments: permits

Auctioned versus freely-distributed permits are equivalent in this model in terms of .hi-blue[efficiency]

--

The permit price in the market under free distribution will match the price that clears the permit auction

--

Output prices will also be the same because all firm and consumer decisions will be identical

--

The one way they will be different is how the .hi-blue[rents (economic profits)] are distributed: who gets the value from the scarcity of permits, the firms or the government?


---

# Distribution of rents in permit markets


.pull-left[
```{r permits_rents, echo = FALSE, fig.show = 'hide', warning = F}
damage_1 <- function(x) 0 + x/2
damage_2 <- function(x) 2 + x/2
abatement_cost <- function(x) 5 - x
intersections_1 <- tibble(x = c(0, 2, 2, 10/3, 2, 10/3),
                        y = c(2, 3, 0, 5/3, 2, 0))
polys_rent <- tibble(x = c(0, 2, 2, 0),
                     y = c(1, 1, 3, 3))
ggplot() +
  geom_polygon(data = polys_rent, aes(x = x, y = y), fill = "red", alpha = 0.2) +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[5],
           y = intersections_1$y[2], yend = intersections_1$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = polys_rent$x[1], xend = polys_rent$x[2],
           y = polys_rent$y[2], yend = polys_rent$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[6],
           y = intersections_1$y[4], yend = intersections_1$y[6],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[4],
           y = intersections_1$y[4], yend = intersections_1$y[4],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = damage_1, color = "#ca5670", size = 1.5) +
  stat_function(fun = damage_2, color = "#ca5670", size = 1.5) +
  stat_function(fun = abatement_cost, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = "A", x = 2.6, y = 3) +
  annotate("text", size = 8, label = expression(C*minute (X) + sigma), x = 2.9, y = 4.75) +
  annotate("text", size = 8, label = expression(P(X)), x = 4.5, y = 1) +
  annotate("text", size = 8, label = expression(C*minute (X)), x = 4, y = 2.55) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Output and Emissions",
       y = "MC and P ($)",
       title = "Rent distribution\nunder permits") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4]), labels = expression(X^"*", X^c)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[2], intersections_1$y[4]), labels = expression(P^"*", P^c))

```

![scarcity rents are distributed to firms or the government depending on the permit allocation scheme](`r knitr::fig_chunk("permits_rents", "png")`)
]

.pull-right[
Assume $x = \delta e$ and $\delta=1$ so we can plot them on the same scale

The red shaded area is the rents from the permit scheme

If freely allocated: they remain with firms

If auctioned: they go to the government as revenue
]

---

# Relative/intensity standards

A common form of standards is called a .hi-red[relative standard]

--

Relative standards regulate firms based on the concentration of pollution relative to some measurable output

--

A relative standard would look something like:
$$ e/x \leq \alpha$$
or equivalently: $e \leq \alpha x$
where $\alpha$ is the policy variable

--

Relative standards are often called intensity standards because $e/x$ measures the pollution intensity of output

---

# Relative/intensity standards

Relative standards are only interesting in the .hi-red[general case] of our model, in the .hi[specific case]:
- If $\delta > \alpha$, the firm has to shut down
- If $\delta \leq \alpha$, the firm complies with the regulation no matter its actions

--

Assume the regulation is binding in the .hi-red[general case] (i.e. it actually affects firm behavior), then the firm will always set $e = \alpha x$<sup>1</sup>

.footnote[
<sup>1</sup>Choosing $e$ strictly less than  $\alpha x$ strictly raises costs and reduces profits.
]

This lets us re-write a firm's profit function as: $\Pi(x) = px - C(x, \alpha x)$

---

# Relative/intensity standards

$$\Pi(x) = px - C(x, \alpha x)$$
--

If the regulation is binding, the firm really is only choosing one variable, $x$

--

The FOC is:
$$p = C_x(x,\alpha x) + \alpha C_e(x, \alpha x)$$

The implicit solution to this, $x(\alpha)$, is the firms supply function, dependent on the policy $\alpha$

---

# Relative standards vs quantity standards

Suppose the regulator wants to hit $e = \bar{e}$ and she knows $x(\alpha)$

All the regulator has to do is set $\alpha = \bar{e}/x(\alpha)$

--

Now suppose the firm just uses a quantity standard and directly sets $e = \bar{e}$

--

The firm's profit function is:
$$\Pi(x,\bar{e}) = px - C(x,\bar{e})$$
and the firm's supply $x(\bar{e})$ is defined by $p = C_x(x,\bar{e})$

--

If the regulator chooses $\bar{e} = e^*$ the regulator can achieve the efficient outcome

---

# Relative standards vs quantity standards

Recall the relative standard FOC if we wanted to set $e = \bar{e}$:
$$p = C_x(x(\alpha),\bar{e}) + \alpha C_e((\alpha), \bar{e})$$

--

and notice that this means that:
$$p - \alpha C_e(x(\alpha), \bar{e}) = C_x(x(\alpha), \bar{e}) > p$$
since $-C_e$ is positive

--

What does this mean?

---

# Relative standards vs quantity standards

This means that:
$$C_x(x(\alpha), \bar{e}) > C_x(x(\bar{e}), \bar{e})$$

--

It follows pretty simply that: $x(\alpha) > x(\bar{e})$ and that:
--

- $C(x(\alpha), \bar{e}) > C(x(\bar{e}), \bar{e})$ since $C_{xx} > 0$
--

- $-C_e(x(\alpha), \bar{e}) > -C_e(x(\bar{e}), \bar{e})$ since $C_{ex} < 0$

--

.hi-blue[Total cost and marginal production and abatement costs are higher under a relative standard]

---

# Relative standards vs quantity standards

Takeaways:

If a regulator sets an emission goal of $\bar{e}$ for a single firm or $J$ firms, then a relative standard will lead to:
--

- higher output
--

- higher total cost
--

- higher marginal abatement cost relative to a quantity standard

--

Why?

---

# Relative standards vs quantity standards

How can the firm achieve compliance under a relative standard $(e/x \leq \alpha)$?

--

Two ways:

--

1. Decrease emissions $e$
--

2. Increase output $x$

--

Relative standards allow the firm to meet the standard *in ways we don't want them to,* so we end up with too much output

This limits us to .hi-blue[second-best] outcomes

---

# Relative standards: optimal policy

If we need to use a relative standard due to political or technical reasons what standard should we set to maximize welfare?

--

The regulator's problem is:
$$W(\alpha) = \int_{0}^{x(\alpha)} P(t) dt - C(x(\alpha), \alpha x(\alpha)) - D(\alpha x( \alpha))$$

--

The FOC is:
$$[P - C_x - \alpha C_e] x'(\alpha) - C_e x - D' \times [x + \alpha x'(\alpha)] = 0$$
where the term in the first bracket is 0 from the firm's $\pi$-max FOCs

---

# Relative standards: optimal policy

This gives us that:
$$- C_e  ={x + \alpha x'(\alpha) \over x } D'$$

.hi-blue[This is not MAC = MD!]

--

$x'(\alpha)$ tells us how output responds to policy and its sign tells us whether MAC $>$ MD or vice versa

--

Suppose $x'(\alpha) > 0$, then the second best policy sets MAC $>$ MD:
- The second-best quantity of emissions is lower than the first-best / optimal level that sets MAC = MD


---

# Environmental policy in the long run

So far we have assumed a fixed number of firms $J$

In the long run, firms may enter or exit the market, and this may affect the efficiency properties of our policies

--

Econ 101 tells us that in perfectly competitive markets:

--

- Firms are identical
--

- Firms enter or exit until profits are driven to zero
--

- Firms produce at minimum average cost
--

- Optimal number of firms is endogenous (usually)

--

We will now look at long run properties of policy with identical firms

---

# Environmental policy in the long run

In the long run, firms have fixed and variable costs:
\begin{align}
C(x,e) = \begin{cases}
VC(x,e) + F \text{  if  } x,e \neq 0 \\
0  \text{  if  } x,e = 0 
\end{cases} \\
\end{align}

where $F$ is the fixed cost of entry, and $VC$ denotes variable costs of operation

--

In the long run, social welfare will depend on $x,e$ .hi-blue[and the number of firms] $J$

---

# Environmental policy in the long run

Social welfare is given by:
$$W(x,e,J) = \max_{x,e,J} \int_0^{x\cdot J} P(t) dt - J\cdot C(x,e)-D(e\cdot J)$$

The FOCs for a social optimum are:
\begin{align}
P(X^*) =& C_x(x^*,e^*) \\
D'(E^*) =& -C_e(x^*,e^*) \\
P(X^*)x^* =& C(x^*,e^*) + D'(E^*)e^*
\end{align}

The last expression is the new one for .hi[long-run efficiency]

---

# Environmental policy in the long run

With some slight rearranging we can get:
$$P(X^*) = { C(x^*,e^*) + D'(E^*)e^*\over x^*}$$

What does this say?

--

First, for a small firm, $D'(E^*)e^*$ is approximately the damage caused by that firm because for sufficiently small $e^*$, $D'(E^*)$ will be approximately constant $(\delta)$ (by a Taylor expansion argument)


---

# Environmental policy in the long run

With some slight rearranging we can get:
$$P(X^*) = \underbrace{{C(x^*,e^*) + \delta e^*\over x^*}}_{\text{average social cost}}$$


This means that it is socially efficient for firms to enter or exit until the price of output (approximately) equals the .hi-blue[average social cost curve]


---

# Standards in the long run

In the short run we had that standards were equivalent to taxes and tradable permits

Is this true in the long run?

--

Suppose the regulator wants to cap total emissions at $E^*$

--

She sets an emission standard $e^* = E^*/J^*$ for all firms where $J^*$ is the optimal long run number of firms

--

Firms can now only choose $x$ since $e$ is fixed at $e^*$

---

# Standards in the long run

Firms choose output according to:
$$p = C_x(x,e^*)$$

--

In the long run equilibrium we will have $\hat{J}$ firms all producing $\hat{x}$ units of output such that:
\begin{align}
  &P(\hat{x}\hat{J}) = C_x(\hat{x},e^*) \\
  &P(\hat{x}\hat{J})\hat{x} - C(\hat{x},e^*) = 0
\end{align}

--

MR = MC, and zero profits are our two equilibrium conditions

--

MR = MC maximizes firm profit, zero profits ensures no change in # of firms

---

# Standards in the long run

Recall that long run efficiency required that:
$$P(X^*)x^* - C(x^*,e^*) = D'(E^*)e^* > 0$$

--

so we have that $\hat{J} \neq J^*$ and $\hat{x} \neq x^*$!

What's the intuition?

--

When we impose a standard:

--

1. Firms cut back output
--

2. This raises marginal profit above zero: MR > MC!
--

3. Firms enter the market so $\hat{J} > J^*$ and we will have that $\hat{x} < x^*$

---

# Standards in the long run

This is important!

--

Conventional wisdom tells us that taxes and permits are only better to use than standards because they equate MACs across firms (achieving least-cost) without requiring knowledge of MACs

--

This is only true in the short run

--

In the long run, standards do not appropriate the damage to the environment $D'(E^*)e^*$ from firms, so we get .hi-red[excess entry] and standards are no longer first-best

---

# Taxes in the long run

Can taxes achieve the efficient outcome in the long run?

--

Yes and it is pretty easy to see, suppose the regulator sets a tax of:
$$\tau = D'(E^*)$$

--

Firm profit is then:
$$\Pi = px - C(x,e) - \tau e$$

--

Giving us FOCs:
$$p = C_x(x^*,e^*) \qquad -C_e(x^*,e^*) = \tau$$

---

# Taxes in the long run

In the long run firms enter until profits are zero so:
$$\Pi = P(X^*)x^* - C(x^*,e^*) - \tau e^* = 0$$
so $\tau = D'(E^*)$ implies that
$$P(X^*)x^* = C(x^*,e^*) + D'(E^*)e^*$$

The firm FOCs for production and the entry zero-profit condition map exactly to the social welfare maximizing conditions!

--

The payment of tax rents from the firms to the regulator of $\tau e^* = D'(E^*)e^*$ limits entry and is what makes taxes efficient in the long run

---

# Permits in the long run

Now what about permit systems?

--

Suppose the regulator auctions off $L=E^*=e^*J$ permits and let $\sigma$ be the market-clearing permit price

--

The long run equilibrium is defined by the two firm FOCs and the entry condition:
\begin{align}
  &P(x^*J) = C_x(x^*,e^*) \\
  &\sigma = -C_e(x^*,e^*) \\
  &P(x^*J)x^* - C(x,e) - \sigma e^* = 0
\end{align}

---

# Permits in the long run

--

Similar to the short run we will have that $\sigma = -C_e(x^*,e^*) = D'(L) = D'(E^*)$

--

Thus the three long run efficiency conditions are satisfied again if we auction off the permits:
- MR = MC
- MAC = MD
- Zero profit

--

Now what if we freely distribute permits?

--

It seems like it might not be long run efficient since firms are not paying the environmental rent, so zero profit might not occur

---

# Permits in the long run

Suppose we allocate $\bar{e}$ permits to the incumbent identical firms, profit for the incumbent firms given permits is then:
$$\Pi = px - C(x,e) - \sigma(e-\bar{e})$$
and profit for any future entrants who were not given an allocation is:
$$\Pi = px - C(x,e) - \sigma e$$

What two things do you notice?

--

First, for any given $\sigma$, entrants and incumbents cannot both have zero profit!

---

# Permits in the long run

Our efficiency condition is now: .hi-red[new firms enter until profits are zero]
$$\Pi = px - C(x,e) - \sigma e = 0$$
so that incumbent firms sustain long run profits of:
$$\sigma\bar{e}$$

--

What is this saying?

--

.hi-blue[Operating profits] of any firms in the market are zero, incumbents had long run profits **only** from their initial permit allocation

---

# Permits in the long run

The second thing you should have noticed is that the firm FOCs will be identical to the auctioned permit case, firms face the .hi-blue[exact same] incentives for output and emissions

--

This means that freely allocated permits are also long run efficient

--

This is just an application of the Coase theorem: the initial assignment of property rights to pollute does not matter for efficiency

---

# Subsidies in the long run

Now what about subsidies?

In the short run they are equivalent to taxes, are they still equivalent in the long run?

--

Think about the incentives for entry...

--

Denote the reference level of emissions as $\hat{e}$, firm profits under a subsidy per unit $\xi$ are:
$$Pi = px - C(x,e) - \xi(e-\hat{e})$$

---

# Subsidies in the long run

For damage efficiency we need to set the subsidy equal to MD $\xi = D'(E^*)$

--

In the long run equilibrium, firms enter until profits are zero
$$\Pi = P(X^*)x^* - C(x^*,e^*) - D'(E^*)(e^*-\hat{e}) = 0$$

--

which implies that
$$P(X^*)x^* - C(x^*,e^*) - D'(E^*)e^* = D'(E^*)\hat{e} > 0$$

Too many firms have entered!

---

# Subsidies in the long run

Why did too many firms enter? 

--

Payments are available to **all** firms and induces excess market entry

--

Permits do not have these problems because the payment was only to incumbent firms, not entrants

--

Incumbent firms are already in the market: giving them the rents from freely distributed permits does not lead to excess entry