---
title: "Lecture 8"
subtitle: "Environmental policy with pre-existing distortions"
author: Ivan Rudik
date: AEM 6510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'


---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, tidylog, xaringanExtra, rlang, patchwork
)
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

---

# Roadmap

So far we have looked at single sector (partial equilibrium) economies with:
- Pollution distortions
- Competitive markets
- Market power distortions

--

Now we will learn about multi-sector (general equilibrium) economies

How does environmental policy spillover into these other sectors?

How does environmental policy interact with revenue-raising taxes (e.g. income taxes)?

---

# Environmental policy with leisure

First we extend the model so that labor supply is .hi-blue[elastic]
  - Households have a choice of either working or leisure time
  
--

To focus on the key intuition we assume: $^1$
  - There is a representative (single) firm
  - There is a representative household

This allows us to treat individual and aggregate behavior the same

.footnote[1: The underlying critical assumption is that utility and profit functions take what's called a Gorman form.]

---

# Environmental policy with leisure

Define the following:
- $X$ is consumption of the polluting good
- $Z$ is consumption of the *numeraire* good (i.e. the relative good)
- $N$ is the hours of leisure time
- $E$ is aggregate emissions

--

Utility is then:
$$U(X,Z,N,E)  = U(X,N) + Z - D(E)$$
where $U_{XX}, U_{NN} < 0$ and  $U_{XX}U_{NN}-U_{NX}^2 > 0$ and the person is endowed with some amount of time $T$ to allocate between work and leisure

---

# Environmental policy with leisure

Wages earned by the household are $w$, and we assume demand for labor is perfectly elastic
  - Demand is horizontal at $w$
  
Household income is then: $w\cdot(T-N)$

We can now write the households utility maximization problem as:
$$\max_{X,N} U(X,Z,N,E)  = U(X,N) + Z - D(E) \text{ subject to: } w\cdot(T-N) = Z + pX$$
---

# Environmental policy with leisure

We can substitute the budget constraint in for $Z$ to get an unconstrained problem:
$$\max_{X,N} U(X,Z,N,E)  = U(X,N) + w\cdot(T-N) - pX - D(E)$$
with FOCs:
$$U_X = p \qquad U_N = w$$
which implicitly define the demand function for consumption $X(p,w)$ and the demand function for leisure $N(p,w)$

---

# Environmental policy with leisure

How do choices of $X,N$ respond to a change in price $p$?

Differentiate both FOCs with respect to p:
$$U_{XX} {\partial X \over \partial p} + U_{XN} {\partial N \over \partial p} = 1 \qquad U_{NN} {\partial N \over \partial p} + U_{XN} {\partial X \over \partial p} = 0$$

We have two equations and two unknowns so we can solve to get:
$${\partial N \over \partial p} = {-U_{XN} \over U_{XX}U_{NN} - U_{XN}^2} \qquad {\partial X \over \partial p} = {U_{NN} \over U_{XX}U_{NN} - U_{XN}^2}$$

---

# Environmental policy with leisure

$${\partial N \over \partial p} = {-U_{XN} \over U_{XX}U_{NN} - U_{XN}^2} \qquad {\partial X \over \partial p} = {U_{NN} \over U_{XX}U_{NN} - U_{XN}^2}$$

${\partial X \over \partial p}$ is negative since $U$ is concave $(U_{NN} < 0, U_{XX}U_{NN} - U_{XN}^2 > 0)$

The sign of ${\partial N \over \partial p}$ equals the sign on $-U_{XN}$

If $X$ and $N$ are substitutes, $-U_{XN} > 0$, and leisure increases in the price of the consumption good

If they are complements, $-U_{XN} < 0$, and leisure decreases in the price of the consumption good

---

# Environmental policy with leisure

The firm side of the economy will be the same as before: it produces $X$ and emits $E$ and for simplicity we will focus on the specific case:
$$\Pi = pX - C(X) \text{ where } E = \delta X$$

We will also assume:

- $\delta=1$ so we can use $E$ and $X$ interchangeably
- $C'(X) > 0, C''(X) \geq 0$
- The polluting industry's demand for labor is small relative to the entire economy, i.e. wages are effectively fixed

---

# Environmental policy with leisure

Now lets solve for the optimum:
$$\max_{X} W = U(X,N) + w\cdot(T-N) - pX - D(X) + pX - C(X)$$
To focus on interactions with non-regulated industries, we assume the regulator cannot determine leisure

--

The consumer chooses $N$ and therefore $Z$ according to the budget constraint $Z = w(T-N) - pX$ and FOC $U_N(X^*,N) = w$

--

One way you can think about this is as if the regulator imposes a quantity standard $X^*$ and then a market price $p^*$ arises which affects leisure demand

---

# Environmental policy with leisure

The FOC for the optimum is:
$$U_X - D'(X) - C'(X) + [U_N -w]{\partial N \over \partial X} = 0$$
where the last term captures the households .hi-blue[indirect] leisure response to the regulator's policy choice

--

Given household utility maximization $U_N -w = 0$ and the condition is then:
$$U_X - C'(X) = D'(X)$$

--

Marginal abatement cost $(U_X - C'(X))$ equals marginal damage $(D'(X))$ !

---

# Environmental policy with labor market distortions

Now suppose the government needs to raise revenue with a labor income tax $m$

--

It needs to finance a budget of size $G$

--

The consumer's utility maximization problem is:
\begin{gather}
  \max_{X,Z,N} U = u(X,N) + Z - D(E) \\
  \text{ subject to } (1-m)w(T-N) = Z + pX
\end{gather}

Where the budget is scaled down by $(1-m)$ reflecting the income tax

---

# Environmental policy with labor market distortions

The FOCs are:
$$u_X = p \qquad u_N = (1-m)w$$

--

The labor tax causes an inefficiency in the labor market: the marginal value of leisure $(u_N)$ is no longer equal to the marginal value of labor $(w)$

--

$u_N < w$ which means that the household overconsumes leisure

--

Another way to see this is to re-write the FOC as:
$$u_N + mw = w$$

--

The tax $m$ makes the consumer act as if there is a subsidy $mw$ on leisure

---

# Environmental policy with labor market distortions

.pull-left[
```{r labor_market_1, echo = FALSE, fig.show = 'hide', warning = F}

# demand <- function(x) 5 - x
demand <- function(x) 3.5 - x
intersections_1 <- tibble(x = c(5, 5, 0, 0, 5, 2.5, 2.5),
                        y = c(0, 2, 2, 1, 1, 0, 2))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
ggplot() +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[2],
           y = intersections_1$y[1], yend = intersections_1$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[5],
           y = intersections_1$y[4], yend = intersections_1$y[5],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[6], xend = intersections_1$x[7],
           y = intersections_1$y[6], yend = intersections_1$y[7],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = demand, color = "darkslateblue", size = 1.5) +
  annotate("text", size = 8, label = expression(N(p^c,w)), x = 0.75, y = 3.5) +
  annotate("text", size = 8, label = "b", x = 2.25, y = 1.75) +
  annotate("text", size = 8, label = "G", x = 4, y = 1.5) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Leisure",
       y = "Wage ($)",
       title = "Income taxation causes DWL") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[6]), labels = expression(T, 0, N^c)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[2], intersections_1$y[4], intersections_1$y[6]), labels = expression(w, "(1-m)w", ))

```


![excess burden is because MU of leisure is less than wage](`r knitr::fig_chunk("labor_market_1", "png")`)
]

.pull-right[

$N^c$ is how much leisure the consumer chooses, since $(1-m)w < w$ this is too much and induces DWL equal to $b$

The tax raises revenues equal to $G$

]


---

# Environmental policy with labor market distortions

.pull-left[
```{r labor_market_2, echo = FALSE, fig.show = 'hide', warning = F}

demand_2 <- function(x) 5 - x
demand <- function(x) 3.5 - x
intersections_1 <- tibble(x = c(5, 5, 0, 0, 5, 2.5, 2.5, 4, 4),
                        y = c(0, 2, 2, 1, 1, 0, 2, 2, 0))
polys_dwl <- tibble(x = c(2, 3, 3),
                     y = c(3, 4, 2))
ggplot() +
  annotate("segment", x = intersections_1$x[1], xend = intersections_1$x[2],
           y = intersections_1$y[1], yend = intersections_1$y[2],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[2], xend = intersections_1$x[3],
           y = intersections_1$y[2], yend = intersections_1$y[3],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[4], xend = intersections_1$x[5],
           y = intersections_1$y[4], yend = intersections_1$y[5],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[6], xend = intersections_1$x[7],
           y = intersections_1$y[6], yend = intersections_1$y[7],
           linetype = "dashed", size = 1, color = "black") +
  annotate("segment", x = intersections_1$x[8], xend = intersections_1$x[9],
           y = intersections_1$y[8], yend = intersections_1$y[9],
           linetype = "dashed", size = 1, color = "black") +
  stat_function(fun = demand, color = "darkslateblue", size = 1.5) +
  stat_function(fun = demand_2, color = "#ca5670", size = 1.5) +
  annotate("text", size = 8, label = expression(N(p^c,w)), x = 0.75, y = 3.5) +
  annotate("text", size = 8, label = expression(N(p^"*",w)), x = 2.25, y = 3.5) +
  annotate("text", size = 8, label = "e", x = 3, y = 1.5) +
  annotate("text", size = 8, label = "c", x = 3.75, y = 1.75) +
  annotate("text", size = 8, label = "d", x = 4.5, y = 1.5) +
  theme_minimal()  +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) +
  labs(x = "Leisure",
       y = "Wage ($)",
       title = "Income taxation causes DWL") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$x[2], intersections_1$x[4], intersections_1$x[6], intersections_1$x[9]), labels = expression(T, 0, N^c, N^"*")) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5),
                     breaks = c(intersections_1$y[2], intersections_1$y[4], intersections_1$y[6]), labels = expression(w, "(1-m)w", ))

```


![excess burden is because MU of leisure is less than wage](`r knitr::fig_chunk("labor_market_2", "png")`)
]

.pull-right[

Suppose the regulator sets $X=X^*$ where $X^* \rightarrow MAC = MD$

This raises the price of $X$, shifts leisure demand to the .hi-red[right]

New DWL is $c$, and government revenues are now only $d$

Change in DWL from $X^c \rightarrow X^*$ is indeterminant

]


---

# Second-best environmental policy

What does the optimal environmental policy look like if there's a pre-existing labor market distortion?

The government has a budget $G$ it needs to finance using labor taxes or emission taxes

First let's consider the case where they can only raise revenue via a labor tax: this is non-revenue raising environmental policy

---

# Second-best non-revenue raising environmental policy

If we cannot raise revenue with the environmental policy, the regulator chooses $X$ (and $E$) and the marginal tax rate $m$ to maximize the sum of profit and utility, subject to the budget constraint

--

The household consumes leisure according to the FOC: $U_N(\bar{X},N) = (1-m)w$ given the regulator chose $X=\bar{X}$

--

The firm obtains profits:
$$\Pi = p\bar{X} - C(\bar{X})$$

---

# Second-best non-revenue raising environmental policy

The marginal value of the dirty good comes from the consumers inverse demand:
$$P(\bar{X}) = u_X(\bar{X},N)$$
which depends on $N$

--

How do the endogenous variables $N$ and $p$ vary with $\bar{X}$?

Let's look at the comparative statics and differentiate the consumer FOCs wrt $\bar{X}$

---

# Second-best non-revenue raising environmental policy

\begin{align}
  u_{XX} + u_{XN} \frac{\partial N}{\partial \bar{X}} &= \frac{\partial p}{\partial \bar{X}} \\
  u_{NX} + u_{NN} \frac{\partial N}{\partial \bar{X}} &= 0
\end{align}

which gives us:
\begin{align}
 \frac{\partial N}{\partial \bar{X}} &= - {u_{XN} \over u_{NN}} \\
 \frac{\partial p}{\partial \bar{X}} &= {u_{XX}u_{NN} - u_{NN}^2 \over u_{NN}} < 0
\end{align}
The sign of $\frac{\partial N}{\partial \bar{X}}$ depends on whether $X$ and $N$ are complements or substitutes

---

# Second-best non-revenue raising environmental policy

The regulator's problem is then:
$$\max_{X,m} u(X,N) + Z - D(X) + pX - C(X) \text{ s.t. } wm(T-N) = G$$

--

We still need to decide what the government does with its revenue

--

For convenience, we assume its returned to the consumer as a lump sum transfer so that:
$$Z=(1-m)w(T-N)-pX+G = w(T-N) - pX$$
Income is unchanged for a given level of $N$ under the tax and transfer


---

# Second-best non-revenue raising environmental policy

The regulator's problem is then:
$$\max_{X,m} u(X,N) +  w(T-N) - D(X) - C(X) + \lambda[ wm(T-N)  -G]$$
$\lambda$ is the marginal welfare cost of public funds

--

It measures the welfare costs of raising revenue by taxing labor

--

What's the FOC for $m$?

---

# Second-best non-revenue raising environmental policy

The FOC for $m$ is:
$$(u_N - w){\partial N \over \partial m} + \lambda \left[w(T-N)-wm {\partial N \over \partial m} \right] = 0$$

The household's optimal choice of $N$ tells us that: $-mw = u_N - w$, we can substitute this in to get $\lambda$:
$$\lambda = \frac{w  m {\partial N \over \partial m}}{w(T-N)-wm{\partial N \over \partial m}}$$

Whats the interpretation of the right hand side?

---

# Second-best non-revenue raising environmental policy

$$\lambda = \frac{w  m {\partial N \over \partial m}}{w(T-N)-wm{\partial N \over \partial m}}$$

--

The numerator is:

--

The welfare cost of changing $m$

--

Why?

--

Higher $m$ increases leisure demand ${\partial N \over \partial m}$

--

This times the tax wedge $mw$ gives us the change in excess burden (i.e. the DWL $d$)

---

# Second-best non-revenue raising environmental policy

$$\lambda = \frac{w  m {\partial N \over \partial m}}{w(T-N)-wm{\partial N \over \partial m}}$$

The denominator is:

--

The change in tax revenue

--

First term is the increase in marginal revenue

--

Second term is the decrease in infra-marginal revenue

---

# Second-best non-revenue raising environmental policy

$$\lambda = \frac{w  m {\partial N \over \partial m}}{w(T-N)-wm{\partial N \over \partial m}}$$

Numerator and denominator combined give us the ratio of change in welfare from a change in $m$ to the change in tax revenue from a change in $m$

--

This is the change in welfare from a change in tax revenue!

---

# Second-best non-revenue raising environmental policy

Now consider the FOC for $X$:
$$u_X - D'(X) - C'(X) + \left[ u_N - w - \lambda w m \right] {\partial N \over \partial X} = 0$$

Recall that we know:
- $-wm = u_N - w$
- $\frac{\partial N}{\partial X} = \frac{\partial N}{\partial p}\frac{\partial p}{\partial X}$

This gives us that:
$$u_X - C'(X) + (1+\lambda)\left[-{\partial N \over \partial p}\frac{\partial p}{\partial X}\right] = D'(X)$$

---

# Second-best non-revenue raising environmental policy

$$u_X - C'(X) + (1+\lambda)\left[-{\partial N \over \partial p}\frac{\partial p}{\partial X}\right]wm = D'(X)$$

--

What are each of the terms:

--

$u_X - C'(X)$ is the marginal abatement cost

--

$D'(X)$ is marginal damage

--

$(1+\lambda)\left[-{\partial N \over \partial p}\frac{\partial p}{\partial X}\right]$ is new

--

What's the interpretation?

---

# Second-best non-revenue raising environmental policy

$(1+\lambda)\left[-{\partial N \over \partial p}\frac{\partial p}{\partial X}\right]wm$ is called the .hi-blue[marginal interaction effect (MIE)]

It tells us how the optimal choice of $X$ departs from $X^*$ because of the labor market distortion

--

We need to account for this because the household's choice of leisure will respond to changes in $X$

--

Suppose $N$ and $X$ are substitutes, what does this mean?

---

# Second-best non-revenue raising environmental policy

Substitutes means that $MIE > 0$

--

The marginal social cost of abatement $(MAC + MIE)$ has become *larger*

--

Intuition?

--

Its more socially costly to reduce $X$ because the household increases $N$ in responses

--

This .hi-red[exacerbates] the distortion caused by the income tax

--

The opposite is true if $X$ and $N$ are complements, $MIE < 0$ and addressing the pollution distortion helps address the labor market distortion


---

# Second-best non-revenue raising environmental policy

.pull-left[
![excess burden is because MU of leisure is less than wage](`r knitr::fig_chunk("labor_market_2", "png")`)
]

.pull-right[

$N^c \rightarrow N^*$ when $p^c \rightarrow p^*$ because of a change in $X$

This is $-{\partial N \over \partial p}\frac{\partial p}{\partial X}$

This reduces tax revenue by $e+c$ which is just 
$$(N^*c* - N^c)(w - (1-m)w) \\
= (N^* - N^c)mw$$

]

---

# Second-best non-revenue raising environmental policy

The marginal welfare cost of recouping the lost tax revenue by raising $m$ is $\lambda$ giving us a total welfare cost of:
$$\lambda (N^* - N^c)mw$$

But $(N^* - N^c)mw$ is also the excess burden: its a .hi-red[direct welfare loss!]

So the total welfare loss is:
$$(1+\lambda) (N^* - N^c)mw$$

The discrete version of MIE!

---

# Findings recap

If there's a gov't revenue constraint, and it can only be met with labor taxes then:

1. The marginal social cost of reducing $X$ is higher than the first-best if $X$ and $N$ are substitutes and lower if they are complements
2. The optimal level of pollution is larger if they are substitutes, lower if they are complements
3. The absolute value of the difference in first and second-best pollution levels is larger if:
  - Labor supply is more elastic
  - Demand for $X$ is more inelastic
  - Elasticity of substitution between $N$ and $X$ is greater
  
  
---

# Second-best non-revenue raising environmental policy

We didn't actually show the last part yet

--

First define:
- $\varepsilon_x$ as the own price elasticity ${\frac{\partial X}{\partial p} / \frac{p}{X}}$
- $\eta_{XN}$ as the elasticity of substitution ${\frac{\partial X}{\partial w} / \frac{(1-m)w}{X}}$

and take advantage of the .hi-blue[Slusky symmetry condition] $\partial N / \partial p = \partial X / \partial w$

We can then use these to substitute into the MIE and get:
$$MIE = (1+\lambda)\left[- {\eta_{XN} \over \varepsilon_X}\right]p {m \over 1-m}$$

---

# Second-best non-revenue raising environmental policy

$$MIE = (1+\lambda)\left[- {\eta_{XN} \over \varepsilon_X}\right]p {m \over 1-m}$$
MIE bigger if $|\eta_{XN}|$ is bigger

--

MIE bigger if $|\varepsilon_X|$ is smaller (more inelastic)

--

Still need to show labor supply part

---

# Second-best non-revenue raising environmental policy

Define the elasticity of labor supply *at the net wage* as:
$$\varepsilon_{L} = - {\partial N \over \partial w}{(1-m)w \over L}$$

And recognize that:
$${\partial N(p,(1-m)w) \over \partial m} = -w {\partial N(p,w) \over \partial w}$$

Combining these two gives:
$${\partial N \over \partial m}m = \varepsilon_{L} L m / (1-m)$$

---

# Second-best non-revenue raising environmental policy

Finally, put:
$${\partial N \over \partial m}m = \varepsilon_{L} L m / (1-m)$$
in
$$\lambda = \frac{w  m {\partial N \over \partial m}}{w(T-N)-wm{\partial N \over \partial m}}$$

---

# Second-best non-revenue raising environmental policy

To get:
$$\lambda = \frac{\varepsilon_{L}m/(1-m)}{1- \varepsilon_{L}m/(1-m)}$$

Side note: 
If labor is more elastic $\varepsilon_{L}$ is larger, the numerator is larger, denominator is smaller $\rightarrow$ MIE is bigger

If labor supply is perfectly inelastic (vertical), there is no welfare cost of public funds!

---

# Second-best non-revenue raising environmental policy

Finally use:
$$\lambda = \frac{\varepsilon_{L}m/(1-m)}{1- \varepsilon_{L}m/(1-m)}$$
to get:
$${m\over(1-m)} = {\lambda \over (1-\lambda)\varepsilon_L}$$

and then substitute into our MIE expression

---

# Second-best non-revenue raising environmental policy

Finally we will get:
$$MIE = \lambda {-\eta_{XN} \over \varepsilon_X} {p \over \varepsilon_L}$$

Bigger $\varepsilon_L$ ... huh?

---

# Revenue raising environmental policy