---
title: "Lecture 10"
subtitle: "Deforestation // Regression discontinuity"
author: Ivan Rudik
date: AEM 6510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'


---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, xaringanExtra, rlang, patchwork, nycflights13, ggthemes
)
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

---

# Roadmap

- Can we exploit situations when we know the mechanism for treatment assignment?
- Can exploit situations where some units are just above some threshold to get treatment, and others are just below the threshold?
- Do deforestation policies work?

---

class: inverse, center, middle
name: hr2021

# Regression discontinuity

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Regression discontinuity design

Suppose we want to estimate the effect of air quality on property values

--

We have a potentially large problem: there's quite a lot of things that affect housing values that may also be correlated with air quality
  - wealth, temperature, density, etc
  
--

This makes $E[Y^0 | D = 1] - D[Y^0 | D = 0] \neq 0$, biasing any estimates because air quality will also pick up on these other factors that determine property values

--

Here we will understand one way we can break this bias by exploiting .hi-red[discontinuities]

---

# Regression discontinuity design

The Clean Air Act mandates a set of National Ambient Air Quality Standards (NAAQS)

--

The NAAQS require counties to keep ambient air concentrations of six pollutants below some (arbitrary-ish) threshold

--

If a county goes above one of these thresholds, call them $c_0$, it is deemed to be in .hi-red[non-attainment]

---

# Regression discontinuity design

Non-attainment counties then face more stringent regulatory oversight and take up a host of actions aimed at reducing pollution back down to acceptable levels

--

Non-attainment counties had .hi-red[much] greater pollution reductions during the 1970s and 1980s compared to attainment counties (Chay and Greenstone, 2005)

---

# Regression discontinuity design

The NAAQS generate a .hi-red[discontinuity:] counties just above the threshold in some year $T$ face greater pressure to reduce pollution in subsequent years $T+1, T+2, \dots$ compared to counties just below the threshold

--

In the absence of this policy we might expect counties with $R + \epsilon$ levels of pollution to be similar to counties with $R + \epsilon$ in terms of .hi-blue[all other factors] (on average)
  - i.e. $E[Y^0 | D = 1] - D[Y^0 | D = 0] = 0$

--

Thus any differences in property values is likely due to the NAAQS-induced decline in pollution in non-attainment counties

---

# Regression discontinuity design

This is the essence of the .hi-blue[regression discontinuity design (RDD)]

--

Two counties with high vs low pollution are generally not comparable (e.g. high pollution areas have other disamenities, lower wealth)

--

What if instead of comparing the two whole groups we did the following

--

Compare two counties that have *virtually identical* levels of pollution,
--
 but 
--
where one is above a threshold that forces it to significantly reduce pollution, and the other is the below the threshold

--

The subsequent difference in pollution and other outcomes we may consider .hi-red[as good as random]

---

# Regression discontinuity design

RDD is appropriate whenever a units entry into treatment $D$ .hi[jumps] in probability when some other variable $X$, called the running variable, is above some threshold $c_0$

--

This makes sense right?

--

Jumps are relatively unnatural, things typically change smoothly

--

In RDD the jump is the chance of being put into treatment (in our example, under more regulatory scrutiny)

---

# Regression discontinuity design

With RDD we use our knowledge about how counties are selected into treatment (by going above the threshold $c_0$) in order to estimate the average treatment effect

--

More precisely: we compare units just below to just above $c_0$ to estimate something called a .hi-blue[local average treatment effect (LATE)]

--

The local is because the estimate is only a valid ATE for units that have pollution levels near the threshold $c_0$

--

Here's what we're doing in pictures

---

# Regression discontinuity design: graphs

```{r}
set.seed(12345)
late <- 0 # local average treatment effect
n_obs <- 1000 # number of observations
rdd_df <- tibble(
  state = seq(1, n_obs)) %>%  # control/untreated potential outcome
  mutate(
    X = rnorm(n(), 50, 25), # running variable
    D = X > 50,
    Y1 = 500 + late*D - 1.5*X + rnorm(n(), 0, 20)
  ) %>% 
  filter(X > 0) %>% 
  select(
    state, D, X, Y1
  )
```

---

# Regression discontinuity design: graphs

.pull-left[

```{r rdd_base, echo = FALSE, fig.show = 'hide', warning = F, fig.width = 8}
ggplot(rdd_df) +
  geom_point(aes(x = X, y = Y1, color = D, shape = D), size = 1) +
  geom_vline(xintercept = 50, color = "grey55", linetype = "dashed") +
  geom_smooth( aes(x = X, y = Y1, color = D), 
               se = FALSE, 
               method = lm,
               formula = y ~ x) +
  annotate("text", size = 8, label = "Attainment", x = 25, y = 300) +
  annotate("text", size = 8, label = "Non-Attainment", x = 85, y = 300) +
  scale_color_colorblind() +
  theme_minimal() +
  labs(
    x = "Current Pollution (X)",
    y = "Future Housing Values (Y1)",
    title = "The relationship without the NAAQS"
  ) +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 24), axis.text.y = element_text(size = 24),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) 
```

![RDD is smooth without treatment](`r knitr::fig_chunk("rdd_base", "png")`)


]

.pull-right[

Suppose the pollution threshold is at 50

In the absence of the NAAQS, we expect a smooth/continuous transition in housing values above vs below the threshold

Next, suppose we implement the NAAQS

]


---

# Regression discontinuity design: graphs

```{r}
set.seed(12345)
late <- 40 # local average treatment effect (NOW 40)
n_obs <- 1000 # number of observations
rdd_df <- tibble(
  state = seq(1, n_obs)) %>%  # control/untreated potential outcome
  mutate(
    X = rnorm(n(), 50, 25), # running variable
    D = X > 50,
    Y1 = 500 + late*D - 1.5*X + rnorm(n(), 0, 20)
  ) %>% 
  filter(X > 0) %>% 
  select(
    state, D, X, Y1
  )
```

---

# Regression discontinuity design: graphs

.pull-left[

```{r rdd_naaqs, echo = FALSE, fig.show = 'hide', warning = F, fig.width = 8}
ggplot(rdd_df) +
  geom_point(aes(x = X, y = Y1, color = D, shape = D), size = 1) +
  geom_vline(xintercept = 50, color = "grey55", linetype = "dashed") +
  geom_smooth( aes(x = X, y = Y1, color = D), 
               se = FALSE, 
               method = lm,
               formula = y ~ x) +
  annotate("text", size = 8, label = "Attainment", x = 25, y = 300) +
  annotate("text", size = 8, label = "Non-Attainment", x = 85, y = 300) +
  annotate("segment", x = 51, xend = 51, 
           y = 425, yend = 465,
           linetype = "solid", size = 2, color = "red") +
  annotate("segment", x = 51, xend = 100, 
           y = 445, yend = 445,
           linetype = "solid", size = 2, color = "red") +
  annotate("text", size = 8, label = "LATE", x = 110, y = 445, color = "red") +
  scale_color_colorblind() +
  theme_minimal() +
  labs(
    x = "Current Pollution (X)",
    y = "Future Housing Values (Y1)",
    title = "The relationship without the NAAQS"
  ) +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 24), axis.text.y = element_text(size = 24),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#eeeeee", colour = NA),
    plot.background = element_rect(fill = "#eeeeee", colour = NA),
    axis.line = element_line(colour = "black")
  ) 
```

![RDD LATE is around the threshold](`r knitr::fig_chunk("rdd_naaqs", "png")`)


]

.pull-right[

The policy induced greater pollution reductions in the non-attainment counties

Housing prices shift up for those counties

The vertical distance between the two groups at 50 is our .hi-blue[local average treatment effect]

]

---

# Regression discontinuity design: the data

We've seen what RDDs look like, what do we need to do one?

--

RDD is about finding these .hi[jumps] in the probability of treatment as we move along some other variable $X$

--

How do we find these jumps?

--

For environmental topics they're often embedded in rules (e.g. the NAAQS), or across space (e.g. deforestation policy)

---

# Regression discontinuity design: the data

Good and plausible RDDs often involve $X$s having a 'hair trigger' that's not tightly related to the outcome
  - e.g. being 10 meters on either side of the Bolivia/Brazil border is pretty arbitrary in the grand scheme of things, but a massive discontinuity in deforestation policy
  
--

We will need to be focused on the area right around this hair trigger threshold: that means we will need a lot of data near $c_0$ in order to have precise estimates of the LATE

---

# Regression discontinuity design: the types

There are two classes of RDD: sharp and fuzzy

.hi-blue[Sharp designs] are where the probability of treatment increases from 0 to 1 at the threshold $c_0$

.hi-red[Fuzzy designs] are where the probability of treatment increases discontinuously at $c_0$, but not necessarily from 0 to 1

--

We will be focusing on sharp designs to keep it simple