---
title: "Lecture 1"
subtitle: "Intro, math and micro review"
author: Ivan Rudik
date: AEM 6510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'


---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, tidylog, xaringanExtra, Ryacas, rlang
)
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
red_pink <- "#e64173"
blue <- "#3C93DC"
red <- "#ff0000"
# A blank theme for ggplot
theme_empty <- theme_minimal() +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_text(size = 24), axis.text.y = element_text(size = 24),
    axis.title.x = element_text(size = 24), axis.title.y = element_text(size = 24),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#ffffff", colour = NA),
    plot.background = element_rect(fill = "#ffffff", colour = NA),
    axis.line = element_line(colour = "black"), axis.ticks = element_line(),
  )
theme_blank <- theme_minimal() +
  theme(
    legend.position = "none",
    title = element_text(size = 24),
    axis.text.x = element_blank(), axis.text.y = element_blank(),
    axis.title.x = element_blank(), axis.title.y = element_blank(),
    panel.grid.minor.x = element_blank(), panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank(),
    panel.background = element_rect(fill = "#ffffff", colour = NA),
    plot.background = element_rect(fill = "#ffffff", colour = NA),
    axis.line = element_blank(), axis.ticks = element_blank(),
  ) 
```

```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

```{css, echo=FALSE}
/* custom.css */
.left-code {
  color: #777;
  width: 38%;
  height: 92%;
  float: left;
}
.right-plot {
  width: 60%;
  float: right;
  padding-left: 1%;
}
.plot-callout {
  height: 225px;
  width: 450px;
  bottom: 5%;
  right: 5%;
  position: absolute;
  padding: 0px;
  z-index: 100;
}
.plot-callout img {
  width: 100%;
  border: 4px solid #23373B;
}
```

---

# AEM 6510: Environmental and resource economics

.hi-blue[Welcome to AEM 6510]

--

**Class time:** Tuesday and Thursday 3:00-4:15

**Instructor:** Ivan Rudik

**Office hours:** Tuesday and Thursday 4:20-4:50

**Teaching assistant:** Weiliang Tan

**TA office hours:** Monday 3:00-4:00

---

# AEM 6510: Environmental and resource economics

**Textbooks:** 
- A Course in Environmental Economics by Phaneuf and Requate
- [Causal Inference: The Mixtape](https://www.scunning.com/mixtape.html) by Cunningham

**Formal Prerequisites:** MATH 1110 or equivalent. 

**Highly recommended:** Intermediate micro and masters-level econometrics

**Lecture notes:** https://github.com/irudik/aem6510, accessible through Canvas

<!-- --- -->

<!-- # Zoom etiquette -->

<!-- A few requests on how you should interact through zoom: -->

<!-- 1. Be on time -->
<!-- 2. Stay muted unless answering/asking a question -->
<!-- 3. Turn on video, have a non-distracting actual or virtual background -->
<!-- 4. Display your full name and any preferred nick name -->
<!-- 5. Be active! -->
<!--   - Ask questions in chat or use the hand raise feature -->
<!--   - If I don't respond in a reasonable timeframe feel free to unmute and ask your question -->

---

# What's the point of the course?

To learn:

Core environmental theory

--

Theory of regulation

--

Applied welfare analysis

--

How to do empirical analysis on the environment

--

How to use R and do geospatial analysis
- This is increasingly important in environmental economics


---

# What do you need to succeed in this class?

Attend lectures if possible

--

Do your homework and prelims

--

Start your final project early

--

Contact me or Weilaing if you have and questions or concerns

---

# Here's how you'll be graded

Prelims (2): 30% and 25% (one theory, one empirical)   
Problem sets (4): 20%  
Literature review paper: 20%  
Literature review presentation: 5%

A: 92-100; A-: 90-91  
B+: 88-89; B: 82-87; B-: 80-81  
C+: 78-79; C: 72-77; C-: 70-71  
D+: 68-69; D: 62-67; D-: 60-61  
F: < 60 


---

# Prelims

There are two, one on theory and one on empirical methods

--

Your highest-scoring prelim is 30% of your grade,  
your lowest-scoring prelim is 25%

--

The empirical prelim is takehome, and you are expected to complete it .hi-red[on your own]

--

You have 48 hours to finish the empirical prelim

--

Late prelims automatically get 30% deducted: don't wait until the last minute to turn them in


---

# Problem sets

There will be 4 problem sets worth 10% each

--

Problem sets will be 3-4 questions long, typically from the Phaneuf and Requate book

--

You will have at least 1 week to complete each problem set

---

# Final project

There is a final project for the course aimed to get you moving toward research

--

You can choose one of two options:

1) Literature review

2) Data dive

---

# Final project: lit review

Write a literature review on up to 3 papers all in a particular subfield of environmental and resource economics

--

- What are the findings? Common threads? Work yet to be done?
- You'll need to run the papers by me before committing

--

You will present your literature review to the class

--

More details later

---

# Final project: data dive

Find a new data set that we will not use in class but is useful for environmental economics

--

- What are the data? How do you get them? How do you use them? What makes them relevant for economics?
- You'll need to run it by me

--

Do some preliminary analyses on the data

--

Present the dataset to the class

--

More details later

---

# Course schedule: Theory

1. Micro & math recap, theory of externalities
2. Intro to the theory of environmental policy
3. Intro to the theory of environmental policy
4. Imperfect information
5. Competitive output markets
6. Non-competitive output markets
7. Pre-existing distortions
8. Theory of applied welfare economics
9. Theory of applied welfare economics
10. Revealed preference models
11. Hedonics (theory)
12. Extra theory prelim office hours

---

# Course schedule: Empirics

13. R and the tidyverse, causal inference
14. R and the tidyverse, causal inference, RCTs
15. Deforestation, remote sensing/simple features, regression discontinuity
16. Hedonics (applied)
17. Climate change science
18. Climate change and the Ricardian Model
19. Climate change and two-way fixed effects
20. Environmental health, difference-in-differences, and event studies
21. Environmental health, difference-in-differences, and event studies
22. Perverse incentives, clean air act, remote sensing
23. Extra empirical prelim office hours


---

# Part 1: Math and micro review

Before getting into the core material we need to review the mathematical foundations underpinning the economics

--

What is central to economic decisionmaking is .hi-blue[constrained optimization]

--

- We want to max utility subject to a budget
- We want to max profits subject to a constraining regulation
- We want to max social welfare subject to an environmental target
- We want to minimize costs subject to meeting electricity demand
- Etc
  

---

# Unconstrained optimization

.hi-blue[Unconstrained optimization] is just Calculus 101

--

Suppose we want to maximize utility from eating fruit:

--

$$ \max_{x_1,...,x_N \in \mathbf{A}} U(\mathbf{x}) $$ 
where $U(\mathbf{x}):\mathbb{R}^N \rightarrow\mathbb{R}$

--

.b[Theorem]: Let $U$ be once continuously-differentiable ( $\mathbb{C}^1$ ) so that first partial derivatives exist and are continuous. If $x^*$ is a local maximum of $U$ on $A$ then

--

$${\partial U(\mathbf{x}^*) \over \partial x_i} = 0 \text{ for } i = 1,...,N$$

---

# Unconstrained optimization

What does this theorem tell us?

--

That the first-order conditions give us the .hi-blue[necessary conditions] for the utility-maximizing amount of each type of fruit to eat

--

These conditions .b[must] be satisfied for any bundle that maximizes utility

--

In a logical statement, .ex[if X then Y], Y is the necessary condition: Y .b[must] be true for X to be true

X is the .hi-blue[sufficient condition]

---

# Unconstrained optimization

```{css echo=FALSE}
.panelset {
  --panel-tab-color: currentColor;
  --panel-tab-color-active: currentColor;
  --panel-tab-color-hover: currentColor;
  --panel-tabs-border-bottom: #ddd;
  --panel-tab-inactive-opacity: 0.25;
  --panel-tab-font-family: Roboto;
}
```


```{r plot_foc, echo = FALSE, fig.show = 'hide'}
fun.1 <- function(x) -x^2 + 4*x + 4
ggplot() +
  stat_function(fun = fun.1, color = "#638ccc", size = 1.5) +
  geom_hline(yintercept = 8, linetype = 'dashed', size = .5) +
  geom_vline(xintercept = 2, color = "#ca5670", size = 1.5) +
  theme_empty +
  lims(x = c(0, 4)) +
  labs(x = "",
       y = "",
       title = "Tangent at optimal x is horizontal")
  
```

.pull-left[
$${\partial U(\mathbf{x}^*) \over \partial x_i} = 0 \text{ for } i = 1,...,N$$

Our necessary conditions tell us that the marginal utility of each fruit equals zero at the optimal choice

If $x^*$ is a maximum of $U$ on $A$, then the tangent line/plane of $U$ at $(U(x^*), x^*)$ must be horizontal
]
.pull-right[

![Tangent line at the optimum has 0 slope](`r knitr::fig_chunk("plot_foc", "png")`)

]




---

# Unconstrained optimization

Do necessary conditions guarantee that we have found the utility-maximizing bundle of fruit?

--

.hi-red[NO!]

--

They are necessary but not .hi-red[sufficient]: you can come up with counter-examples where a function's derivative equals zero but it is not maximized ( e.g. $x^3$ )

--

We need additional conditions to impose sufficiency

---

# Unconstrained optimization

These are the second-order sufficient conditions

--

.b[Theorem]: Let $U$ be twice continuously differentiable. Suppose that $x^*$ is a critical point of $U$ on $A$ and its Hessian matrix is negative (semi-)definite at $x^*$ ( $D^2 U(x^*)$ is negative (semi-)definite ). Then $x^*$ is a strict (weak) local maximum.

--

Negative semi-definite is basically just saying that a multivariate function is weakly concave

---

# Unconstrained optimization

```{css echo=FALSE}
.panelset {
  --panel-tab-color: currentColor;
  --panel-tab-color-active: currentColor;
  --panel-tab-color-hover: currentColor;
  --panel-tabs-border-bottom: #ddd;
  --panel-tab-inactive-opacity: 0.25;
  --panel-tab-font-family: Roboto;
}
```

.panelset[

.panel[.panel-name[Words Details]

We need second-order conditions to guarantee we have found a maximum


We need marginal utility to be .b[decreasing] (i.e. utility be concave) at $x^*$


This guarantees $x^*$ maximizes $U$ if $x^*$ satisfies the first-order necessary conditions

]

.panel[.panel-name[Math Details]

Approximate $U$ around $x^*$ with a second-order Taylor expansions:
$$U(x^*+h) = U(x^*) + DU(x^*)h + {1 \over 2} h^TD^2U(x^*)h + O(h^2)$$
$DU(x^*) = 0$ by definition of $x^*$ so that
$$U(x^*+h)-U(x^*) = {1\over2} h^T D^2U(x^*)h$$
$U$ is always decreasing away from $x^*$ if ${1\over2}h^T D^2U(x^*)h < 0$

]

]


---

# Constrained optimization

Before we were able to eat as much fruit as we wanted

--

What if our actions are now constrained?

--

$$ \max_{x_1,...,x_N \in \mathbf{A}} U(\mathbf{x}), \,\,\,\,\, A = \\{ \mathbf{x} \in \mathbb{R}^n: h_1(\mathbf{x}) = a_1, \dots, h_m(\mathbf{x}) = a_m \\}$$

where $m < n$

---

# Constrained optimization

```{r budget_constraint, echo = FALSE, fig.show = 'hide', warning = F}
ic <- function(x) 25/x
budget <- function(x) 10 - x
ggplot() +
  stat_function(fun = ic, color = "#638ccc", size = 1.5) +
  stat_function(fun = budget, color = "#ca5670", size = 1.5) +
  theme_empty +
  lims(x = c(0, 15), y = c(0, 15)) +
  labs(x = "",
       y = "",
       title = "")
  
```

.pull-left[

A simple and familiar example is utility maximization:
$$ \max_{x_1,x_2} U(x_1,x_2) \text{ subject to } h(x_1, x_2) = a $$

We want to be on the .hi-blue[highest indifference curve] subject to our .hi-red[budget constraint]

]

.pull-right[

![Tangent line at the optimum has 0 slope](`r knitr::fig_chunk("budget_constraint", "png")`)

]

---

# Constrained optimization

Because of our budget constraint, marginal utility may not equal zero at the utility-maximizing bundle!

--

No proof, but for some $\lambda$ we can characterize this bundle by

--

$$DU(x^*_1, x^*_2) - \lambda Dh(x_1^*, x_2^*) = (0,0)$$
and

--

$$h(x_1^*,x_2^*) = a$$

--

We now have 3 equations, and 3 unknowns $(x^*_1,x^*_2,\lambda^*)$, we just transformed our constrained problem into an unconstrained problem

---

# Constrained optimization

How does this magic happen?

--

.hi-blue[Intuition:] We can reframe a constrained optimization problem in an alternative fashion where we allow the agent to pick whatever $(x_1,x_2)$ they want, .b[but], they must pay a fine $\lambda$ per unit violation of the restriction

--

This gives us an alternative, unconstrained problem, known as a *Lagrangian*:
$$\mathcal{L}(x_1,x_2,\lambda) = \max_{x_1,x_2,\lambda} U(x_1,x_2) - \lambda[h(x_1,x_2) - a]$$

--

Maximize utility, but also pay a fine $\lambda$ for each unit difference between $h(x_1,x_2)$ and $a$

---

# Constrained optimization

$$\max_{x_1,x_2,\lambda} U(x_1,x_2) - \lambda[h(x_1,x_2) - a]$$

with first-order conditions:

--

$$ {\partial \mathcal{L} \over \partial x_i} = {\partial U \over \partial x_i} - \lambda {\partial h \over \partial x_i} = 0 \text{ for } i = 1,2$$

--

Notice marginal utility is no longer zero at the optimum allocation unless $\lambda = 0$

---

# Constrained optimization

We have one remaining issue: some arbitrary $\lambda$ won't guarantee that the solution of our Lagrangian will match the actual constrained problem

--

We need to find a way to pick the .hi-blue[correct] $\lambda^*$ that incentivizes the agent to exactly satisfy the constraint $h(x_1,x_2) = a$

--

In other words, we need a third condition to pin down $\lambda^*$, this is the feasibility condition:
$${\partial \mathcal{L} \over \partial \lambda} = h(x_1,x_2) - a = 0$$
Which is just the constraint!

---

# Constrained optimization

We can translate a constrained optimization problem into an unconstrained one (the Lagrangian) by:

--

1. Adding a new variable $\lambda$, which is called the *Lagrange Multiplier* or the *shadow value* of the constraint
--

2. Adding another first-order condition (feasibility condition) and modifying the original first-order conditions to reflect the constraint

--

The intuition for $\lambda$ is that it is the marginal value of loosening (or tightening) the constraint

